---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Sustainability

```{r setup-01, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Define** sustainability and sustainable design, and explain the three pillars (social, economic, ecological) and the "triple bottom line."
2. **Identify** the four points of the sustainability platform --- recycling, bio-based/biodegradable polymers, material reduction, and design for recyclability --- and give an engineering example of each.
3. **Interpret** the seven plastic recycling codes and distinguish post-consumer from post-industrial recyclate.
4. **Analyze** the trade-off between recyclability and biodegradability, using the corn-based PLA container case as the worked example.
5. **Calculate** the material, energy, and carbon consequences of a design decision (material substitution, wall-thickness reduction) for a production run.
6. **Evaluate** a product on whole-life cost, and explain why "cheaper to buy" and "cheaper to own" often disagree.
7. **Apply** a sustainability checklist to a project: supplier selection, contract clauses, impact assessment, circular-economy principles, and sustainability KPIs.

<details>
<summary>Before we begin: think of the last thing you threw away today</summary>

Where did its materials come from? How long did you use it before it became
waste? Where is it now --- landfill, recycling stream, incinerator? Could a
different design decision, made years ago by an engineer you will never meet,
have changed that answer? Keep the object in mind; we will come back to it.

</details>

## Introduction: every gram, multiplied by a hundred thousand

Here is the situation that runs through this whole book. Your company designs
and manages the launch of a **1/18-scale radio-controlled vehicle replica** for
the toy market. If the prototype is accepted, you win a contract to build
**100 000+ units** for the next holiday season. Early on, you must choose the
body material --- **steel, aluminium, or carbon fibre** --- and the frame
material --- aluminium or carbon fibre.

That single choice is not just a strength-and-weight problem. Multiply it by
100 000 units and it becomes a decision about:

- how many tonnes of raw material leave the ground,
- how many megajoules of energy are spent turning that material into parts,
- how much scrap and packaging waste the line produces every shift,
- whether the finished toy can be recycled when a child outgrows it in three years,
- and whether your customer --- a large retailer with its own public sustainability targets --- will even put it on the shelf.

This is why **sustainability** belongs at the front of a project-management
course, not tacked on at the end. The tools you meet in later chapters --- scope,
WBS, schedule, budget, risk --- all decide how resources get spent. Sustainability
is the question of whether that spend was worth it once you look past the launch
date.

::: {.callout-note}
**Industry snapshot.** General Motors' CAMI Assembly plant in Ingersoll,
Ontario --- an hour from where this course is taught --- runs **landfill-free**:
every waste stream from daily operations is reused, recycled, or converted to
energy. That is not a marketing slogan bolted onto an existing plant; it is a
set of thousands of small design and process decisions of exactly the kind this
chapter is about.
:::

```{r video-story-of-stuff, echo=FALSE, results='asis'}
embed_youtube("9GorqroigqM", "The Story of Stuff (Annie Leonard) - the materials economy in 20 minutes")
```

## What sustainability means

The most quoted definition comes from the 1987 Brundtland report and is used
almost word-for-word by the U.S. Environmental Protection Agency:

> Sustainability is *"meeting the needs of the present without compromising the
> ability of future generations to meet their own needs."*

The idea is older than the word. Thomas Jefferson expressed the same principle
in financial terms: **do not leave a debt for future generations to pay.**
Swap "debt" for "depleted resources, degraded ecosystems, and accumulated
waste" and you have the modern version.

**Sustainable design** applies that principle to the things engineers actually
produce. Wikipedia's working definition is useful: *the philosophy of designing
physical objects, the built environment, and services to comply with the
principles of social, economic, and ecological sustainability.* The intent is to
*minimize negative environmental impact through skilful, sensitive design* while
holding a **dynamic balance between economy and society**.

### The three pillars and the triple bottom line

Sustainability is usually drawn as three overlapping domains. A decision is
sustainable only where all three overlap: it has to be **socially** acceptable,
**economically** viable, and **ecologically** sound at the same time.

```{r fig-three-pillars-01, fig.cap="The three pillars of sustainability. Only decisions in the central overlap are genuinely sustainable; the pairwise overlaps (bearable, equitable, viable) each miss one dimension.", fig.align='center', fig.width=9, fig.height=6.5}
circles <- data.frame(
  x0 = c(-0.55, 0.55, 0),
  y0 = c(0.35, 0.35, -0.6),
  r  = rep(1.05, 3),
  pillar = c("Social\n(bearable)", "Economic\n(viable)", "Ecological\n(sound)"),
  fill = c("#2E86C1", "#F39C12", "#27AE60")
)

ggplot(circles) +
  ggforce::geom_circle(aes(x0 = x0, y0 = y0, r = r, fill = fill),
                       alpha = 0.35, linewidth = 0.8, color = "grey30") +
  scale_fill_identity() +
  annotate("text", x = -0.95, y = 0.9, label = "SOCIAL",
           fontface = "bold", size = 4.5 * diagram_text_scale, color = "#1B4F72") +
  annotate("text", x = 0.95, y = 0.9, label = "ECONOMIC",
           fontface = "bold", size = 4.5 * diagram_text_scale, color = "#9C640C") +
  annotate("text", x = 0, y = -1.55, label = "ECOLOGICAL",
           fontface = "bold", size = 4.5 * diagram_text_scale, color = "#186A3B") +
  annotate("text", x = 0, y = 0.55, label = "bearable",
           size = 3.2 * diagram_text_scale, fontface = "italic") +
  annotate("text", x = -0.7, y = -0.35, label = "equitable",
           size = 3.2 * diagram_text_scale, fontface = "italic") +
  annotate("text", x = 0.7, y = -0.35, label = "viable",
           size = 3.2 * diagram_text_scale, fontface = "italic") +
  annotate("text", x = 0, y = -0.1, label = "SUSTAINABLE",
           fontface = "bold", size = 3.6 * diagram_text_scale, color = "grey15") +
  coord_equal() +
  labs(title = "Sustainability lives in the overlap") +
  theme_void() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5)
  )
```

<details>
<summary>Discussion: which pillar does each of these decisions serve --- and which does it ignore?</summary>

- A supplier offers motors 20 % cheaper, made in a plant with a poor safety record. *(economic yes; social no)*
- You switch the body to recycled aluminium, adding \$0.40 per unit. *(ecological yes; economic maybe)*
- You keep an inefficient old assembly line running because retraining staff on a new one is disruptive. *(social short-term yes; ecological and economic long-term no)*

The point of the diagram is that a "win" on one pillar is not a win. You have
to check all three.

</details>

## The four points of the sustainability platform

For a product engineer, sustainability comes down to four levers:

1. **Recycling** --- can the material be recovered and used again?
2. **Bio-based or biodegradable polymers** --- can the material come from a renewable source, or return safely to the environment?
3. **Material reduction** --- can the product do its job with less material in the first place?
4. **Designing for recyclability** --- is the product built so that recovery is actually practical, not just theoretically possible?

The rest of the chapter takes these one at a time, then widens out to what
sustainability means when you are managing a whole project rather than designing
one part.

### Sustainable design principles

The design literature expands the four levers into a set of working principles:

| Principle | What it means in practice |
|---|---|
| **Low-impact materials** | Non-toxic, sustainably produced or recycled materials that need less energy to process. |
| **Energy efficiency** | Choose manufacturing processes --- and design products --- that require less energy over their life. |
| **Emotionally durable design** | Build things people want to keep. A product kept for ten years instead of three cuts consumption and waste by more than any recycling scheme. |
| **Design for reuse and recycling** | Design for a commercial "afterlife" --- but without compromising performance or service life to get there. It is a balancing act. |

Notice the tension already built in: "make it last" and "make it recyclable"
pull in opposite directions, and "use less material" can fight "make it
durable." Good sustainable design is the art of resolving those tensions, not
pretending they do not exist.

## Recycling codes

Most rigid plastics carry one of seven **resin identification codes** --- the
number inside the chasing-arrows triangle. The code tells a sorting facility
what polymer it is looking at. The three most commonly recycled are **PET
(#1)**, **HDPE (#2)**, and **PP (#5)**.

```{r tbl-recycling-codes-01}
recycling_codes <- data.frame(
  Code = 1:7,
  Abbreviation = c("PETE / PET", "HDPE", "PVC / V", "LDPE", "PP", "PS", "OTHER"),
  Polymer = c(
    "Polyethylene terephthalate", "High-density polyethylene",
    "Polyvinyl chloride", "Low-density polyethylene",
    "Polypropylene", "Polystyrene",
    "Acrylic, polycarbonate, PLA, nylon, multi-layer"
  ),
  `Typical products` = c(
    "Drink bottles, food jars",
    "Milk jugs, detergent bottles, piping",
    "Pipe, window frames, cable insulation",
    "Films, bags, squeeze bottles",
    "Bumpers, battery cases, living hinges, tubs",
    "Foam packaging, disposable cups, CD cases",
    "Eyewear, headlamp lenses, RC-car bodies, PLA parts"
  ),
  `Recycled easily?` = c("Yes", "Yes", "Rarely", "Sometimes", "Yes", "Rarely", "Rarely"),
  check.names = FALSE
)

knitr::kable(recycling_codes, align = c("c", "l", "l", "l", "c"),
             caption = "The seven resin identification codes.") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(1, bold = TRUE) |>
  kableExtra::row_spec(c(1, 2, 5), background = "#EAF6EC")
```

Two grades of recyclate matter to a manufacturer:

- **Post-industrial** (a.k.a. regrind): clean scrap from your own or a supplier's process --- runners, sprues, start-up shots, off-spec parts. Easy to reuse because you know exactly what it is.
- **Post-consumer**: material that has been through a consumer's hands. It must be collected, sorted, washed, ground, and often re-pelletized, and it can carry metal, labels, food residue, and other polymers. Much harder to work with, but it is the stream that actually diverts waste from landfill.

<details>
<summary>Think about it: your RC-car body is moulded from a glass-filled ABS/PC blend. Which recycling code does it carry, and what does that tell you?</summary>

It carries **#7 (OTHER)** --- any multi-polymer blend or filled material that is
not one of codes 1--6 lands here. In practice, #7 is rarely recycled through
curbside programs: the blend cannot be melted back into a single clean polymer.
If recyclability matters for this product, that is a signal to reconsider the
material *before* tooling is cut, not after.

</details>

### Try it yourself: which resin codes get recycled?

```{r tryit-resin-01, echo=FALSE, results='asis'}
resin_bins <- c("Recycled easily", "Sometimes", "Rarely")
tryit_sort(
  "resin-codes", "How well does each product's plastic recycle?",
  bins = resin_bins,
  cards = data.frame(
    text = c("Soft-drink bottle (PET, #1)", "Laundry-detergent jug (HDPE, #2)",
             "Yoghurt tub (PP, #5)", "Honey squeeze bottle (LDPE, #4)",
             "Grocery-bag film (LDPE, #4)", "Window frame (PVC, #3)",
             "Foam takeaway clamshell (PS, #6)", "Polycarbonate eyewear lens (#7)"),
    bin = c("Recycled easily", "Recycled easily", "Recycled easily", "Sometimes",
            "Sometimes", "Rarely", "Rarely", "Rarely"),
    why = c("PET (#1) is one of the three most commonly recycled codes.",
            "HDPE (#2) is one of the three most commonly recycled codes.",
            "PP (#5) is one of the three most commonly recycled codes.",
            "LDPE (#4) is only sometimes accepted, depending on the municipality.",
            "Films jam sorting equipment; LDPE is recycled only sometimes.",
            "PVC (#3) is rarely recycled and contaminates other streams.",
            "PS (#6) is bulky, light and rarely recycled.",
            "#7 (other) covers blends and engineering plastics: rarely recycled."),
    stringsAsFactors = FALSE),
  intro = "Use the code table above. Pick one class for every product.", question = NULL)

tryit_sort(
  "resin-twin", "How well does each product's plastic recycle? (graded)",
  bins = resin_bins, show = FALSE,
  cards = data.frame(
    text = c("Milk jug (HDPE, #2)", "Water bottle (PET, #1)", "Margarine tub (PP, #5)",
             "CD case (PS, #6)", "Sandwich-bag film (LDPE, #4)", "Cable insulation (PVC, #3)"),
    bin = c("Recycled easily", "Recycled easily", "Recycled easily", "Rarely", "Sometimes", "Rarely"),
    stringsAsFactors = FALSE),
  question = "<p>Match each product to how readily its resin code is recycled, using the seven resin identification codes.</p>")
```

## Recyclable vs. biodegradable: a real tension

Years ago the plastics industry pursued two waste strategies at once ---
**recycling** and **biodegradability** --- and they turned out to be partly
opposed. If you make a part from recyclate and want it to last (think plastic
lumber, or a pallet), the last thing you want is biodegradable material mixed
in, causing the part to fail early. Contaminate a recycling stream with enough
compostable polymer and the whole batch can become unusable.

The industry chose recycling. But with fossil-fuel prices and supply security
back in the spotlight, **biopolymers** have returned. Some of the earliest
plastics were bio-based (cellulosics, natural rubber); many new ones are built
from **corn or potato starch**. **Polylactic acid (PLA)** is made from corn
starch and behaves a lot like PET, but it is harder to process and degrades
faster at elevated temperature.

And "biodegradable" has an asterisk. Degradation is done by **microbes**, which
need **oxygen and moisture**. A modern landfill is compacted, dry, and
low-oxygen --- almost nothing degrades well there. You can dig up decades-old
newspaper and still read it.

::: {.callout-important}
**Case study --- EcoPlast Packaging (MGMT-3105, Week 1).** EcoPlast launched
corn-based PLA food containers. Then: customers found them brittle in the cold
and warped in heat; recyclers reported PLA contaminating conventional-plastic
batches; consumers wanted eco-friendly *and* the same durability as regular
plastic; and environmental groups argued the containers mostly ended up in
landfill, where they did not meaningfully biodegrade --- calling it "green
marketing."
:::

<details>
<summary>Discussion (Week 1 case): was it unreasonably risky for EcoPlast to trade durability for sustainability, and what would you change?</summary>

**Model answer.** A product that fails in real-world use creates *more* waste,
erodes consumer trust, and undermines the point of sustainability --- so yes, the
trade-off as executed was unreasonably risky. Consumers share *some*
responsibility for unrealistic expectations, but the company had a duty to
manage those expectations and to test properly before launch. Rushing to market
compromised quality.

What to change: (1) real R&D and performance testing across the actual
temperature and handling range before launch; (2) an accurate **caution** label
("not suitable for hot liquids") --- a danger label would be excessive; (3)
consumer education on correct disposal (industrial compost, not curbside or
landfill); (4) stop marketing the product as fully "sustainable" until there is
evidence of an environmental benefit in real disposal conditions --- otherwise
the claim is exposed to a misleading-marketing challenge.

</details>

```{r video-tshirt-lifecycle, echo=FALSE, results='asis'}
embed_youtube("BiSYoeqb_VY", "The Life Cycle of a T-Shirt (TED-Ed) - why 'where does it go?' is the hard question")
```

## Material reduction

Reducing the material that goes into a product --- at the **design stage**, by
thinning walls, removing bosses, right-sizing ribs, or substituting a lighter
material --- pays off several times over:

- the product **costs less** (less material bought per unit),
- there is **less to recycle or dispose of** at end of life,
- **more material is left** for other uses,
- and **less energy** is spent producing *and transporting* every unit.

The catch: you cannot thin your way past the product's functional requirements.
Material reduction is enabled by better material properties, better processes,
and better analysis --- not by wishful thinking.

### Worked Example 1: thinning the RC-car body

Your baseline RC-car body is injection-moulded ABS with a mass of **48 g**.
Simulation says you can drop the nominal wall from 2.5 mm to 2.1 mm and still
pass the drop test, cutting body mass to **41 g**. The contract run is
**120 000 units**. ABS costs **\$3.10/kg**, its production embodied energy is
about **90 MJ/kg**, and its cradle-to-gate carbon intensity is about
**3.4 kg CO2e/kg**.

Live cell --- change the mass reduction, the run size, or the material's cost and energy factors.

```{r calc-material-reduction-example-01, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "material-reduction-example-01",
  title = "Material Reduction Calculator",
  icon = "leaf",
  inputs = list(
    qwebr_calc_input("mass_before_g", "Baseline body mass", value = 48, min = 10, max = 500, step = 1, suffix = "g"),
    qwebr_calc_input("mass_after_g", "Thinned body mass", value = 41, min = 5, max = 500, step = 1, suffix = "g"),
    qwebr_calc_input("abs_cost_per_kg", "ABS cost", value = 3.1, min = 0.5, max = 20, step = 0.1, prefix = "$", suffix = "/ kg"),
    qwebr_calc_input("abs_energy_MJ_kg", "ABS embodied energy", value = 90, min = 10, max = 400, step = 5, suffix = "MJ / kg"),
    qwebr_calc_input("abs_co2e_kg_kg", "ABS carbon intensity", value = 3.4, min = 0.5, max = 30, step = 0.1, suffix = "kg CO2e / kg"),
    qwebr_calc_input("units", "Production run", value = 120000, min = 5000, max = 500000, step = 5000, type = "slider")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: material-reduction-example-01
mass_before_g <- 48
mass_after_g  <- 41
units         <- 120000
abs_cost_per_kg   <- 3.10
abs_energy_MJ_kg  <- 90
abs_co2e_kg_kg    <- 3.4

saved_kg <- (mass_before_g - mass_after_g) / 1000 * units

cost_saved   <- saved_kg * abs_cost_per_kg
energy_saved <- saved_kg * abs_energy_MJ_kg
co2_saved    <- saved_kg * abs_co2e_kg_kg

fmt <- function(x) format(round(x), big.mark = " ", scientific = FALSE, trim = TRUE)

# Chart: CO2e avoided grows with the production run
runs <- seq(0, 500000, length.out = 100)
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
plot(runs / 1000, (mass_before_g - mass_after_g) / 1000 * runs * abs_co2e_kg_kg / 1000,
     type = "l", lwd = 4, col = "#16a085", xlab = "Production run (thousand units)", ylab = "CO2e avoided (tonnes)",
     main = "Carbon avoided vs. production run")
abline(v = units / 1000, lty = 2, col = "#bdc3c7")
points(units / 1000, co2_saved / 1000, pch = 19, cex = 2.2, col = "#e74c3c")

cat(sprintf("%s kg of material saved over the run\n", fmt(saved_kg)))
cat(sprintf("Mass removed per unit  %s g\n", mass_before_g - mass_after_g))
cat(sprintf("Material cost saved  $%s\n", fmt(cost_saved)))
cat(sprintf("Production energy saved  %s MJ\n", fmt(energy_saved)))
cat(sprintf("Equivalent electricity  ~%s kWh\n", fmt(energy_saved / 3.6)))
cat(sprintf("Cradle-to-gate CO2e avoided  %s kg (~%.1f t)\n", fmt(co2_saved), co2_saved / 1000))
```

</details>

A 7 g decision, taken once on a screen, moves almost a tonne of material and
several tonnes of CO2e once it is multiplied by the production run --- before
counting the lighter shipping weight.

### Try it yourself: material reduction by hand

Do the same calculation for a part of your own. Work out the material,
cost, energy and carbon consequences of thinning the part over the whole run.
Each team picks its own **data set**.

```{r tryit-material-01, echo=FALSE, results='asis'}
mat_sets <- list(
  c(part = "Washer-fluid bottle (HDPE)", before = 62, after = 54, units = 300000, cost = 1.9,  energy = 80, co2 = 1.9),
  c(part = "Battery cover (PP)",        before = 45, after = 38, units = 250000, cost = 2.2,  energy = 75, co2 = 2.1),
  c(part = "Cable tray (ABS)",          before = 120, after = 104, units = 80000, cost = 3.4, energy = 95, co2 = 3.6),
  c(part = "Tool-box lid (PP)",         before = 30, after = 26, units = 1200000, cost = 1.7, energy = 70, co2 = 2.0))
mat_given <- function(v) c(
  Part = v[["part"]],
  "Mass before / after" = sprintf("%s g / %s g", v[["before"]], v[["after"]]),
  "Production run" = format(as.numeric(v[["units"]]), big.mark = " ", scientific = FALSE),
  "Material cost" = sprintf("$%s / kg", v[["cost"]]),
  "Embodied energy" = sprintf("%s MJ / kg", v[["energy"]]),
  "Carbon intensity" = sprintf("%s kg CO2e / kg", v[["co2"]]))
mat_answers <- function(v) {
  n <- function(k) as.numeric(v[[k]])
  kg <- (n("before") - n("after")) / 1000 * n("units")
  list(tryit_answer("Material saved over the run", kg, tol = 1, tol_pct = 0.1, suffix = "kg"),
       tryit_answer("Material cost saved", kg * n("cost"), tol = 1, tol_pct = 0.1, prefix = "$"),
       tryit_answer("Production energy saved", kg * n("energy"), tol = 1, tol_pct = 0.1, suffix = "MJ"),
       tryit_answer("Cradle-to-gate CO2e avoided", kg * n("co2"), tol = 1, tol_pct = 0.1, suffix = "kg"))
}
tryit_number(
  "material-by-hand", "Material reduction over a production run",
  answers = mat_answers(mat_sets[[1]]), given = mat_given(mat_sets[[1]]),
  teams = lapply(mat_sets[-1], function(v) list(given = mat_given(v), answers = mat_answers(v))),
  intro = "Mass saved per unit x units, then multiply by each factor.")

# Graded twin: different part and figures.
mat_twin <- c(part = "Pump housing (PA6)", before = 85, after = 71, units = 150000, cost = 3.9, energy = 120, co2 = 5.6)
tryit_number(
  "material-twin", "Material reduction (graded)", show = FALSE,
  answers = mat_answers(mat_twin),
  question = paste0("<p>A pump housing in nylon is thinned from 85 g to 71 g per unit. The run is ",
    "150 000 units. The material costs $3.90/kg, has an embodied energy of 120 MJ/kg and a ",
    "cradle-to-gate carbon intensity of 5.6 kg CO2e/kg. Calculate the totals over the run.</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

Saved mass = (mass before - mass after) / 1000 x units, in kg. Each of cost,
energy and carbon is that mass times its per-kg factor. For data set 1:
8 g x 300 000 = 2 400 kg, which at \$1.90/kg is \$4 560, at 80 MJ/kg is
192 000 MJ and at 1.9 kg CO2e/kg is 4 560 kg CO2e.

</details>

### Try it yourself: the material-choice decision

Now compare the three candidate body materials head to head. The live cell below
holds rough figures for **steel**, **aluminium**, and **carbon-fibre composite**
bodies of the mass each material would need for equal stiffness. Change the
numbers, add a recycled-content option, or change the run size and see what
happens to cost, energy, and carbon.

```{webr-r}
#| label: material-choice-compare-01
# Rough per-unit body mass for equal stiffness, and cradle-to-gate factors.
material <- c("Steel", "Aluminium", "Carbon-fibre")
mass_g   <- c(140,      70,          38)          # grams per body
cost_kg  <- c(1.20,     3.00,        28.00)       # $ per kg
energy_kg<- c(25,       155,         300)         # MJ per kg (primary)
co2_kg   <- c(2.0,      11.0,        24.0)        # kg CO2e per kg

units <- 120000

kg_total     <- mass_g / 1000 * units
cost_total   <- kg_total * cost_kg
energy_total <- kg_total * energy_kg
co2_total    <- kg_total * co2_kg

result <- data.frame(
  Material     = material,
  Body_mass_g  = mass_g,
  Material_t   = round(kg_total / 1000, 1),
  Cost_usd     = round(cost_total),
  Energy_MWh   = round(energy_total / 3600, 1),
  CO2e_t       = round(co2_total / 1000, 1)
)
print(result, row.names = FALSE)

cat("\nLightest body:", material[which.min(mass_g)], "\n")
cat("Lowest material cost:", material[which.min(cost_total)], "\n")
cat("Lowest cradle-to-gate CO2e:", material[which.min(co2_total)], "\n")
```

<details>
<summary>What the numbers usually show --- and why the "greenest" answer is not obvious</summary>

Carbon fibre gives the lightest body, which cuts shipping energy and, in a real
vehicle, fuel over the whole service life. But its *production* energy and carbon
per kilogram are the highest of the three, and it is the hardest to recycle
(code #7, thermoset matrix). Steel is cheap and highly recycled but heavy.
Aluminium sits in between and is excellent if you can get **recycled** stock ---
secondary aluminium uses roughly 5 % of the energy of primary.

For a *toy* that is shipped once and never driven under its own power for
thousands of kilometres, the use-phase fuel saving from carbon fibre never
arrives --- so the decision tilts back toward recycled aluminium or steel. The
lesson: **the right material depends on where the impact actually lands in the
life cycle.**

</details>

## Designing for recyclability

Recyclability is designed in, not hoped for. The checklist a plastics engineer
can actually influence:

- Make the assembly from a **single material**, or from **compatible** materials that can be recycled together.
- Make it **easy to disassemble**.
- Prefer **snap-fits and press-fits** over screws and metal fasteners.
- **Reduce the part count.**
- **Avoid labels** (or use same-polymer labels / in-mould labelling).
- Keep **thermosets** out of thermoplastic assemblies --- use thermoplastic elastomers instead of rubber seals where you can.

You may not have the final word on a design, but you can always **offer the
greener alternative**. And the mindset has shifted: products used to be designed
for *end of life* --- last for period X, then done. Now good products are designed
for **next life**: what can this material become *after* its first use?

<details>
<summary>Try it: redesign the RC-car battery door for recyclability</summary>

**Current design:** ABS door, two stainless M2 screws, a printed polyester
label, a moulded-in TPE gasket (thermoset-like behaviour), a steel spring.

<details>
<summary>One redesign</summary>

- Replace the two screws with a **living hinge + snap latch** moulded into the door (same ABS) --- removes two dissimilar-metal fasteners and a driver operation on the line.
- Replace the adhesive polyester label with **in-mould labelling** in an ABS-compatible film, or mould the text into the part.
- Replace the steel spring with a **moulded ABS cantilever spring** if the cycle life allows.
- Keep the TPE gasket but make it a **removable press-in** part in a recycling-compatible grade, or switch to a moulded-in flexible ABS lip if the seal spec allows.

Result: the door goes from five materials and two fasteners to essentially one
polymer, snaps apart by hand, and carries no adhesive.

</details>

</details>

## The consumer's perspective

Sustainable design has to survive contact with real buyers. Market research is
consistent on two points:

1. Environmentally conscious consumers **will pay marginally more** and **accept marginally lower fit and finish** for a product that is genuinely more environmentally friendly.
2. They have **zero tolerance** for a drop in **performance/functionality** or in **service life**. They bought the product to do a job. If a "sustainable" design wears out sooner, it just gets replaced sooner --- consuming more resources and defeating the purpose.

```{r fig-consumer-tolerance-01, fig.cap="Consumer tolerance for trade-offs made in the name of sustainability: some room on price and cosmetics, essentially none on how well the product works or how long it lasts.", fig.align='center', fig.width=9, fig.height=5}
tol <- data.frame(
  attribute = c("Higher price", "Cosmetic fit & finish",
                "Core performance", "Service life / durability"),
  tolerance = c(0.55, 0.45, 0.03, 0.05)
)
tol$attribute <- factor(tol$attribute, levels = rev(tol$attribute))
tol$kind <- ifelse(tol$tolerance > 0.2, "Some tolerance", "Near-zero tolerance")

ggplot(tol, aes(x = tolerance, y = attribute, fill = kind)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = scales::percent(tolerance, accuracy = 1)),
            hjust = -0.15, size = 3.8 * diagram_text_scale) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 0.7),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c("Some tolerance" = "#27AE60",
                               "Near-zero tolerance" = "#C0392B")) +
  labs(title = "Where consumers will --- and will not --- give ground",
       x = "Approximate share of consumers willing to accept a trade-off",
       y = NULL, fill = NULL) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Whole-life cost: cheaper to buy vs. cheaper to own

This is where **planned obsolescence** enters. A business may have a commercial
reason to limit a product's life --- it protects repeat sales. Sustainable design
pushes the other way: make it last. The buyer's own wallet is often on the side
of durability, even when the sticker price is higher.

The live cell compares a **cheap** version of a product (low price, short life,
replaced repeatedly) with a **durable** version (higher price, lasts the whole
period). Change the prices, lifetimes, and horizon and watch which one wins on
**total cost of ownership** --- and on **units sent to landfill**.

```{r calc-whole-life-cost-01, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "whole-life-cost-01",
  title = "Whole-Life Cost Calculator",
  icon = "scale-balanced",
  inputs = list(
    qwebr_calc_input("cheap_price", "Cheap version: price", value = 18, min = 1, max = 200, step = 1, prefix = "$"),
    qwebr_calc_input("cheap_life_yrs", "Cheap version: lifetime", value = 2, min = 0.5, max = 20, step = 0.5, suffix = "yr"),
    qwebr_calc_input("durable_price", "Durable version: price", value = 42, min = 1, max = 500, step = 1, prefix = "$"),
    qwebr_calc_input("durable_life_yrs", "Durable version: lifetime", value = 10, min = 0.5, max = 30, step = 0.5, suffix = "yr"),
    qwebr_calc_input("horizon_years", "Years you need the product", value = 10, min = 1, max = 20, step = 1, type = "slider")
  ),
  goal = "Make the cheap version the cheaper one to own, without changing either price. Which input can you move?",
  goal_test = "cheap_tco < durable_tco"
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: whole-life-cost-01
horizon_years <- 10

# Cheap option
cheap_price     <- 18
cheap_life_yrs  <- 2

# Durable option
durable_price     <- 42
durable_life_yrs  <- 10

cheap_purchases   <- ceiling(horizon_years / cheap_life_yrs)
durable_purchases <- ceiling(horizon_years / durable_life_yrs)

cheap_tco   <- cheap_purchases   * cheap_price
durable_tco <- durable_purchases * durable_price

better <- if (durable_tco < cheap_tco) "Durable" else if (durable_tco > cheap_tco) "Cheap" else "Neither"

tco <- c(Cheap = cheap_tco, Durable = durable_tco)
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
bp <- barplot(tco, col = c("#e67e22", "#16a085"), ylim = c(0, max(tco) * 1.2),
              ylab = "Total cost of ownership ($)",
              main = paste("Cost over", horizon_years, "years"))
text(bp, tco, labels = paste0("$", tco), pos = 3, cex = 1.3)

cat(sprintf("Lower total cost: %s (saves $%s)\n", better, abs(cheap_tco - durable_tco)))
cat(sprintf("Cheap option  %d purchase(s) x $%s = $%s\n", cheap_purchases, cheap_price, cheap_tco))
cat(sprintf("Durable option  %d purchase(s) x $%s = $%s\n", durable_purchases, durable_price, durable_tco))
cat(sprintf("Units to landfill, cheap  %d\n", cheap_purchases))
cat(sprintf("Units to landfill, durable  %d\n", durable_purchases))
```

</details>

<details>
<summary>Where does the cheap option actually win?</summary>

Shorten the horizon (you only need the product for a year), or make the durable
version's premium extreme relative to its extra life, and "cheap" wins on cost.
But the **waste count** almost always favours durable: five landfilled units
versus one. Whole-life thinking means putting *both* numbers --- dollars and
units --- in front of the decision-maker, not just the sticker price.

</details>

## Sustainability in project management

Everything above was about designing a *part*. When you are running a *project*,
sustainability becomes a set of management actions spread across the whole life
cycle. The MGMT-3105 checklist:

**Procurement and contracts**

- Choose **eco-friendly and ethical suppliers**; audit them for compliance.
- Put **sustainability clauses** in contracts.
- Perform an **Environmental & Social Impact Assessment (ESIA)**.

**Design and materials**

- Minimize raw-material use; prioritize **recyclable, renewable, and local** resources.
- Design for **energy and water efficiency**.
- Apply **circular-economy** principles: reuse, refurbish, recycle.
- Plan the **product/service lifecycle** --- what happens to it after the project ends.

**Operations**

- Reduce **carbon emissions** from transport and operations.
- Use **digital tools** to cut paper and travel.
- Run proper **waste management**: segregation, recycling, safe disposal.

**People and governance**

- Safeguard **health, safety, and well-being** of the team and the community.
- Support **social responsibility**: local hiring, fair wages, training.
- Track **sustainability KPIs**: carbon footprint, waste-reduction %, energy saved.
- Monitor suppliers and contractors; **identify and mitigate** environmental and social risks; run **sustainability audits** when needed.

```{r tryit-checklist-01, echo=FALSE, results='asis'}
check_bins <- c("Procurement and contracts", "Design and materials", "Operations", "People and governance")
tryit_sort(
  "sustainability-checklist", "Try it yourself: where does each action belong on the checklist?",
  bins = check_bins,
  cards = data.frame(
    text = c("Add a clause requiring suppliers to report CO2e per part",
             "Require a sustainability audit before approving a new supplier",
             "Redesign the housing as a single polymer so it can be recycled",
             "Specify 30 % recycled content in the packaging",
             "Ship by consolidated sea freight instead of air",
             "Replace paper travellers on the line with tablets",
             "Segregate scrap aluminium and plastic into labelled bins",
             "Report the waste-diversion rate on the monthly KPI sheet",
             "Hire apprentices from the local community college",
             "Brief the team on the new recycling station's safety rules"),
    bin = c(rep("Procurement and contracts", 2), rep("Design and materials", 2),
            rep("Operations", 3), rep("People and governance", 3)),
    why = c("Sustainability clauses belong in contracts.", "Auditing suppliers is a procurement action.",
            "Design for recyclability is a design decision.", "Recycled content is a material choice.",
            "Reducing transport emissions is an operations action.", "Digital tools cut paper: operations.",
            "Waste management (segregation, recycling) is operations.",
            "Tracking sustainability KPIs is governance.", "Local hiring is social responsibility.",
            "Health and safety of the team is people and governance."),
    stringsAsFactors = FALSE),
  intro = "The MGMT-3105 checklist has four groups. Pick one for every action.", question = NULL)

tryit_sort(
  "checklist-twin", "Sustainability checklist (graded)",
  bins = check_bins, show = FALSE,
  cards = data.frame(
    text = c("Write a take-back obligation into the packaging contract",
             "Prefer a locally sourced steel grade",
             "Switch plant lighting to LED with occupancy sensors",
             "Publish carbon per unit to the steering committee",
             "Choose a supplier with an ISO 14001 certificate",
             "Run a sustainability audit of the contractor on site"),
    bin = c("Procurement and contracts", "Design and materials", "Operations",
            "People and governance", "Procurement and contracts", "People and governance"),
    stringsAsFactors = FALSE),
  question = "<p>For each project action, choose the group of the MGMT-3105 sustainability checklist it belongs to.</p>")
```

### The circular economy

The circular economy is the alternative to the linear **take --- make --- use ---
dispose** model. Instead, material is kept in use through loops of increasing
size: maintain and reuse first, then refurbish and remanufacture, then recycle,
with disposal as the last resort.

```{r fig-circular-economy-01, fig.cap="The circular economy: keep materials circulating through the tightest loop that still works. Inner loops (maintain, reuse) retain the most value; recycling is the outermost loop before disposal.", fig.align='center', fig.width=9, fig.height=7}
stages <- data.frame(
  label = stringr::str_wrap(c("Design", "Make", "Distribute", "Use", "Maintain / Reuse",
            "Refurbish / Remanufacture", "Recycle"), width = 12),
  angle = seq(90, 90 - 360 + 360/7, length.out = 7) * pi / 180
)
stages$x <- cos(stages$angle)
stages$y <- sin(stages$angle)

seg <- data.frame(
  x    = stages$x,
  y    = stages$y,
  xend = stages$x[c(2:7, 1)],
  yend = stages$y[c(2:7, 1)]
)

ggplot() +
  geom_curve(data = seg, aes(x = x, y = y, xend = xend, yend = yend),
             curvature = -0.25, linewidth = 0.9, color = "#27AE60",
             arrow = arrow(length = unit(0.18, "cm"), type = "closed")) +
  geom_point(data = stages, aes(x = x, y = y), size = 20, color = "#EAF6EC") +
  geom_point(data = stages, aes(x = x, y = y), size = 20, shape = 1, color = "#27AE60") +
  geom_text(data = stages, aes(x = x, y = y, label = label),
            size = 3 * diagram_text_scale, fontface = "bold",
            lineheight = 0.9, color = "grey15") +
  annotate("text", x = 0, y = 0, label = "keep materials\nin use",
           size = 3.2 * diagram_text_scale, fontface = "italic", color = "grey40") +
  coord_equal(xlim = c(-1.4, 1.4), ylim = c(-1.4, 1.4)) +
  labs(title = "Design out waste; circulate materials") +
  theme_void() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale,
                                  face = "bold", hjust = 0.5))
```

```{r video-circular-economy, echo=FALSE, results='asis'}
embed_youtube("zCRKvDyyHmI", "Explaining the Circular Economy (Ellen MacArthur Foundation)")
```

### Try it yourself: a sustainability KPI snapshot

Sustainability only gets managed if it gets measured. The live cell computes a
few of the KPIs from the checklist for the RC-vehicle project: carbon footprint
per unit, waste-diversion rate, and recycled-content share. Adjust the inputs to
your own project assumptions.

```{r calc-sustainability-kpi-01, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "sustainability-kpi-01",
  title = "Sustainability KPI Snapshot",
  icon = "seedling",
  inputs = list(
    qwebr_calc_input("units", "Units produced", value = 120000, min = 10000, max = 500000, step = 10000, type = "slider"),
    qwebr_calc_input("material_co2_t", "Materials (cradle-to-gate)", value = 71, min = 0, max = 300, step = 1, suffix = "t CO2e"),
    qwebr_calc_input("process_co2_t", "Moulding + assembly energy", value = 18, min = 0, max = 100, step = 1, suffix = "t CO2e"),
    qwebr_calc_input("transport_co2_t", "Inbound + outbound freight", value = 26, min = 0, max = 100, step = 1, suffix = "t CO2e"),
    qwebr_calc_input("process_waste_t", "Scrap + packaging generated", value = 9.0, min = 0.1, max = 50, step = 0.1, suffix = "t"),
    qwebr_calc_input("waste_diverted_t", "Reused / recycled / to-energy", value = 8.1, min = 0, max = 50, step = 0.1, suffix = "t"),
    qwebr_calc_input("recycled_input_t", "Recycled feedstock", value = 15.5, min = 0, max = 100, step = 0.5, suffix = "t"),
    qwebr_calc_input("total_input_t", "Total material input", value = 62.0, min = 1, max = 200, step = 0.5, suffix = "t")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: sustainability-kpi-01
units            <- 120000

# Carbon
material_co2_t   <- 71      # tonnes CO2e, cradle-to-gate materials
process_co2_t    <- 18      # tonnes CO2e, moulding + assembly energy
transport_co2_t  <- 26      # tonnes CO2e, inbound + outbound freight

# Waste
process_waste_t  <- 9.0     # tonnes of scrap + packaging generated
waste_diverted_t <- 8.1     # tonnes reused / recycled / to-energy

# Materials
recycled_input_t <- 15.5    # tonnes of recycled feedstock
total_input_t    <- 62.0    # tonnes of material input

total_co2_t <- material_co2_t + process_co2_t + transport_co2_t
diversion   <- 100 * waste_diverted_t / process_waste_t

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
bp <- barplot(c(Materials = material_co2_t, Process = process_co2_t, Transport = transport_co2_t),
              col = c("#2c3e50", "#2980b9", "#16a085"), ylab = "t CO2e", main = "Carbon footprint by source")

cat(sprintf("Waste KPI status: %s\n", if (diversion >= 90) "PASS" else "!!BELOW TARGET!!"))
cat(sprintf("Waste diversion rate  %.1f %% (target >= 90 %%)\n", diversion))
cat(sprintf("Landfilled waste  %.2f t\n", process_waste_t - waste_diverted_t))
cat(sprintf("Carbon footprint  %.1f t CO2e\n", total_co2_t))
cat(sprintf("Carbon per unit  %.3f kg CO2e\n", total_co2_t * 1000 / units))
cat(sprintf("Recycled content  %.1f %% of input\n", 100 * recycled_input_t / total_input_t))
```

</details>

## Key Takeaways

- **Sustainability** = meeting present needs without compromising future generations' ability to meet theirs. **Sustainable design** applies that across social, economic, and ecological pillars at once --- a "win" on one pillar is not a win.
- The four levers for a product engineer: **recycle, bio-source/biodegrade, reduce material, design for recyclability.**
- The **seven resin codes** tell a sorter what a plastic is; PET (#1), HDPE (#2), and PP (#5) recycle well, #7 (blends/PLA/filled) rarely does. **Post-consumer** recyclate diverts real waste but is hard to process; **post-industrial** regrind is easy.
- **Recyclable and biodegradable can conflict**, and "biodegradable" does little in a dry, compacted landfill --- the EcoPlast case shows what happens when that asterisk is ignored.
- A small **material-reduction** decision, multiplied by the production run, moves tonnes of material, energy, and CO2e. The "greenest" material depends on **where in the life cycle the impact lands** --- use phase vs. production vs. end of life.
- Consumers give ground on **price and cosmetics** but not on **performance or durability**; **whole-life cost** usually favours the durable design on both dollars and landfill count.
- In **project management**, sustainability is a checklist across procurement, design, operations, and governance --- backed by **circular-economy** loops and tracked with **KPIs** (carbon per unit, waste diversion, recycled content).

## Case Study: Meridian R/C — Week 1

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 1: The Body & Frame Material Decision"}

Meridian R/C Vehicles is racing a deadline neither you nor your sponsor can
move. Kestrel Retail Group, a national toy retailer, has told Daniel Cho,
Meridian's VP of Product, that if the **Meridian Trailblazer** — a
1/18-scale rally-truck replica — passes durability testing and toy-safety
certification, Kestrel's Category Buyer, Renata Ford, will place a
**100,000+ unit** production order in time for next year's holiday season.
Before a single tool is cut, you and your manufacturing engineer, **Priya
Nandakumar**, have to settle the one decision that shapes every cost, mass,
and end-of-life number that follows: what is the Trailblazer's body and
frame made of?

Priya frames it as a three-way trade study. Steel bodies would weigh about
140 g each and cost roughly $1.20/kg in raw material — cheap, but steel is
the heaviest of the three candidates and, for this die-cast toy geometry,
the least attractive to recycle back into value. Carbon-fibre composite is
the opposite extreme: a 38 g body, the best strength-to-weight ratio by
far, but at roughly $28/kg in material alone — and once a thermoset
carbon-fibre part reaches end of life in a child's toy box, there is no
practical way to recycle it. Aluminium lands in the middle on every axis:
about 70 g per body, $3/kg, and — critically — genuinely recyclable at
scale.

| Candidate | Body mass | Material cost | End-of-life recyclability |
|---|---|---|---|
| Steel | 140 g | $1.20/kg | Heavy; low recyclate value for this design |
| **Aluminium (chosen)** | 70 g | $3.00/kg | High — genuinely recyclable |
| Carbon fibre | 38 g | $28.00/kg | Very poor — thermoset, effectively unrecyclable |

You, Priya, procurement lead **Marcus Webb**, and quality engineer **Sofia
Reyes** agree: aluminium is the call for both the body and the frame. It is
not the lightest option and it is not the cheapest raw material, but it is
the only one that clears the bar on cost *and* recyclability at once —
exactly the kind of "no pillar loses" decision the sustainability framework
asks for. Carbon fibre does not disappear from the story, though. **Fibro
Source**, a carbon-fibre supplier in Sherbrooke, Quebec that is
mid-restructuring after a recent French acquisition, stays on the supplier
list — not for the core prototype, but as an optional carbon-fibre trim and
spoiler package Kestrel could offer later as a premium upsell. Daniel Cho
signs off: aluminium body and frame, carbon fibre trim only if the business
case for it ever appears.

:::

## Practice Problems

<details>
<summary>Problem 1: Packaging substitution (food industry)</summary>

A food-processing client ships 2.4 million meal trays a year. The current tray
is 22 g of black CPET (#1, but often rejected by sorters because black plastic
is invisible to optical sorting). A proposed natural-coloured PP (#5) tray
weighs 19 g, costs \$0.006/g versus \$0.005/g for the CPET, and is accepted by
the client's regional recycler.

(a) What is the annual change in material mass and material cost?
(b) Give two non-cost arguments for switching, and one argument against.

<details>
<summary>Solution</summary>

```{r p1-packaging-solution-01, echo=TRUE}
trays        <- 2.4e6
mass_cpet_g  <- 22
mass_pp_g    <- 19
cost_cpet    <- 0.005
cost_pp      <- 0.006

mass_change_kg <- (mass_pp_g - mass_cpet_g) / 1000 * trays
cost_change    <- (mass_pp_g * cost_pp - mass_cpet_g * cost_cpet) * trays

cat("Annual material mass change :", format(round(mass_change_kg), big.mark = " "), "kg",
    "(a reduction)\n")
cat("Annual material cost change : ",
    sprintf("$%s", format(round(cost_change), big.mark = " ")),
    " (an increase)\n", sep = "")
```

(a) About 7 200 kg less plastic per year, but roughly \$9 600/yr more spent on
material because PP costs more per gram here and the 3 g saving does not fully
offset it.

(b) *For:* the natural-coloured PP is actually recovered by the recycler
(black CPET usually is not), and PP trays tolerate reheating better. *Against:*
higher material spend, and a tooling change is needed --- capital plus a
qualification run.

</details>

</details>

<details>
<summary>Problem 2: Recycled-aluminium body (automotive/toy)</summary>

Switching the RC-car body from primary to **recycled** aluminium keeps the mass
at 70 g/unit but changes the cradle-to-gate energy from 155 MJ/kg to 8 MJ/kg and
the carbon from 11 kg CO2e/kg to 0.6 kg CO2e/kg. Recycled stock costs \$0.25/kg
more. Run size 120 000 units.

Estimate the energy saved, the CO2e avoided, and the extra material cost.

<details>
<summary>Solution</summary>

```{r p2-recycled-al-solution-01, echo=TRUE}
mass_kg   <- 70 / 1000
units     <- 120000
kg_total  <- mass_kg * units

energy_saved <- kg_total * (155 - 8)
co2_saved    <- kg_total * (11 - 0.6)
extra_cost   <- kg_total * 0.25

cat("Aluminium in the run :", format(round(kg_total), big.mark = " "), "kg\n")
cat("Production energy saved :", format(round(energy_saved), big.mark = " "), "MJ",
    sprintf("(~%s kWh)\n", format(round(energy_saved / 3.6), big.mark = " ")))
cat("CO2e avoided :", format(round(co2_saved), big.mark = " "), "kg",
    sprintf("(~%.1f t)\n", co2_saved / 1000))
cat("Extra material cost :",
    sprintf("$%s\n", format(round(extra_cost), big.mark = " ")))
```

Roughly 1.2 million MJ (about 340 000 kWh) and 87 t of CO2e avoided for about
\$2 100 of extra material cost --- one of the highest-leverage sustainability
moves available on this project.

</details>

</details>

<details>
<summary>Problem 3: Design-for-recyclability audit (defense)</summary>

A defense client requires that field-returned electronic enclosures be
**materially recoverable**. You are handed this enclosure: glass-filled nylon
(#7) housing, 6 zinc-plated steel screws, a bonded aluminium RFI shield, a
polyurethane potting compound over the PCB, and two adhesive asset-tracking
labels.

List, in priority order, the four changes that would most improve material
recovery, and note the functional risk of each.

<details>
<summary>Model answer</summary>

1. **Replace the potting compound** with a removable conformal coating or a
   gasketed enclosure. *Risk:* reduced shock/moisture protection --- must
   re-qualify to the environmental spec.
2. **Eliminate the bonded aluminium shield**; use a conductive coating on the
   nylon or a snap-in shield of the same-family metal. *Risk:* RFI performance
   must be re-verified.
3. **Swap the 6 steel screws** for moulded snap features or same-metal
   fasteners. *Risk:* fewer re-assembly cycles; captive-hardware requirement may
   forbid it.
4. **Drop adhesive labels** for moulded-in or laser-etched marking. *Risk:*
   contrast/legibility and traceability sign-off.

The housing polymer itself (#7 glass-filled) is hard to change if the structural
spec depends on it --- note it as a known limitation rather than a quick fix.

</details>

</details>

<details>
<summary>Problem 4 (discussion): planned obsolescence vs. sustainable design</summary>

"Sustainable design requires that our products last as long as possible. But
business decisions may deliberately shorten product life to protect repeat
sales --- planned obsolescence. Which do you prefer, and how would you reconcile
the two in a company that has to stay profitable?"

<details>
<summary>Discussion</summary>

There is no single right answer --- this is a position you argue. Strong answers
usually land on **service and circular business models**: the manufacturer keeps
revenue through repair, upgrades, refurbishment, take-back, or
product-as-a-service, so "make it last" and "keep earning" stop being opposites.
Weak answers just pick a side and ignore the constraint on the other.

</details>

</details>

## Further Resources

- **U.S. EPA --- Learn About Sustainability:** <https://www.epa.gov/sustainability/learn-about-sustainability>
- **Brundtland Report (*Our Common Future*, 1987):** <https://sustainabledevelopment.un.org/content/documents/5987our-common-future.pdf>
- **Ellen MacArthur Foundation --- Circular Economy Introduction:** <https://www.ellenmacarthurfoundation.org/topics/circular-economy-introduction/overview>

```{r video-resources-01, echo=FALSE, results='asis'}
embed_youtube("9GorqroigqM", "The Story of Stuff")
embed_youtube("zCRKvDyyHmI", "Explaining the Circular Economy (Ellen MacArthur Foundation)")
embed_youtube("7V8oFI4GYMY", "What Is Sustainable Development?")
```
