---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Introduction to Project Management and Communication

```{r setup-02, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Distinguish** a project from a process, and explain why project work needs its own discipline.
2. **Describe** the triple constraint (scope, time, cost) and its relationship to quality and resources.
3. **Identify** the five process phases of a project and what happens in each.
4. **Apply** the project manager's decision hierarchy and name the four pillars of the PMI Code of Ethics.
5. **Build** a project communication plan and choose the right method (formal/informal, written/verbal) for a message.
6. **Calculate** the number of communication channels on a team and explain why they grow faster than headcount.
7. **Apply** active- and effective-listening techniques and recognize the listening filters that distort a message.
8. **Select** the appropriate performance report (status, progress, trend, forecast, variance, earned value) for a stakeholder need.

<details>
<summary>Before we begin: think of a group assignment that went badly</summary>

Was the problem that nobody knew *what* to do, or that people knew but never
told each other? Most project failures are communication failures wearing a
different coat. Keep that example in mind.

</details>

## Introduction: only one project in four

Here is a sobering number that opens this course: by common industry estimates,
only about **26 % of projects** are delivered successfully while respecting the
triple constraint --- on scope, on time, on budget. The other three in four slip,
overspend, descope, or fail outright.

That is why project management exists as a discipline. Running the RC-vehicle
launch from the last chapter is not the same as running a production line. A
line is a **process**: repetitive, permanent, tuned once and then kept stable. A
**project** is a one-time push toward a unique goal, under a deadline and a
budget, and it needs different tools and a different mindset.

::: {.callout-important}
**Case study --- Orion Tech (MGMT-3105, Week 2).** A software company built a
healthcare app. The team was skilled, but: developers were never clear on
requirements and reworked constantly; the PM ran everything through email and
buried the team; stakeholders got no status reports; a few voices dominated
meetings while others went silent; and tone and urgency were misread between the
local team and offshore developers. Deadlines slipped, the budget grew, and the
client's satisfaction collapsed. Leadership asked whether the PM had followed
any communication protocol at all.
:::

```{r video-what-is-pm-02, echo=FALSE, results='asis'}
embed_youtube("Jk-JwtScIlw", "What Is Project Management? (Association for Project Management)")
```

## Project versus process

| | **Project** | **Process** |
|---|---|---|
| Duration | Temporary, one-time | Permanent, ongoing |
| Work | Unique, complex, connected activities toward one goal | Repetitive, designed to be repeated |
| Ends when | The goal is met (by a specific time, within budget, to spec) | It doesn't --- it is improved and kept running |
| Example on our project | *Designing and launching the RC-vehicle replica* | *Moulding 100 000 bodies once the line is running* |

Projects help companies cope with **shortened product life cycles**, **narrow
launch windows**, **increasingly complex products**, and **global markets**. The
RC-vehicle brief has all four: a holiday launch window, a technical product, and
suppliers on two continents.

## The triple constraint

Every project is boxed in by **scope**, **time**, and **cost**, with **quality**
sitting in the middle as the measure of whether stakeholders are satisfied.
Resources (people, equipment, facilities) are the fourth practical limit.

```{r fig-triple-constraint-02, fig.cap="The triple constraint. Scope, time, and cost form the sides; quality is the area they enclose. Change one side and at least one other must move, or quality suffers.", fig.align='center', fig.width=8, fig.height=7}
tri <- data.frame(
  x = c(0, 2, 4, 0),
  y = c(0, 3.4, 0, 0)
)
ggplot(tri, aes(x, y)) +
  geom_polygon(fill = "#5DADE2", alpha = 0.25, color = "#2E86C1", linewidth = 1.1) +
  annotate("text", x = 1, y = 1.9, label = "SCOPE", angle = 60,
           fontface = "bold", size = 4.2 * diagram_text_scale, color = "#1B4F72") +
  annotate("text", x = 3, y = 1.9, label = "TIME", angle = -60,
           fontface = "bold", size = 4.2 * diagram_text_scale, color = "#1B4F72") +
  annotate("text", x = 2, y = -0.28, label = "COST",
           fontface = "bold", size = 4.2 * diagram_text_scale, color = "#1B4F72") +
  annotate("text", x = 2, y = 1.15, label = "QUALITY",
           fontface = "bold", size = 4 * diagram_text_scale, color = "#B03A2E") +
  annotate("text", x = 2, y = 0.72,
           label = "(stakeholder satisfaction)", fontface = "italic",
           size = 3 * diagram_text_scale, color = "grey35") +
  coord_equal(xlim = c(-0.6, 4.6), ylim = c(-0.7, 3.8)) +
  labs(title = "Pick two to fix; the third must flex") +
  theme_void() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale,
                                  face = "bold", hjust = 0.5))
```

<details>
<summary>Think about it: the retailer moves the RC-vehicle launch four weeks earlier. Which constraint gives?</summary>

Time just shrank. To hold the launch you must either **cut scope** (fewer body
options, defer a feature), **add cost** (overtime, expedited tooling, a second
supplier), or **accept lower quality** (less testing) --- and that last one is how
projects like Orion Tech's die. The PM's job is to make that trade explicit and
get a decision, not to quietly absorb it.

</details>

### Try it yourself: put a number on each side of the triangle

"Which constraint gives" is a much easier decision once each option has a
dollar figure attached. Live cell --- change the weeks compressed, the costs,
or the failure probability and see how the three options compare.

```{r calc-triple-constraint-tradeoff-02, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "triple-constraint-tradeoff-02",
  title = "Triple-Constraint Trade-off Calculator",
  icon = "triangle-exclamation",
  inputs = list(
    qwebr_calc_input("weeks_compressed", "Weeks to compress", value = 4, min = 1, max = 12, step = 1, type = "slider"),
    qwebr_calc_input("overtime_cost_per_week", "Overtime cost", value = 9000, min = 1000, max = 30000, step = 500, prefix = "$", suffix = "/ wk"),
    qwebr_calc_input("expedite_fee", "Expedite fee", value = 15000, min = 0, max = 100000, step = 1000, prefix = "$"),
    qwebr_calc_input("colourway_revenue_at_risk", "Revenue lost by cutting scope", value = 62000, min = 0, max = 300000, step = 1000, prefix = "$"),
    qwebr_calc_input("skip_test_prob_failure", "Chance a defect escapes", value = 0.08, min = 0, max = 1, step = 0.01),
    qwebr_calc_input("field_failure_cost", "Cost if it does", value = 240000, min = 0, max = 1000000, step = 10000, prefix = "$")
  ),
  goal = "Find the chance of a defect escaping at which skipping the test stops being the cheapest option.",
  goal_test = "expected_quality_cost > min(cost_of_holding_scope, colourway_revenue_at_risk)"
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: triple-constraint-tradeoff-02
weeks_compressed <- 4

# Option A: hold scope -- pay to compress the schedule
overtime_cost_per_week <- 9000
expedite_fee           <- 15000
cost_of_holding_scope  <- weeks_compressed * overtime_cost_per_week + expedite_fee

# Option B: hold cost -- cut scope (drop one body colourway)
colourway_revenue_at_risk <- 62000   # forecast contribution margin lost

# Option C: hold cost and scope -- accept lower quality (skip one test cycle)
skip_test_prob_failure <- 0.08       # chance a defect reaches the customer
field_failure_cost     <- 240000     # recall/rework cost if it does
expected_quality_cost  <- skip_test_prob_failure * field_failure_cost

fmt <- function(x) format(round(x), big.mark = " ", scientific = FALSE, trim = TRUE)

costs <- c(A = cost_of_holding_scope, B = colourway_revenue_at_risk, C = expected_quality_cost)
cheapest <- c(A = "A, pay to compress", B = "B, cut scope", C = "C, skip a test cycle")[[names(which.min(costs))]]

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
bp <- barplot(costs / 1000, names.arg = c("A: compress", "B: cut scope", "C: cut testing"),
              col = c("#2980b9", "#e67e22", "#8e44ad"), ylim = c(0, max(costs / 1000) * 1.2),
              ylab = "Cost of flexing ($ thousand)", main = "What each side of the triangle costs")
text(bp, costs / 1000, labels = paste0("$", round(costs / 1000, 1), "k"), pos = 3, cex = 1.3)

cat(sprintf("Cheapest side to flex: Option %s\n", cheapest))
cat(sprintf("Option A: pay to compress  $%s\n", fmt(cost_of_holding_scope)))
cat(sprintf("Option B: cut scope  $%s revenue lost\n", fmt(colourway_revenue_at_risk)))
cat(sprintf("Option C: skip a test cycle  $%s expected cost\n", fmt(expected_quality_cost)))
cat(sprintf("Option C risk  %.0f%% chance of a $%s failure\n",
            skip_test_prob_failure * 100, fmt(field_failure_cost)))
```

</details>

## The five process phases

```{r fig-project-lifecycle-02, fig.cap="Effort and resource use across the project life cycle. Little happens during initiation, effort peaks through execution and control, and it falls away at closing.", fig.align='center', fig.width=10, fig.height=5}
phase_x <- seq(0, 10, length.out = 200)
effort  <- dnorm(phase_x, mean = 6, sd = 2.2)
df <- data.frame(x = phase_x, effort = effort / max(effort))

ggplot(df, aes(x, effort)) +
  geom_area(fill = "#5DADE2", alpha = 0.4) +
  geom_line(color = "#2E86C1", linewidth = 1.1) +
  annotate("segment", x = c(2, 4, 8, 9.2), xend = c(2, 4, 8, 9.2),
           y = 0, yend = 1.05, linetype = "dotted", color = "grey55") +
  annotate("text", x = c(1, 3, 6, 8.6, 9.7), y = 1.12,
           label = c("Initiating", "Planning", "Executing", "Controlling", "Closing"),
           size = 3.2 * diagram_text_scale, fontface = "bold") +
  coord_cartesian(ylim = c(0, 1.2)) +
  labs(title = "Project management process life cycle",
       x = "Project timeline", y = "Resources and effort") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale)
  )
```

| Phase | Key work |
|---|---|
| **1. Initiating** | Business case, feasibility, project sponsors, **project charter** + sign-off. |
| **2. Planning** | Scope statement, WBS, network diagram (AON), budget, risk plan, quality standards, control system, communication protocols. *This phase decides success or failure.* |
| **3. Executing** | Execute the plan, monitor progress, quality assurance. The PM shifts from doing to leading. |
| **4. Controlling** | Scope/change/schedule/budget/risk control become daily; overlaps execution. |
| **5. Closing** | Final review of budget, schedule, scope; customer sign-off; **lessons learned**; release resources; final payments. The hardest phase to finish well. |

### The project manager's decision hierarchy

When decisions conflict, this is the order of consideration:

1. **Safety** (always first)
2. What is best for **the company I represent**
3. What is best for **the project sponsor**
4. What is best for **the customer**
5. What is best for **the suppliers**

And the **PMI Code of Ethics** rests on four pillars: **responsibility, respect,
fairness, and honesty**. You can violate ethics without breaking a law --- but as
an accredited PM you would swear to uphold the code, and breaking it puts your
professional standing at risk.

## Project communications

Communication is not a side task. **About 90 % of a project manager's time is
spent communicating.** The three communication processes are **plan**, **manage**,
and **control**.

### The communication plan

A written communication plan is built from the stakeholders' information needs.
Typical elements:

- What information is collected, and when
- Who receives it
- How it is gathered and stored
- Limits, if any, on who may tell whom
- Reporting relationships
- Contact information for all stakeholders
- The distribution schedule for each type of communication

<details>
<summary>Orion Tech fix: what should have been in the plan on day one?</summary>

A stakeholder list with each party's information needs; the report types (status,
cost, risk, performance) and their frequency; the mix of channels (email for the
record, chat for quick updates, dashboards for tracking, video for
collaboration); and **feedback loops** so every team member confirms they
understood the requirements. None of that existed --- so requirements were guessed
at and stakeholders were left in the dark.

</details>

### The communication model

Every communication has three elements: a **sender**, a **message**, and a
**receiver**. The sender **encodes** the message based on the receiver's
education, experience, language, and culture, picks the right method, and
**confirms it was understood**. The receiver **decodes** carefully and checks
their understanding. **Feedback** ("Does that make sense?", a nod, a clarifying
question) closes the loop.

### Communication channels grow faster than the team

Add one person to a team and you do not add one relationship --- you add one for
every person already there. The number of two-way channels on a team of $n$
people is:

$$\text{channels} = \frac{n(n-1)}{2}$$

```{r fig-channels-02, fig.cap="Two-way communication channels versus team size. Doubling the team roughly quadruples the channels that have to be kept coherent.", fig.align='center', fig.width=9, fig.height=5}
n <- 2:15
ch <- n * (n - 1) / 2
df <- data.frame(n, ch)

ggplot(df, aes(n, ch)) +
  geom_col(fill = "#5DADE2", width = 0.7) +
  geom_text(aes(label = ch), vjust = -0.4, size = 3.3 * diagram_text_scale) +
  scale_x_continuous(breaks = n) +
  labs(title = "Communication channels = n(n - 1) / 2",
       x = "People on the team (n)", y = "Two-way channels") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Worked Example 1: channels on the RC-vehicle project

The core project has the PM, 4 team members, an accountant, and a sponsor liaison
--- **7 people**. The extended stakeholder circle adds 2 supplier contacts, a
retailer contact, and a compliance officer --- **11 people**.

Live cell --- try different core-team and extended-circle sizes.

```{r calc-channels-example-02, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "channels-example-02",
  title = "Communication Channels Calculator",
  icon = "diagram-project",
  inputs = list(
    qwebr_calc_input("core", "Core team size", value = 7, min = 2, max = 30, step = 1),
    qwebr_calc_input("extended", "Extended circle size", value = 11, min = 2, max = 40, step = 1, type = "slider")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: channels-example-02
core <- 7
extended <- 11

ch_core     <- core * (core - 1) / 2
ch_extended <- extended * (extended - 1) / 2

n <- 2:40
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
plot(n, n * (n - 1) / 2, type = "l", lwd = 4, col = "#2c3e50", xlab = "People in the group", ylab = "Communication channels",
     main = "Channels grow with the square of group size")
points(c(core, extended), c(ch_core, ch_extended), pch = 19, cex = 2.2, col = c("#2980b9", "#e74c3c"))

cat(sprintf("Widening the circle from %d to %d adds %d channels\n", core, extended, ch_extended - ch_core))
cat(sprintf("Core team channels  %d\n", ch_core))
cat(sprintf("Extended circle channels  %d\n", ch_extended))
```

</details>

Four extra people did not add four conversations --- they added 34. That is the
argument for a **hub-and-spoke** communication plan (routes defined, the PM as
the hub) instead of letting everyone talk to everyone about everything.

### Try it yourself: channel growth and the cost of one more stakeholder

```{r calc-channels-calc-02, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "channels-calc-02",
  title = "Cost of One More Stakeholder",
  icon = "user-plus",
  inputs = list(
    qwebr_calc_input("team_size", "Baseline team size", value = 9, min = 2, max = 30, step = 1, type = "slider")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: channels-calc-02
team_size <- 9

channels <- function(n) n * (n - 1) / 2

sizes <- team_size + 0:4
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
plot(2:40, channels(2:40), type = "l", lwd = 4, col = "#bdc3c7", xlab = "People in the group", ylab = "Communication channels",
     main = "Each added person costs more than the last")
points(sizes, channels(sizes), pch = 19, cex = 2.2, col = c("#2980b9", rep("#e74c3c", 4)))

cat(sprintf("Adding person #%d alone creates %d new channels\n",
            team_size + 1, channels(team_size + 1) - channels(team_size)))
for (extra in 0:4) {
  n <- team_size + extra
  cat(sprintf("Team of %d  %d channels", n, channels(n)))
  if (extra > 0) cat(sprintf(" (+%d vs. baseline)", channels(n) - channels(team_size)))
  cat("\n")
}
```

</details>

### Try it yourself: a merger by hand

A team is about to take on extra people. Work out how much the coordination
load grows. Each team picks its own **data set**.

```{r tryit-channels-02, echo=FALSE, results='asis'}
ch_fun <- function(n) n * (n - 1) / 2
ch_sets <- list(c(core = 6, added = 3), c(core = 8, added = 4), c(core = 5, added = 5), c(core = 10, added = 2))
ch_given <- function(v) c("Current team size" = v[["core"]], "People joining" = v[["added"]])
ch_answers <- function(v) {
  b <- ch_fun(v[["core"]]); a <- ch_fun(v[["core"]] + v[["added"]])
  list(tryit_answer("Channels before", b, tol = 0),
       tryit_answer("Channels after", a, tol = 0),
       tryit_answer("Increase in headcount", 100 * v[["added"]] / v[["core"]], tol = 0.5, suffix = "%"),
       tryit_answer("Increase in channels", 100 * (a / b - 1), tol = 0.5, suffix = "%"))
}
tryit_number(
  "channels-by-hand", "Two-way channels before and after a merger",
  answers = ch_answers(ch_sets[[1]]), given = ch_given(ch_sets[[1]]),
  teams = lapply(ch_sets[-1], function(v) list(given = ch_given(v), answers = ch_answers(v))),
  intro = "Channels = n(n - 1) / 2. Percentages to the nearest whole number.")

tryit_number(
  "channels-twin", "Channels after a merger (graded)", show = FALSE,
  answers = ch_answers(c(core = 7, added = 4)),
  question = paste0("<p>A project team of 7 people is merged with 4 supplier engineers, all becoming full ",
    "team members. Calculate the number of two-way channels before and after, the percentage increase in ",
    "headcount, and the percentage increase in channels (nearest whole number).</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

Channels = n(n - 1)/2. For data set 1: before, 6 x 5 / 2 = 15; after,
9 x 8 / 2 = 36. Headcount rises 50 % but channels rise 140 %: adding people
grows the load faster than the headcount.

</details>

## Listening: the mode we use most and understand least

The four modes of communication, by share of a typical day and by how well they
are scientifically understood:

| Mode | Share | Notes |
|---|---:|---|
| Listening | ~55 % | The most used, the **least** studied and taught |
| Speaking | ~23 % | |
| Reading | ~13 % | |
| Writing | ~9 % | |

And most of the meaning is **nonverbal**: roughly **93 %** of person-to-person
communication is carried by body language (~55 %) and tone of voice (~38 %) ---
and tone's share rises toward 86 % on the phone, where body language is gone.

- **Active listening:** the receiver signals engagement --- agreeing, asking for clarification.
- **Effective listening:** watching the speaker for gesture and expression, thinking before responding, asking questions, repeating back, giving feedback.

### The listening filters

Every message passes through three filters before it lands:

- **Selective attention** --- we choose what to focus on.
- **Selective interpretation** --- we impose our own meaning (shaped by gender, generation, geography, culture, education, occupation).
- **Selective retention** --- we choose what to keep, and for how long.

There is also a raw arithmetic problem. People **listen** at about 500 words per
minute but **speak** at about 150. That leaves ~350 words per minute of spare
capacity --- room to tune out.

### Try it yourself: the tune-out budget

```{r calc-tune-out-02, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "tune-out-02",
  title = "Tune-Out Budget Calculator",
  icon = "headphones",
  inputs = list(
    qwebr_calc_input("listen_wpm", "Listening speed", value = 500, min = 250, max = 900, step = 50, suffix = "words / min"),
    qwebr_calc_input("speak_wpm", "Speaking speed", value = 150, min = 50, max = 250, step = 10, suffix = "words / min"),
    qwebr_calc_input("briefing_minutes", "Briefing length", value = 12, min = 1, max = 60, step = 1, type = "slider")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: tune-out-02
listen_wpm <- 500
speak_wpm  <- 150
briefing_minutes <- 12

spare_wpm    <- listen_wpm - speak_wpm
spare_words  <- spare_wpm * briefing_minutes
# Time the listener could "leave" if they mentally processed at listening speed:
tune_out_seconds_per_min <- spare_wpm / listen_wpm * 60

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
bp <- barplot(c(Speaking = speak_wpm, Listening = listen_wpm), col = c("#2980b9", "#16a085"),
              ylab = "Words per minute", main = "The gap the listener's mind fills")
segments(bp[1], speak_wpm, bp[2], listen_wpm, lty = 2, lwd = 3, col = "#e74c3c")
text(mean(bp), (speak_wpm + listen_wpm) / 2 + 20, labels = paste(spare_wpm, "spare"), cex = 1.3, col = "#e74c3c")

cat(sprintf("Spare listening capacity: %d words/min\n", spare_wpm))
cat(sprintf("Spare words over the briefing  %s\n", format(spare_words, big.mark = " ")))
cat(sprintf("Tune-out time available  %.0f seconds per minute\n", tune_out_seconds_per_min))
```

</details>

<details>
<summary>Listening de-railers --- which one is Orion Tech's problem?</summary>

- **Listening without focus** --- mind multitasking.
- **Listening without respect (impatience)** --- deciding the other person is wrong and interrupting.
- **Listening without respect (ego)** --- hearing only from your own perspective, not theirs.
- **Listening with bias (defensive / adverse)** --- building your rebuttal while they talk, or listening only for points to attack.

On Orion Tech it was mostly the first and last: dominant voices building
responses instead of listening, quieter voices tuned out, and cross-cultural
tone read through the listener's own filter rather than the speaker's intent.

</details>

## Choosing a communication method

| | **Written** | **Verbal** |
|---|---|---|
| **Formal** | Complex problems, project plans, charters, long-distance --- *formal written* | Presentations and speeches --- *formal verbal* |
| **Informal** | Memos, emails (careful: email is also a contract), texts, notes --- *informal written* | Meetings and conversations --- *informal verbal* |

Should a PM try to control **all** communication between team members and
outsiders? **Yes** --- and can they? **No.** The plan defines the important
routes; it cannot police every hallway conversation.

### Try it yourself: pick the method

```{r tryit-methods-02, echo=FALSE, results='asis'}
method_bins <- c("Formal written", "Formal verbal", "Informal written", "Informal verbal")
tryit_sort(
  "comm-method", "Which communication method fits each message?",
  bins = method_bins,
  cards = data.frame(
    text = c("A signed change order to a supplier",
             "The project charter, sent for approval",
             "Presenting the risk register to the steering committee",
             "A closing speech to the client's executives",
             "A text: the truck arrives at 3 p.m.",
             "A short email reminding a colleague a drawing is due Friday",
             "A hallway chat with a quality technician about a gauge",
             "The daily 10-minute stand-up with the assembly team"),
    bin = c("Formal written", "Formal written", "Formal verbal", "Formal verbal",
            "Informal written", "Informal written", "Informal verbal", "Informal verbal"),
    why = c("Contractual weight and a record: formal written.",
            "Plans and charters are formal written documents.",
            "A presentation to a committee is formal verbal.",
            "Speeches and presentations are formal verbal.",
            "Texts and notes are informal written.",
            "Short emails are informal written (but email can still be a contract).",
            "Conversations are informal verbal.",
            "Meetings and conversations are informal verbal."),
    stringsAsFactors = FALSE),
  intro = "Pick formal or informal, and written or verbal, for every message.", question = NULL)

tryit_sort(
  "comm-method-twin", "Communication method (graded)", bins = method_bins, show = FALSE,
  cards = data.frame(
    text = c("A formal notice that a delivered lot failed acceptance testing",
             "A weekly progress presentation to the sponsor",
             "A quick chat-app message that the jig is free",
             "A one-to-one conversation to settle a scheduling clash",
             "A statement of work for a new subcontractor"),
    bin = c("Formal written", "Formal verbal", "Informal written", "Informal verbal", "Formal written"),
    stringsAsFactors = FALSE),
  question = "<p>For each message, choose the communication method that fits it best.</p>")
```

## Performance reporting

Different stakeholder questions need different reports:

| Report | Answers |
|---|---|
| **Status** | Where does the project stand right now? |
| **Progress** | What has been accomplished? |
| **Trend** | Is performance improving or deteriorating over time? |
| **Forecast** | What will status and performance be in the future? |
| **Variance** | How do actual results compare to the plan? |
| **Earned value** | Integrating scope, cost, and schedule into one performance picture (see [Chapter 12](12-Project-Control-Systems.qmd)). |

Orion Tech's stakeholders were right to expect regular **status, progress, and
forecast** reports --- they fund and approve the project, and reports give them
transparency, accountability, and early warning.

### Try it yourself: which report answers the question?

```{r tryit-reports-02, echo=FALSE, results='asis'}
report_bins <- c("Status", "Progress", "Trend", "Forecast", "Variance", "Earned value")
tryit_sort(
  "report-type", "Which performance report answers each stakeholder question?",
  bins = report_bins,
  cards = data.frame(
    text = c("Where does the project stand today?",
             "What has the team finished since the last report?",
             "Is the defect find-rate getting better or worse over six weeks?",
             "On the current rate, when will we launch?",
             "How far is tooling cost from what we budgeted?",
             "Given what is spent and what is done, are we ahead or behind overall?",
             "Are issues being closed faster than they are raised?",
             "Will we finish under budget at completion?"),
    bin = c("Status", "Progress", "Trend", "Forecast", "Variance", "Earned value", "Trend", "Forecast"),
    why = c("Status is the position right now.", "Progress is what has been accomplished.",
            "Trend: improving or deteriorating over time.", "Forecast: the future status.",
            "Variance compares actual results with the plan.",
            "Earned value integrates scope, cost and schedule.",
            "A direction over time: trend.", "Looking ahead to completion: forecast."),
    stringsAsFactors = FALSE),
  intro = "Pick the report type for each question.", question = NULL)

tryit_sort(
  "report-type-twin", "Performance report type (graded)", bins = report_bins, show = FALSE,
  cards = data.frame(
    text = c("What share of the work packages is complete?",
             "Has the schedule slipped further each week this month?",
             "What is the difference between actual and planned spend?",
             "What will the project cost at completion?",
             "Is the project on track right now?"),
    bin = c("Progress", "Trend", "Variance", "Forecast", "Status"),
    stringsAsFactors = FALSE),
  question = "<p>For each stakeholder question, choose the report type that answers it.</p>")
```

## Key Takeaways

- A **project** is temporary and unique; a **process** is permanent and repetitive. Only ~26 % of projects hit the full triple constraint --- which is why the discipline exists.
- The **triple constraint** is scope--time--cost, enclosing **quality**. Move one side and another must move, or quality pays.
- Five phases: **initiating, planning, executing, controlling, closing.** Planning decides the outcome; closing is the hardest to finish.
- Decisions are weighed **safety → company → sponsor → customer → suppliers**; ethics rest on **responsibility, respect, fairness, honesty**.
- ~**90 %** of a PM's time is communication. Build a written **communication plan** from stakeholder needs, and close every loop with **feedback**.
- Channels grow as $n(n-1)/2$ --- faster than headcount. Plan routes; don't let everyone talk to everyone.
- **Listening** is ~55 % of communication and the least trained; ~93 % of meaning is nonverbal. Watch for the **selective attention/interpretation/retention** filters and the de-railers.
- Match the **method** (formal/informal × written/verbal) to the message, and the **report** (status/progress/trend/forecast/variance/EVM) to the stakeholder's question.

## Case Study: Meridian R/C — Week 2

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 2: The Email That Didn't Reach Vancouver"}

Three weeks after Daniel Cho green-lit the Meridian Trailblazer prototype, Sofia
Reyes flagged a durability concern in the early drop-test data: the die-cast
aluminium body and frame were flexing more than expected at the wheel-well
corners. Priya Nandakumar's fix was straightforward --- thicken the aluminium
wall from **1.2 mm to 1.5 mm**, a change that would not touch the tooling
budget but would add a fraction of a gram per unit. You approved it the same
afternoon and asked Marcus Webb to notify everyone downstream who depended on
the old spec sheet.

That is where it went wrong. Marcus fired off the update to the internal
engineering distribution list --- Priya, Sofia, the CAD vendor, and you ---
because that was the list he had bookmarked from the Week 1 kickoff. It was
the right list for people *building* the part. It was not the right list for
people *selling* it. Kestrel Retail Group's Category Buyer, **Renata Ford**,
based out of **Vancouver**, had been working from the original 1.2 mm spec
sheet to brief her own packaging and merchandising team on box dimensions and
weight claims. Nobody had ever added her, or Kestrel's Vancouver office
generally, to any distribution list --- because no such list existed. Renata
found out about the change only when her packaging vendor's proof came back
listing a spec that did not match Meridian's shipped sample.

The fallout: a **3-day delay** while Kestrel's packaging vendor reprinted spec
sheets and box art, and a **$4,000** rush-reprint charge nobody had budgeted
for. Daniel Cho was patient about the money but blunt about the cause: *"This
wasn't a design problem. It was a mailing-list problem."*

You rebuilt the communication plan that week. A **weekly status email** now
goes to a named, role-based distribution list --- not a bookmarked group of
whoever was in the room at kickoff. A **biweekly steering call** with Daniel
and Kestrel keeps the client in sync in real time instead of by forwarded
email. And a **RACI matrix** now spells out, for every deliverable, who is
Responsible, Accountable, Consulted, and Informed --- so "who else needs to
know this" stops being a guess. The timing mattered too: the team was about
to grow from 4 people to 7, taking the number of two-way communication
channels from $n(n-1)/2 = 6$ to **21** --- more than triple the coordination
load the new plan now had to carry.

| Metric | Before the fix | After the fix |
|---|---|---|
| Distribution list | Ad hoc, engineering-only | Named, role-based (RACI-driven) |
| Client sync | None scheduled | Weekly status email + biweekly steering call |
| Team size | 4 | 7 |
| Two-way channels ($n(n-1)/2$) | 6 | 21 |

:::

## Practice Problems

<details>
<summary>Problem 1: Channel count and a reorganization</summary>

The RC-vehicle project currently routes all communication through a 6-person
core team. A proposal would merge in 3 supplier engineers as full team members.

(a) How many channels before and after?
(b) By what percentage do channels increase versus the 50 % increase in headcount?

<details>
<summary>Solution</summary>

```{r p1-channels-02, echo=TRUE}
before <- 6; after <- 9
ch <- function(n) n * (n - 1) / 2
cat("Before:", ch(before), "channels\n")
cat("After :", ch(after), "channels\n")
cat(sprintf("Headcount up %.0f%%, channels up %.0f%%\n",
            100 * (after / before - 1),
            100 * (ch(after) / ch(before) - 1)))
```

Headcount rises 50 %, channels rise 125 %. Merging them in as full members
triples the coordination load of those three people; a defined liaison route
would add far less.

</details>

</details>

<details>
<summary>Problem 2: Pick the method (automotive, food, defense)</summary>

Choose formal/informal and written/verbal for each:

1. Notifying a defense customer that a delivered lot failed an acceptance test.
2. A daily 10-minute stand-up with the assembly team.
3. The project charter for the RC-vehicle launch.
4. Reminding a teammate that a drawing is due Friday.

<details>
<summary>Answers</summary>

1 --- **formal written** (contractual weight, complex, long-distance).
2 --- **informal verbal** (meeting). 3 --- **formal written** (plan/charter).
4 --- **informal written** (a text or short email; keep a light record).

</details>

</details>

<details>
<summary>Problem 3 (discussion): rebuild Orion Tech's communication system</summary>

If you were the PM, redesign the tools, reporting, and meetings to prevent the
failure.

<details>
<summary>Model answer</summary>

Write a **communication plan before launch**. Use a mix:
**email** for records, **Teams/Slack** for quick updates, **dashboards** for
visual tracking, **video calls** for collaboration (so tone and context survive
the offshore gap). Run **regular status meetings with an agenda and action
items**, and have offshore developers present their own weekly updates so no
group dominates. Send stakeholders a **weekly progress report** (status, cost,
risk). Build **feedback loops**: every developer confirms, in their own words,
what a requirement means before coding it. Use plain language, avoid idioms, and
summarize decisions in writing.

</details>

</details>

## Further Resources

```{r video-resources-02, echo=FALSE, results='asis'}
embed_youtube("cSohjlYQI2A", "5 Ways to Listen Better (Julian Treasure, TED)")
embed_youtube("R1vskiVDwl4", "10 Ways to Have a Better Conversation (Celeste Headlee, TED)")
```
