---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Project Scope, Charter, and Organizational Structures

```{r setup-03, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Evaluate** a candidate project with payback period and net present value (NPV).
2. **Write** a project charter and explain what authority its sign-off grants.
3. **Identify** the five inputs to a charter and the benefits of using one.
4. **Build** a stakeholder register and place stakeholders on a power--interest grid.
5. **Develop** a scope statement --- deliverables, acceptance criteria, exclusions, constraints, assumptions --- and distinguish product scope from project scope.
6. **Recognize** scope creep and run a change-control process.
7. **Compare** functional, projectized, and matrix (weak/balanced/strong) structures and the trade-offs each imposes on the project manager.

<details>
<summary>Before we begin: who told you it was okay to start?</summary>

On your last group project, was there a moment where everyone agreed *this is
what we are building and this is who signed off*? Or did work just... start? The
charter is that moment, written down.

</details>

## Introduction: nothing official without a charter

Your company must get **Senior Management to approve the Scope and Charter
before work begins** on the RC-vehicle replica. That is not bureaucracy --- it is
the difference between a project and a hobby. *"There is no official project
without an approved Project Charter."*

Two questions come first: **should we do this project at all** (selection), and
**exactly what are we committing to** (scope). Then the organization's shape ---
functional, projectized, or matrix --- decides how much authority you actually
have to deliver it.

::: {.callout-note}
**Industry snapshot.** In automotive new-product introduction, a project cannot
pass its first gate without a signed charter and a business case showing
positive NPV. A food company scoping a new line writes a **statement of work**
that names the throughput, the allergen controls, and what is explicitly *out*.
A defense contractor's **SOW** is a contract exhibit --- every deliverable and
exclusion in it is legally enforceable.
:::

```{r video-scope-03, echo=FALSE, results='asis'}
embed_youtube("DZr3ZMWGGwc", "What Is Scope in Project Management?")
```

## Project selection: is it worth doing?

### Payback period

The simplest screen: how long until the project's net cash inflows repay the
initial outlay?

$$\text{Payback} = \text{years until cumulative cash flow} \ge 0$$

### Worked Example 1: payback on the RC-vehicle tooling investment

The project needs **\$450 000** up front (tooling, setup, approvals). Once the
100 000-unit contract runs, it returns an estimated **\$180 000** of net cash
per year.

Live cell --- change the outlay, the annual cash flow, or the horizon.

```{r calc-payback-example-03, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "payback-example-03",
  title = "Payback Period Calculator",
  icon = "clock-rotate-left",
  inputs = list(
    qwebr_calc_input("initial", "Initial outlay", value = 450000, min = 10000, max = 2000000, step = 10000, prefix = "$"),
    qwebr_calc_input("annual_cf", "Annual cash flow", value = 180000, min = 10000, max = 1000000, step = 10000, prefix = "$"),
    qwebr_calc_input("horizon", "Horizon (years)", value = 5, min = 1, max = 10, step = 1, type = "slider")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: payback-example-03
initial   <- 450000
annual_cf <- 180000
horizon   <- 5

money <- function(x) paste0(ifelse(x < 0, "-", ""), "$", format(round(abs(x)), big.mark = " ", trim = TRUE))

cum <- cumsum(rep(annual_cf, horizon)) - initial
year_paid <- which(cum >= 0)[1]

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
barplot(cum / 1000, names.arg = 1:horizon, col = ifelse(cum >= 0, "#16a085", "#e74c3c"),
        xlab = "Year", ylab = "Cumulative cash flow ($ thousand)", main = "Cumulative cash flow")
abline(h = 0, lwd = 2)

if (is.na(year_paid)) {
  cat(sprintf("!!Not paid back within %d years!!\n", horizon))
} else {
  # Fractional year within the payback year:
  prev <- if (year_paid == 1) -initial else cum[year_paid - 1]
  frac <- -prev / annual_cf
  cat(sprintf("Payback period: %.2f years\n", (year_paid - 1) + frac))
}
for (t in 1:horizon) cat(sprintf("End of year %d  %s\n", t, money(cum[t])))
```

</details>

Payback is quick to compute but ignores the **time value of money** and anything
that happens after the payback point. NPV fixes both.

The same story as a picture: the balance starts at \$0, is spent down to
**-\$450 000** as tooling, setup, and approvals are paid for ahead of **Start
of Production (SOP)**, then turns upward once production starts, as
\$180 000/year of net cash flow closes the gap until it crosses breakeven.

```{r fig-payback-curve-03, fig.cap="Cumulative cash position for the RC-vehicle tooling investment. The balance starts at $0 and is spent down to -$450,000 by Start of Production (SOP); once production begins, $180,000/year of net cash flow closes the gap, crossing breakeven at 2.5 years.", fig.align='center', fig.width=9, fig.height=5.5}
initial   <- 450000
annual_cf <- 180000
horizon   <- 5
payback_t <- initial / annual_cf

pre_sop  <- data.frame(t = seq(-1, 0, by = 0.05))
pre_sop$balance <- -initial * (pre_sop$t + 1)
post_sop <- data.frame(t = seq(0, horizon, by = 0.05))
post_sop$balance <- -initial + annual_cf * post_sop$t
curve_df <- rbind(pre_sop, post_sop)

ggplot(curve_df, aes(t, balance)) +
  geom_ribbon(aes(ymin = pmin(balance, 0), ymax = 0),
              fill = "#C0392B", alpha = 0.15) +
  geom_ribbon(data = subset(curve_df, balance > 0),
              aes(ymin = 0, ymax = balance),
              fill = "#27AE60", alpha = 0.15) +
  geom_hline(yintercept = 0, color = "grey40", linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.7) +
  geom_line(color = "#1B4F72", linewidth = 1.2) +
  geom_point(data = data.frame(t = payback_t, balance = 0),
             color = "#186A3B", size = 3.5) +
  annotate("text", x = 0.08, y = -initial * 0.82, label = "-$450,000\n(tooling outlay, at SOP)",
           hjust = 0, vjust = 1, size = 3.2 * diagram_text_scale, color = "#C0392B") +
  annotate("text", x = 0.08, y = -initial * 0.3, label = "SOP",
           hjust = 0, size = 3.2 * diagram_text_scale, color = "grey30", fontface = "bold") +
  annotate("text", x = payback_t, y = 0,
           label = paste0("Payback: ", payback_t, " yrs"),
           vjust = -1.2, hjust = -0.08, size = 3.2 * diagram_text_scale,
           color = "#186A3B", fontface = "bold") +
  scale_x_continuous(breaks = seq(-1, horizon, 1)) +
  scale_y_continuous(labels = scales::dollar_format(big.mark = " "),
                      expand = expansion(mult = c(0.14, 0.08))) +
  labs(title = "Time vs. cumulative project cash position",
       x = "Time (years)", y = "Cumulative cash position") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Net present value

NPV discounts every future cash flow back to today's dollars at the
organization's required rate of return $r$, then subtracts the initial outlay.
A positive NPV means the project earns more than that hurdle rate.

$$NPV = -C_0 + \sum_{t=1}^{n} \frac{CF_t}{(1 + r)^t}$$

### Nominal vs. discounted payback for Example 1

Apply a 12 % hurdle rate to Example 1's own numbers and put the nominal and
discounted cash flows side by side:

```{r tbl-payback-npv-03}
initial   <- 450000
annual_cf <- 180000
horizon   <- 5
r         <- 0.12

years       <- 0:horizon
cash_flow   <- c(-initial, rep(annual_cf, horizon))
cum_cf      <- cumsum(cash_flow)
disc_factor <- 1 / (1 + r)^years
pv          <- cash_flow * disc_factor
cum_pv      <- cumsum(pv)

tbl <- data.frame(
  Year = years,
  `Net cash flow` = scales::dollar(cash_flow, big.mark = " "),
  `Cumulative cash flow` = scales::dollar(cum_cf, big.mark = " "),
  `Discount factor @ 12%` = sprintf("%.3f", disc_factor),
  `Present value` = scales::dollar(round(pv), big.mark = " "),
  `Cumulative PV` = scales::dollar(round(cum_pv), big.mark = " "),
  check.names = FALSE
)

knitr::kable(tbl, align = c("c", "r", "r", "c", "r", "r"),
             caption = "Example 1: nominal cash flow and cumulative cash flow vs. the discounted present value and cumulative PV (NPV running total) at a 12% hurdle rate.") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(3, bold = TRUE) |>
  kableExtra::column_spec(6, bold = TRUE)
```

The **nominal** cumulative cash flow crosses zero between year 2 and year 3 ---
the 2.5-year payback already computed. Once each year's cash flow is
discounted at 12 %, the **cumulative PV** does not cross zero until between
year 3 and year 4: a **discounted payback of about 3.15 years**. Those extra
0.65 years are the cost of the time value of money --- and the year-5
cumulative PV, **\$198 860**, is exactly the project's NPV.

```{r fig-npv-breakeven-03, fig.cap="Discounted breakeven for Example 1. The dashed grey line is the nominal cumulative cash flow from the earlier figure; the solid navy line discounts each year's cash flow at the 12% hurdle rate before accumulating it, pushing the crossing point from 2.5 years out to about 3.15 years.", fig.align='center', fig.width=9, fig.height=5.5}
initial   <- 450000
annual_cf <- 180000
horizon   <- 5
r         <- 0.12

years       <- 0:horizon
cash_flow   <- c(-initial, rep(annual_cf, horizon))
disc_factor <- 1 / (1 + r)^years
pv          <- cash_flow * disc_factor
cum_pv      <- cumsum(pv)

# Discounted payback: fractional year where cumulative PV crosses zero
cross_i      <- which(cum_pv >= 0)[1]
frac_pv      <- -cum_pv[cross_i - 1] / pv[cross_i]
payback_pv_t <- years[cross_i - 1] + frac_pv

# Fine-resolution curves so the shaded ribbon tracks the line exactly
pre_sop        <- data.frame(t = seq(-1, 0, by = 0.02))
pre_sop$balance <- -initial * (pre_sop$t + 1)

post_t          <- seq(0, horizon, by = 0.02)
post_sop        <- data.frame(t = post_t,
                               balance = approx(years, cum_pv, xout = post_t)$y)
curve_df   <- rbind(pre_sop, post_sop)

# Nominal curve (from the earlier figure) for comparison
nominal_df <- data.frame(t = post_t, balance = -initial + annual_cf * post_t)
nominal_df <- rbind(data.frame(t = pre_sop$t, balance = pre_sop$balance), nominal_df)

ggplot(curve_df, aes(t, balance)) +
  geom_ribbon(aes(ymin = pmin(balance, 0), ymax = 0),
              fill = "#C0392B", alpha = 0.15) +
  geom_ribbon(data = subset(curve_df, balance > 0),
              aes(ymin = 0, ymax = balance),
              fill = "#27AE60", alpha = 0.15) +
  geom_hline(yintercept = 0, color = "grey40", linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.7) +
  geom_line(data = nominal_df, aes(t, balance), color = "grey55",
            linetype = "dashed", linewidth = 1) +
  geom_line(color = "#1B4F72", linewidth = 1.2) +
  geom_point(data = data.frame(t = payback_pv_t, balance = 0),
             color = "#186A3B", size = 3.5) +
  annotate("text", x = 0.08, y = -initial * 0.82, label = "-$450,000\n(tooling outlay, at SOP)",
           hjust = 0, vjust = 1, size = 3.2 * diagram_text_scale, color = "#C0392B") +
  annotate("text", x = 0.08, y = -initial * 0.3, label = "SOP",
           hjust = 0, size = 3.2 * diagram_text_scale, color = "grey30", fontface = "bold") +
  annotate("text", x = payback_pv_t, y = 0,
           label = paste0("Discounted payback: ", round(payback_pv_t, 2), " yrs"),
           vjust = -1.2, hjust = 1.05, size = 3.2 * diagram_text_scale,
           color = "#186A3B", fontface = "bold") +
  annotate("text", x = horizon - 0.1, y = -initial + annual_cf * (horizon - 0.1),
           label = "nominal (undiscounted)", hjust = 1, vjust = -0.8,
           size = 3 * diagram_text_scale, color = "grey45", fontface = "italic") +
  scale_x_continuous(breaks = seq(-1, horizon, 1)) +
  scale_y_continuous(labels = scales::dollar_format(big.mark = " "),
                      expand = expansion(mult = c(0.14, 0.08))) +
  labs(title = "Nominal vs. discounted (NPV) cumulative cash position",
       x = "Time (years)", y = "Cumulative cash position") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Try it yourself: NPV of the project

```{r calc-npv-calc-03, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "npv-calc-03",
  title = "NPV Calculator",
  icon = "coins",
  inputs = list(
    qwebr_calc_input("initial", "Initial outlay", value = 450000, min = 10000, max = 2000000, step = 10000, prefix = "$"),
    qwebr_calc_input("r", "Required rate of return", value = 0.12, min = 0, max = 0.3, step = 0.01, type = "slider")
  ),
  goal = "Make the project fail its hurdle: push the NPV below zero. Which input has to move, and how far?",
  goal_test = "npv < 0"
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: npv-calc-03
initial   <- 450000
cash_flows<- rep(180000, 5)   # years 1..5, as in Worked Example 1
r         <- 0.12   # required rate of return

t   <- seq_along(cash_flows)
pv  <- cash_flows / (1 + r)^t
npv <- sum(pv) - initial

money <- function(x) paste0(ifelse(x < 0, "-", ""), "$", format(round(abs(x)), big.mark = " ", trim = TRUE))

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
bp <- barplot(rbind(cash_flows, pv) / 1000, beside = TRUE, names.arg = t, col = c("#bdc3c7", "#2980b9"), xlab = "Year", ylab = "$ thousand",
              main = "Cash flow (grey) vs. present value (blue)")

if (npv > 0) {
  cat(sprintf("NPV %s: ACCEPT (beats the %s%% hurdle)\n", money(npv), format(100 * r)))
} else {
  cat(sprintf("NPV %s: !!REJECT!! (misses the %s%% hurdle)\n", money(npv), format(100 * r)))
}
for (i in t) cat(sprintf("Year %d present value  %s\n", i, money(pv[i])))
cat(sprintf("Sum of PVs  %s\n", money(sum(pv))))
cat(sprintf("Less outlay  %s\n", money(initial)))
```

</details>

<details>
<summary>What happens to NPV if the hurdle rate rises to 18 %?</summary>

Raise `r` in the cell and re-run. Higher discount rates punish cash flows that
arrive later, so NPV falls --- a project that clears a 12 % hurdle may fail an
18 % one. This is why a company with expensive capital (or a risky project that
demands a risk premium) is choosier about which projects it starts. Course
Learning Outcome 1 is exactly this: *communicate the philosophy of project
selection using the NPV method.*

</details>

### Try it yourself: a discounting table by hand

Build the year-by-year table for a fresh project, then read off its NPV and
discounted payback. Amounts are in \$ thousand. Each team picks its own **data
set**.

```{r tryit-npv-03, echo=FALSE, results='asis'}
npv_sets <- list(
  list(outlay = 320, r = 0.10, cf = c(110, 120, 130, 90, 70)),
  list(outlay = 250, r = 0.08, cf = c(70, 80, 90, 100, 60)),
  list(outlay = 500, r = 0.12, cf = c(150, 160, 170, 120, 100)),
  list(outlay = 180, r = 0.15, cf = c(60, 70, 80, 50, 40)))
npv_table <- function(s) {
  t <- 0:5
  cash <- c(-s$outlay, s$cf)
  df <- 1 / (1 + s$r)^t
  pv <- cash * df
  data.frame(Year = t, `Rate` = sprintf("%g %%", 100 * s$r), `Cash flow ($k)` = cash, `Discount factor` = df,
             `Present value ($k)` = pv, `Cumulative PV ($k)` = cumsum(pv), check.names = FALSE)
}
npv_answers <- function(s) {
  d <- npv_table(s)
  cum <- d$`Cumulative PV ($k)`; pv <- d$`Present value ($k)`
  k <- which(cum >= 0)[1]
  list(tryit_answer("NPV at the end of year 5", tail(cum, 1), tol = 0.3, suffix = "$k"),
       tryit_answer("Discounted payback", (k - 2) + (-cum[k - 1] / pv[k]), tol = 0.05, suffix = "years"))
}
npv_cols <- c("Discount factor", "Present value ($k)", "Cumulative PV ($k)")
tryit_grid(
  "npv-by-hand", "Discounted cash flow table",
  data = npv_table(npv_sets[[1]]), blank = npv_cols,
  tol = c(`Discount factor` = 0.001, `Present value ($k)` = 0.1, `Cumulative PV ($k)` = 0.3),
  digits = 1, answers = npv_answers(npv_sets[[1]]),
  teams = lapply(npv_sets[-1], function(s) list(data = npv_table(s), answers = npv_answers(s))),
  hints = c(`Discount factor` = "1 / (1 + r)^t. Year 0 is not discounted, so its factor is 1.",
            `Present value ($k)` = "Cash flow x discount factor.",
            `Cumulative PV ($k)` = "Running total of the present values, starting with year 0."),
  intro = "The required rate of return is in the Rate column. Year 0 is the outlay.")

# Graded twin: a different project.
npv_twin <- list(outlay = 400, r = 0.09, cf = c(130, 140, 150, 110, 90))
tryit_grid(
  "npv-twin", "Discounted cash flow table (graded)", show = FALSE,
  data = npv_table(npv_twin), blank = npv_cols,
  tol = c(`Discount factor` = 0.001, `Present value ($k)` = 0.1, `Cumulative PV ($k)` = 0.3),
  digits = 1, answers = npv_answers(npv_twin),
  question = paste0("<p>A project costs $400 000 at year 0 and returns $130k, $140k, $150k, $110k ",
    "and $90k in years 1 to 5. The required rate of return is 9 %. Complete the table (amounts in ",
    "$ thousand), then give the NPV at the end of year 5 and the discounted payback in years ",
    "(interpolate within the year the cumulative present value turns positive).</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

For data set 1 (outlay \$320k, r = 10 %): the present values are 100.0, 99.2,
97.7, 61.5 and 43.5, so the cumulative PV runs -220.0, -120.8, -23.2, +38.3 and
+81.8. The NPV is **\$81.8k**. The cumulative PV turns positive in year 4:
discounted payback = 3 + 23.2 / 61.5 = **3.38 years**. The nominal payback
would be shorter; the gap is the cost of the time value of money.

</details>

## The project charter

> **PMI:** developing a document that formally **authorizes** a project and
> documents the initial requirements that satisfy the stakeholders' needs and
> expectations. **Approval of the charter formally initiates the project.**

### Five inputs to a charter

1. **Project Statement of Work** --- the high-level "project brief": business need and product scope, fewer details than the scope statement.
2. **Business Case** --- why the project is being done (the NPV/payback story).
3. **Contract** (if the client is external).
4. **Enterprise Environmental Factors (EEFs)** --- conditions you cannot control (market, regulation, weather).
5. **Organizational Process Assets (OPAs)** --- the company's plans, processes, policies, and accumulated knowledge.

### What a charter contains

Title and description; business case; pre-assigned resources; stakeholders and
their known requirements; product deliverables and the end result; measurable
objectives tied to strategy; a **summary** milestone schedule; a **summary**
budget (gross only --- the real budget comes after planning); approval
requirements and who signs off; high-level risks; and **signatures**.

### Why bother

- Stakeholders are clearly defined
- Communication channels are defined
- Roles and responsibilities are outlined
- Scope is defined --- helping prevent **scope creep**
- Better project processes in later phases
- Higher probability of success
- **Buy-in** from the team

The charter is a **living document**: update it as the project moves. Used well,
it is *"the single most important tool the project manager uses to manage the
expectations of the sponsor and all other stakeholders."*

## Stakeholders

A **stakeholder** is any individual or group with an active interest in the
project who can affect it.

| Internal | External |
|---|---|
| Senior management, accountant, team members, IT department | Client, competitors, suppliers, environmental agencies, government agencies, research labs |

Managing them is a loop: **identify → gather information → determine strengths
and weaknesses → predict behaviour → build an engagement strategy → implement**.
The point of engagement is to **increase support and reduce resistance** --- by
listening, informing, and helping stakeholders along.

```{r fig-stakeholder-grid-03, fig.cap="Power-interest grid for the RC-vehicle project. Effort goes where power and interest are both high; low-low stakeholders are monitored, not managed.", fig.align='center', fig.width=9, fig.height=6.5}
sh <- data.frame(
  who = c("Senior mgmt / sponsor", "Retailer (client)", "Motor supplier",
          "Carbon-fibre supplier", "Compliance officer", "Assembly team",
          "Accountant", "Competitors", "Local community"),
  power    = c(0.92, 0.80, 0.55, 0.45, 0.70, 0.40, 0.50, 0.30, 0.20),
  interest = c(0.75, 0.85, 0.60, 0.55, 0.65, 0.90, 0.55, 0.65, 0.35)
)

ggplot(sh, aes(interest, power)) +
  annotate("rect", xmin = 0.5, xmax = 1, ymin = 0.5, ymax = 1.12, fill = "#27AE60", alpha = 0.10) +
  annotate("rect", xmin = 0, xmax = 0.5, ymin = 0.5, ymax = 1.12, fill = "#F39C12", alpha = 0.10) +
  annotate("rect", xmin = 0.5, xmax = 1, ymin = 0, ymax = 0.5, fill = "#5DADE2", alpha = 0.10) +
  annotate("rect", xmin = 0, xmax = 0.5, ymin = 0, ymax = 0.5, fill = "grey70", alpha = 0.12) +
  geom_point(color = "#1B4F72", size = 3) +
  geom_text(aes(label = who), size = 2.9 * diagram_text_scale,
            vjust = ifelse(sh$power > 0.85, 1.8, -0.8)) +
  annotate("text", x = 0.75, y = 1.08, label = "MANAGE CLOSELY",
           fontface = "bold", size = 3 * diagram_text_scale, color = "#186A3B") +
  annotate("text", x = 0.25, y = 1.08, label = "KEEP SATISFIED",
           fontface = "bold", size = 3 * diagram_text_scale, color = "#9C640C") +
  annotate("text", x = 0.75, y = 0.03, label = "KEEP INFORMED",
           fontface = "bold", size = 3 * diagram_text_scale, color = "#1B4F72") +
  annotate("text", x = 0.25, y = 0.03, label = "MONITOR",
           fontface = "bold", size = 3 * diagram_text_scale, color = "grey30") +
  scale_x_continuous(limits = c(0, 1), breaks = c(0.1, 0.9), labels = c("low", "high")) +
  scale_y_continuous(limits = c(0, 1.12), breaks = c(0.1, 0.9), labels = c("low", "high")) +
  labs(title = "Stakeholder power vs. interest", x = "Interest", y = "Power") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Try it yourself: score your own stakeholders onto the grid

Rate each stakeholder's **power** and **interest** on a 1--10 scale and the
quadrant follows automatically --- the same rule the figure above was built
from. Live cell --- edit the table or add rows for your own project.

```{webr-r}
#| label: stakeholder-score-03
stakeholders <- data.frame(
  who      = c("Kestrel Retail Group", "Daniel Cho (sponsor)", "MCS (motor supplier)",
               "Internal QA", "Fibro Source (trim supplier)"),
  power    = c(9, 9, 5, 5, 2),
  interest = c(8, 8, 7, 5, 2)
)

stakeholders$quadrant <- with(stakeholders, ifelse(
  power >= 5 & interest >= 5, "Manage closely",
  ifelse(power >= 5 & interest < 5, "Keep satisfied",
  ifelse(power < 5 & interest >= 5, "Keep informed", "Monitor"))))

print(stakeholders, row.names = FALSE)
cat("\nChange the power/interest scores (1-10) for your own project's stakeholders\n")
cat("and re-run -- the quadrant assignment updates automatically.\n")
```

### Try it yourself: place each stakeholder on the grid

```{r tryit-stakeholders-03, echo=FALSE, results='asis'}
sh_bins <- c("Manage closely", "Keep satisfied", "Keep informed", "Monitor")
tryit_sort(
  "stakeholder-quadrants", "Which quadrant is each stakeholder in?",
  bins = sh_bins,
  cards = data.frame(
    text = c("City building inspector (power 8, interest 3)",
             "Assembly team lead (power 4, interest 9)",
             "Retail buyer (power 9, interest 9)",
             "Neighbouring business (power 2, interest 3)",
             "Environmental agency (power 7, interest 2)",
             "Motor supplier (power 6, interest 8)",
             "Local newspaper (power 3, interest 6)",
             "Finance director (power 9, interest 4)"),
    bin = c("Keep satisfied", "Keep informed", "Manage closely", "Monitor",
            "Keep satisfied", "Manage closely", "Keep informed", "Keep satisfied"),
    why = c("High power, low interest: keep satisfied.", "Low power, high interest: keep informed.",
            "High on both: manage closely.", "Low on both: monitor.",
            "High power, low interest: keep satisfied.", "High on both: manage closely.",
            "Low power, high interest: keep informed.", "High power, low interest: keep satisfied."),
    stringsAsFactors = FALSE),
  intro = "Scores are out of 10; 5 and above counts as high. Pick the quadrant for each stakeholder.",
  question = NULL)

tryit_sort(
  "stakeholder-twin", "Power-interest grid (graded)", bins = sh_bins, show = FALSE,
  cards = data.frame(
    text = c("Union steward (power 3, interest 8)", "Project sponsor (power 9, interest 8)",
             "Municipal planner (power 7, interest 3)", "Industry association (power 2, interest 2)",
             "Key customer (power 8, interest 9)"),
    bin = c("Keep informed", "Manage closely", "Keep satisfied", "Monitor", "Manage closely"),
    stringsAsFactors = FALSE),
  question = "<p>Using the power-interest grid (a score of 5 or more is high), choose the strategy for each stakeholder.</p>")
```

## The scope statement

Written early in **planning**, the scope statement is more detailed than the
charter and should include:

- The **project statement** (what will be done)
- **Acceptance criteria** and the **validation/testing plan**
- **Deliverables**
- **Exclusions** (what is explicitly *not* in scope)
- **Constraints** and **assumptions**
- Project coordination
- **Signed-off** review with the customer and with corporate management

**Product scope** = the features and functions of the thing you are building.
**Project scope** = the work needed to deliver it. A scope statement covers
both, plus boundaries and references to specs. *As the project progresses, scope
should get clearer and more specific --- and changes here should be avoided,
though not forbidden.*

### The scope management processes

**Plan scope → collect requirements → define scope → create WBS → validate scope
(at closing) → control scope (throughout).** The scope management plan is often
run as an extension of **Management by Objectives (MBO)**: set unambiguous,
realistic objectives; check periodically whether they are being met; take
corrective action.

## Scope creep

**Scope creep** is the uncontrolled growth of requirements over the project's
life:

- one deliverable quietly becomes five
- three "essential" features become ten
- mid-project, the customer's needs change and requirements are reopened

Its usual causes are **stakeholders changing requirements** and **internal
miscommunication**. It is *not always bad* --- customer needs genuinely evolve,
and refusing all change can deliver the wrong product. The problem is
**uncontrolled** change. A good PM **expects it and plans for it** with a
**change-control process**: every change request is logged, its impact on scope,
schedule, and cost is assessed, and it is approved or rejected before work
changes.

### Try it yourself: the cost of an "innocent" change

```{r calc-scope-change-03, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "scope-change-03",
  title = "Cost of a Scope Change",
  icon = "file-circle-exclamation",
  inputs = list(
    qwebr_calc_input("extra_design_hours", "Extra design hours", value = 60, min = 0, max = 400, step = 5, suffix = "h"),
    qwebr_calc_input("designer_rate", "Designer rate", value = 46, min = 10, max = 150, step = 1, prefix = "$", suffix = "/ h"),
    qwebr_calc_input("extra_tooling_cost", "Extra tooling cost", value = 28000, min = 0, max = 200000, step = 1000, prefix = "$"),
    qwebr_calc_input("indirect_rate", "Indirect rate", value = 0.35, min = 0, max = 1, step = 0.05),
    qwebr_calc_input("schedule_days_added", "Schedule days added", value = 9, min = 0, max = 60, step = 1, type = "slider"),
    qwebr_calc_input("penalty_per_day_late", "Late penalty", value = 1500, min = 0, max = 10000, step = 100, prefix = "$", suffix = "/ day")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: scope-change-03
# A stakeholder asks to add a second body colourway "while we're at it".
extra_design_hours   <- 60
extra_tooling_cost   <- 28000
designer_rate        <- 46
indirect_rate        <- 0.35
schedule_days_added  <- 9
penalty_per_day_late <- 1500     # retailer contract clause

direct  <- extra_design_hours * designer_rate + extra_tooling_cost
loaded  <- direct * (1 + indirect_rate)
sched   <- schedule_days_added * penalty_per_day_late

fmt <- function(x) paste0("$", format(round(x), big.mark = " ", scientific = FALSE, trim = TRUE))

parts <- c(Direct = direct, Indirect = loaded - direct, "Late penalty" = sched)
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
bp <- barplot(parts / 1000, col = c("#2980b9", "#8e44ad", "#e74c3c"), ylim = c(0, max(parts / 1000) * 1.2), ylab = "$ thousand",
              main = "Where the cost of the change lands")
text(bp, parts / 1000, labels = paste0("$", round(parts / 1000, 1), "k"), pos = 3, cex = 1.3)

cat(sprintf("Total impact of one 'small' change: %s\n", fmt(loaded + sched)))
cat(sprintf("Direct cost  %s\n", fmt(direct)))
cat(sprintf("Fully burdened (x%.2f)  %s\n", 1 + indirect_rate, fmt(loaded)))
cat(sprintf("Schedule penalty (%g days)  %s\n", schedule_days_added, fmt(sched)))
```

</details>

## Organizational culture

Every organization develops its own outlook, policies, habits of thought, and
norms --- *"as unique as a fingerprint."* Culture is *the solution to problems
that has worked consistently for a group and is therefore taught to new members
as the correct way to perceive and think.* It is shaped by **technology,
environment, geographic location, reward systems, rules and procedures, key
organizational members, and critical incidents.**

Culture affects projects through:

- **Department interaction** --- cooperative cultures run projects far better than adversarial ones.
- **Employee commitment to goals** --- teams need to understand the *why* to be motivated.
- **Project planning** --- some cultures tolerate padding estimates but not being late.
- **Performance evaluation** --- a culture that rewards innovation breeds aggressive, high-payoff project management (with occasional big losses).

## Organizational structures

How the lines of authority and communication are arranged decides **when people
are available** and **how you get them**.

```{r fig-org-authority-03, fig.cap="Project manager authority across organizational structures. It rises from almost none in a functional organization to near-total in a projectized one, with the matrix band in between.", fig.align='center', fig.width=10, fig.height=4.5}
org <- data.frame(
  type = factor(c("Functional", "Weak matrix", "Balanced matrix",
                  "Strong matrix", "Projectized"),
                levels = c("Functional", "Weak matrix", "Balanced matrix",
                           "Strong matrix", "Projectized")),
  authority = c(0.08, 0.28, 0.5, 0.72, 0.95)
)
ggplot(org, aes(type, authority, fill = authority)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = scales::percent(authority, accuracy = 1)),
            vjust = -0.4, size = 3.6 * diagram_text_scale) +
  scale_fill_gradient(low = "#AED6F1", high = "#1B4F72", guide = "none") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.05),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(title = "How much authority does the PM actually have?",
       x = NULL, y = "PM authority over resources") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

| Structure | Strengths for the PM | Weaknesses for the PM |
|---|---|---|
| **Functional** (grouped by specialty; one clear boss; each department runs its own projects) | No restructuring needed; deep expertise; everyone knows their place | Many layers to reach people; weak customer focus; projects drag because staff have "day jobs" and other priorities |
| **Projectized** (the company just does projects; team co-located) | One reporting line; easy communication; fast decisions; strong PMs; rapid market response | Teams are expensive to set up and maintain; loyalty to the project over the organization; hard to keep a shared pool of expertise; team anxiety about "what next" when the project ends |
| **Matrix** (blend; classified **weak / balanced / strong** by PM influence) | Suited to dynamic environments; values both project and functional efficiency; coordinates across units; maximizes scarce shared resources | **Two bosses**; heavy time spent negotiating shared resources; frustrating for staff caught between project and functional demands |

<details>
<summary>Which structure is the RC-vehicle project running in?</summary>

Almost certainly a **matrix** --- the company has an existing manufacturing line
and functional departments (design, quality, purchasing), and it borrows people
into the project. That means the PM shares the design engineers with their
functional manager and must negotiate their time. Expect the "administrative
conflict" of [Chapter 10](10-Negotiation-and-Conflict-Management.qmd): every
matrix team member has two bosses.

</details>

### Try it yourself: which structure is it?

```{r tryit-structures-03, echo=FALSE, results='asis'}
org_bins <- c("Functional", "Projectized", "Weak matrix", "Balanced matrix", "Strong matrix")
tryit_sort(
  "org-structures", "Which organizational structure does each description fit?",
  bins = org_bins,
  cards = data.frame(
    text = c("Each department runs its own projects; any PM is a part-time coordinator inside one department.",
             "The company only does projects; the whole team is co-located and reports to the PM.",
             "Staff report to functional managers; the PM mainly coordinates and expedites.",
             "The PM and the functional managers share authority and negotiate over people.",
             "A full-time PM with a project office controls the budget and assignments; functions supply specialists.",
             "The main worry is team anxiety about 'what next' when the project ends.",
             "Projects drag because staff have day jobs and other priorities.",
             "Every team member has two bosses, and the PM and functional manager negotiate for their time as equals."),
    bin = c("Functional", "Projectized", "Weak matrix", "Balanced matrix", "Strong matrix",
            "Projectized", "Functional", "Balanced matrix"),
    why = c("Grouped by specialty, one clear boss, each department runs its own projects.",
            "One reporting line, fast decisions, strong PM.",
            "Little PM authority: the weak end of the matrix.", "Authority split evenly: balanced.",
            "More PM authority than the functional managers: strong.",
            "A weakness of projectized teams: where do we go next?",
            "A weakness of functional organizations.", "Two bosses with equal say: balanced matrix."),
    stringsAsFactors = FALSE),
  intro = "Pick the structure each description fits best.", question = NULL)

tryit_sort(
  "org-structures-twin", "Organizational structure (graded)", bins = org_bins, show = FALSE,
  cards = data.frame(
    text = c("A team seconded from several departments; the PM has formal authority over budget and schedule.",
             "Engineers sit in their department; the PM must ask each manager for hours.",
             "A dedicated design-build team is set up for each contract and disbanded afterwards.",
             "Authority over people is shared equally between the PM and the function heads.",
             "Customers must pass through several layers to reach the people doing the work."),
    bin = c("Strong matrix", "Weak matrix", "Projectized", "Balanced matrix", "Functional"),
    stringsAsFactors = FALSE),
  question = "<p>For each description, choose the organizational structure it fits best.</p>")
```

## Key Takeaways

- **Select** projects on merit: **payback** is quick but ignores the time value of money; **NPV** discounts every future cash flow and is the course's standard (CLO 1). Positive NPV = beats the hurdle rate.
- The **charter** *formally initiates* the project. Inputs: SOW, business case, contract, EEFs, OPAs. It defines stakeholders, roles, and scope, and earns team **buy-in**. No approved charter, no official project.
- Manage **stakeholders** on a **power--interest grid**: manage the high-power/high-interest closely, monitor the rest.
- The **scope statement** covers **product and project scope**, plus acceptance criteria, exclusions, constraints, and assumptions, and is signed off.
- **Scope creep** is uncontrolled requirement growth. Change is normal; *uncontrolled* change is the enemy --- run every request through **change control** for a cost/schedule impact and a decision.
- **Culture** shapes project outcomes through department cooperation, commitment, planning honesty, and what performance evaluation rewards.
- **Structure** sets the PM's real authority: near zero in **functional**, near total in **projectized**, negotiated in the **matrix** (weak/balanced/strong) --- where everyone has two bosses.

## Case Study: Meridian R/C — Week 3

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 3: The Charter Lands on Your Desk"}

This week the paperwork catches up with the promise. Daniel Cho, Meridian's VP
of Product and the project's sponsor, signs off on the project charter for the
Meridian Trailblazer, the 1/18-scale rally-truck replica that Kestrel Retail
Group has all but promised to order in bulk --- 100,000 units or more for next
year's holiday season --- if the prototype survives durability testing and
toy-safety certification. Cho's signature does more than start the clock: it
sets a hard ceiling. You are authorized to spend **up to $800,000** to get a
certified, sellable prototype in front of Kestrel. Every budget you build from
here --- the detailed estimate due in a few weeks, the crashing decisions
later, the earned-value tracking after that --- has to fit under that number.

The charter also settles a structural question you've been dancing around
since the aluminium-vs-steel-vs-carbon-fibre debate in Week 1: Meridian is not
spinning up a dedicated project team. You are running a **matrix
organization**. Priya Nandakumar keeps her desk in Manufacturing Engineering
and is loaned to you for body and frame tooling. Marcus Webb still reports to
Procurement and negotiates your supplier contracts on top of his other
accounts. Sofia Reyes stays inside Quality and splits her week between your
durability and certification testing and her department's other work. None of
the three has you as their only boss --- and neither do you have them full
time.

With the charter signed, you sit down and sketch a power-interest grid to
decide where your attention goes this quarter:

| Stakeholder | Power | Interest | Quadrant |
|---|---|---|---|
| Kestrel Retail Group | High | High | Manage closely |
| Daniel Cho (sponsor) | High | High | Manage closely |
| MCS (motor supplier) | Medium | High | Keep informed |
| Paragon (motor supplier) | Medium | High | Keep informed |
| Internal QA department | Medium | Medium | Keep satisfied |
| Fibro Source (trim supplier) | Low | Low | Monitor |

Kestrel and Cho land in the top-right corner --- high power, high interest ---
because either one can stall or kill the project. MCS and Paragon, your two
candidate motor suppliers, sit at medium power but high interest: they care
intensely about winning the contract, so you keep them informed even though
neither can unilaterally derail you. Internal QA, medium on both axes, gets
kept satisfied: not a daily conversation, but not ignored either. Fibro
Source, supplying only the optional carbon-fibre trim package, is low power
and low interest --- worth monitoring, nothing more.

:::

## Practice Problems

<details>
<summary>Problem 1: Payback vs. NPV disagree</summary>

Project A: \$300 000 outlay, then \$150 000/yr for 3 years, nothing after.
Project B: \$300 000 outlay, then \$90 000/yr for 6 years. Hurdle rate 10 %.
Which has the shorter payback, and which has the higher NPV?

<details>
<summary>Solution</summary>

```{r p1-select-03, echo=TRUE}
npv <- function(c0, cfs, r) -c0 + sum(cfs / (1 + r)^seq_along(cfs))
A <- c(150000, 150000, 150000)
B <- rep(90000, 6)
cat(sprintf("A: payback 2.0 yr, NPV $%s\n", format(round(npv(3e5, A, 0.10)), big.mark = " ")))
cat(sprintf("B: payback ~3.3 yr, NPV $%s\n", format(round(npv(3e5, B, 0.10)), big.mark = " ")))
```

A pays back faster; B has the higher NPV because its cash flows keep coming for
three more years. Payback alone would pick the worse project.

</details>

</details>

<details>
<summary>Problem 2: Charter or scope statement?</summary>

Which document does each belong in?

1. "Gross budget: approximately \$650 000."
2. "Body colourways: Racing Red and Gunmetal only; no custom colours."
3. "Sponsor: VP Operations. Signs off at Gate 1."
4. "Drop test: survives 10 drops from 1.5 m onto concrete, no functional damage."

<details>
<summary>Answers</summary>

1. **Charter** (summary budget).
2. **Scope statement** (exclusions).
3. **Charter** (approval/signatures).
4. **Scope statement** (acceptance criteria).

</details>

</details>

<details>
<summary>Problem 3 (discussion): the functional manager won't release the good engineer</summary>

In a weak matrix, your best-choice design engineer's functional manager keeps
assigning them to department work and offers you a junior instead. Using the
culture and structure ideas in this chapter, explain why this happens and what
you can do.

<details>
<summary>Discussion points</summary>

The functional manager is **evaluated on their
department's output**, so they hoard their strongest people --- a reward-system
effect. In a weak matrix you have little formal authority to override it. Levers:
escalate to the sponsor (whose NPV depends on the project), negotiate a
part-time or time-boxed assignment, trade a deliverable the functional manager
cares about, or get the resource commitment written into the **charter** so it is
a documented agreement rather than a favour.

</details>

</details>

## Further Resources

```{r video-resources-03, echo=FALSE, results='asis'}
embed_youtube("30a6zZP5EVI", "What Is Scope Creep? (Explained in 3 Minutes)")
embed_youtube("HHey0EXZR0w", "Scope Creep in Project Management: Causes, Examples, Prevention")
```
