---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Leadership and the Project Manager

```{r setup-04, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Distinguish** leadership from management in a project context.
2. **Explain** the four foundations of a project partnership and Bennis's four competencies of the "new project leader."
3. **Match** a leadership style (command & control, relating, democratic, hands-on, coaching) to a team situation, and state each style's downside.
4. **Score** a leadership self-assessment and interpret the result.
5. **Describe** the stages of team development and what the leader does in each.
6. **Apply** motivation theory --- Maslow, Herzberg, and expectancy --- to diagnose a performance problem.
7. **Identify** the five components of emotional intelligence and why they matter to a project leader.

<details>
<summary>Before we begin: recall a team you would follow anywhere</summary>

Who led it, and what did they *do* --- not what title they held? Most people
describe behaviour, not authority. That is the whole chapter in one observation.

</details>

## Introduction: the project manager as mini-CEO

Leading the RC-vehicle project means more than tracking a Gantt chart. The brief
has your team learning a **new assembly line** with a real learning curve,
working with an **unproven supplier**, and racing a **holiday deadline**. The PM
*"encapsulates the role of a mini-CEO"* --- expected to manage holistically from
start to finish: acquiring resources, motivating and building the team, holding
a vision *and* fighting fires, and communicating relentlessly. *"A PM cannot get
up in the morning and think 'I don't want to communicate today.'"*

::: {.callout-note}
**Industry snapshot.** On an automotive launch, the line lead who can only give
orders stalls the moment a novel problem appears; the one who has built a team
that raises issues early hits the job-one date. In food processing, a
sanitation crew that understands *why* the changeover procedure matters follows
it at 3 a.m. when no one is watching. Leadership is what fills the gap between
the procedure and reality.
:::

```{r video-leaders-safe-04, echo=FALSE, results='asis'}
embed_youtube("lmyZMtPVodo", "Why Good Leaders Make You Feel Safe (Simon Sinek, TED)")
```

## Leadership versus management

**Leadership** is *the ability to inspire confidence and support among the
people needed to achieve organizational goals.* A **manager** holds a title that
lets them *plan, organize, direct, and control* others. Leadership may be part
of the manager's job, but the other roles are more administrative.

Leaders create a sense of **partnership**. Four things build it:

| Foundation | What it means |
|---|---|
| **Exchange of purpose** | Every member knows the project's goals and vision. |
| **A right to say "no"** | Members can voice professional disagreement. *Losing an argument is acceptable; losing the right to disagree is not.* |
| **Joint accountability** | Every member owns the project's outcomes --- good or bad. |
| **Absolute honesty** | All information becomes community information. |

```{r fig-leader-vs-manager-04, fig.cap="Manager and leader emphases. Neither column is 'bad' --- a project manager needs both, but leans on the right-hand behaviours during execution.", fig.align='center', fig.width=10, fig.height=5}
lm <- data.frame(
  aspect = c("Focus", "Asks", "Relies on", "Time horizon",
             "Handles change by", "Success is"),
  manager = c("Systems & structure", "How & when", "Control", "Short term",
              "Administering", "Doing things right"),
  leader  = c("People", "What & why", "Trust", "Longer term",
              "Setting direction", "Doing the right things")
)
lm_long <- tidyr::pivot_longer(lm, c(manager, leader),
                               names_to = "role", values_to = "text")
lm_long$aspect <- factor(lm_long$aspect, levels = rev(lm$aspect))
lm_long$role <- factor(lm_long$role, levels = c("manager", "leader"),
                       labels = c("MANAGER", "LEADER"))

ggplot(lm_long, aes(role, aspect)) +
  geom_tile(aes(fill = role), color = "white", linewidth = 1.5) +
  geom_text(aes(label = text), size = 3.1 * diagram_text_scale, lineheight = 0.9) +
  scale_fill_manual(values = c("MANAGER" = "#AED6F1", "LEADER" = "#A9DFBF"),
                    guide = "none") +
  labs(title = "Two hats the project manager wears", x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.text.x = element_text(size = 11 * diagram_text_scale, face = "bold"),
    axis.text.y = element_text(size = 10 * diagram_text_scale)
  )
```

### Bennis's four competencies of the new project leader

1. **The power of appreciation** --- recognize and reward others' talent; your job is not to outshine the team but to let it develop.
2. **Keep reminding people what's important** --- amid the noise of problems, keep everyone's eye on the prize.
3. **Generate and sustain trust** --- through authenticity, fairness, honesty, and caring. Honesty is the trait people look for most.
4. **The leader and the led are intimate allies** --- leadership supports the team's effort; it does not dominate it.

## The five leadership styles

Effective PMs switch styles to fit the person and the situation.

| Style | Use it when | The downside if overused |
|---|---|---|
| **Command & control** | You know the best way and the other person doesn't; beginners; a crisis | Team feels devalued; learned helplessness; kills creativity and initiative |
| **Relating** | Building rapport, bonding the team, focusing people on common goals | Can't make tough calls or take charge in a crisis; seen as weak |
| **Democratic** | *Planning* --- getting buy-in and reality-based plans from those who will execute | *Execution* often needs autocracy; consensus-building wastes a critical deadline |
| **Hands-on** | You must step in and model the way to get a stuck task done | Disempowering; low morale; you become the bottleneck |
| **Coaching** | Long-term development; asking questions that lead people to their own solutions | Slow for short-term objectives --- but it is the style most managers most need to build |

```{r fig-situational-04, fig.cap="Situational leadership: match the style to how ready the team is. New or struggling teams need direction; capable, committed teams need delegation.", fig.align='center', fig.width=9, fig.height=6}
sl <- data.frame(
  x = c(0.8, 0.35, 0.65, 0.2),
  y = c(0.8, 0.8, 0.3, 0.3),
  label = c("COACHING\n(high direction,\nhigh support)",
            "SUPPORTING\n(low direction,\nhigh support)",
            "DIRECTING\n(high direction,\nlow support)",
            "DELEGATING\n(low direction,\nlow support)")
)
ggplot(sl, aes(x, y)) +
  geom_point(size = 42, shape = 21, fill = "#D6EAF8", color = "#2E86C1") +
  geom_text(aes(label = label), size = 2.6 * diagram_text_scale, lineheight = 0.9) +
  annotate("segment", x = 0, xend = 1, y = 0, yend = 0,
           arrow = arrow(length = unit(0.2, "cm")), color = "grey40") +
  annotate("segment", x = 0, xend = 0, y = 0, yend = 1,
           arrow = arrow(length = unit(0.2, "cm")), color = "grey40") +
  annotate("text", x = 0.5, y = -0.08, label = "Directive behaviour  ->",
           size = 3 * diagram_text_scale) +
  annotate("text", x = -0.06, y = 0.5, label = "Supportive behaviour  ->",
           angle = 90, size = 3 * diagram_text_scale) +
  coord_equal(xlim = c(-0.12, 1.05), ylim = c(-0.12, 1.05)) +
  labs(title = "Match the style to team readiness") +
  theme_void() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale,
                                  face = "bold", hjust = 0.5))
```

### Try it yourself: the leadership self-assessment

The course quiz has 18 statements rated 1 (strongly disagree) to 5 (strongly
agree). Items **1, 4, 7, 10, 13, 16** sum to a **command & control** score;
items **2, 5, 8, 11, 14, 17** to a **democratic** score; items **3, 6, 9, 12,
15, 18** to a **relating** score. Enter your 18 answers and run.

```{webr-r}
#| label: leadership-score-04
# Replace with your own 1-5 answers to items 1..18, in order:
responses <- c(2, 4, 3, 2, 5, 4,   # items 1-6
               2, 4, 4, 2, 4, 4,   # items 7-12
               2, 5, 2, 2, 4, 3)   # items 13-18

cc  <- sum(responses[c(1, 4, 7, 10, 13, 16)])
dem <- sum(responses[c(2, 5, 8, 11, 14, 17)])
rel <- sum(responses[c(3, 6, 9, 12, 15, 18)])

scores <- c("Command & Control" = cc, "Democratic" = dem, "Relating" = rel)
for (nm in names(scores)) cat(sprintf("%-18s %2d / 30\n", nm, scores[nm]))
cat(sprintf("\nYour dominant style: %s\n", names(which.max(scores))))
cat("Remember: the best PMs deliberately use all three depending on the situation.\n")
```

## Building the team

Teams develop through predictable stages, and the leader's job changes with each:

```{r fig-tuckman-04, fig.cap="Team performance through Tuckman's stages. Performance dips during storming before rising; a leader who expects the dip manages it instead of panicking.", fig.align='center', fig.width=10, fig.height=5}
stage_x <- seq(0, 10, length.out = 200)
perf <- 0.4 + 0.15 * sin(stage_x) +
        pmin(0.5, 0.06 * stage_x) -
        0.22 * exp(-((stage_x - 2.2)^2) / 0.6)
df <- data.frame(x = stage_x, perf = pmin(pmax(perf, 0.1), 1))

ggplot(df, aes(x, perf)) +
  geom_line(color = "#2E86C1", linewidth = 1.2) +
  annotate("text", x = c(1, 2.4, 4.6, 7.2, 9.4), y = 1.03,
           label = c("Forming", "Storming", "Norming", "Performing", "Adjourning"),
           fontface = "bold", size = 3.1 * diagram_text_scale) +
  annotate("segment", x = c(1.7, 3.5, 6, 8.6), xend = c(1.7, 3.5, 6, 8.6),
           y = 0.1, yend = 1, linetype = "dotted", color = "grey60") +
  coord_cartesian(ylim = c(0.1, 1.1)) +
  labs(title = "Tuckman's stages of team development",
       x = "Time together", y = "Team performance") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale)
  )
```

- **Forming** --- polite, uncertain. Leader: **direct** --- clarify goals, roles, the charter.
- **Storming** --- conflict over approach and authority; performance dips. Leader: **coach** --- surface conflict, don't suppress it.
- **Norming** --- norms settle, trust builds. Leader: **support**.
- **Performing** --- the team runs itself. Leader: **delegate**.
- **Adjourning** --- the project ends; recognize contributions and manage the transition (see [Chapter 13](13-Project-Closeout-and-Termination.qmd)).

## Motivating the team

### Worked Example 1: Herzberg --- is this a hygiene problem or a motivator problem?

Herzberg splits job factors into **hygiene factors** (pay, conditions, policy,
supervision --- their *absence* causes dissatisfaction, but their presence does not
motivate) and **motivators** (achievement, recognition, the work itself,
responsibility, growth --- these drive engagement).

Live cell --- re-classify the complaints, or add your own, and see which get fixed first.

```{webr-r}
#| label: herzberg-example-04
complaints <- data.frame(
  issue = c("Cramped, hot project room", "No recognition for hitting the drop-test milestone",
            "Confusing expense policy", "Same repetitive CAD cleanup every week",
            "Parking is far from the building"),
  type = c("Hygiene", "Motivator", "Hygiene", "Motivator", "Hygiene")
)
print(complaints, row.names = FALSE)

fix_first <- complaints$issue[complaints$type == "Hygiene"]
cat("\nFix the hygiene issues to stop dissatisfaction:\n - ",
    paste(fix_first, collapse = "\n - "), "\n", sep = "")
cat("\nThen add motivators (recognition, varied/owned work) to actually raise engagement.\n")
```

### Try it yourself: expectancy theory

Vroom's expectancy model says motivation is the product of three beliefs, each
0--1: **Expectancy** (effort → performance), **Instrumentality** (performance →
reward), and **Valence** (how much the person wants the reward).

$$\text{Motivation} = E \times I \times V$$

Because it is a product, **any factor near zero kills motivation**.

```{webr-r}
#| label: expectancy-04
team_member <- data.frame(
  name          = c("Priya", "Sam", "Dev"),
  expectancy    = c(0.9, 0.8, 0.4),   # "if I try, I can hit the target"
  instrumentality = c(0.8, 0.3, 0.9), # "hitting it gets me the reward"
  valence       = c(0.7, 0.9, 0.8)    # "I want that reward"
)
team_member$motivation <- with(team_member, expectancy * instrumentality * valence)

print(team_member, row.names = FALSE)
weakest <- team_member$name[which.min(team_member$motivation)]
cat(sprintf("\nLowest motivation: %s\n", weakest))
cat("Look at which of E, I, V is dragging the product down -- that is where the leader intervenes.\n")
cat("(Sam doesn't believe performance leads to the reward; Dev doesn't believe he can hit the target.)\n")
```

## Emotional intelligence

Goleman: *the most effective leaders all have a high degree of emotional
intelligence* --- IQ and technical skill are just the entry ticket. Five
components:

- **Self-awareness** --- a clear view of your strengths, weaknesses, drives, and ego needs.
- **Self-regulation** --- keeping yourself under control; think before you act.
- **Motivation** --- driven to your own potential *and* to drawing the best from each team member.
- **Empathy** --- weighing team members' feelings when making a decision.
- **Social skill** --- managing relationships with purpose.

## Key Takeaways

- **Management** administers systems; **leadership** inspires people. A PM needs both and leans on leadership during execution.
- Partnership rests on **exchange of purpose, the right to say no, joint accountability, and absolute honesty**. Bennis: **appreciate, remind what matters, build trust, be an ally.**
- Switch among the **five styles** --- command & control, relating, democratic, hands-on, coaching --- by situation. Plan democratically; often execute autocratically. Every style overused has a cost.
- Teams move through **forming → storming → norming → performing → adjourning**; performance dips in storming --- expect it and coach through it.
- Diagnose motivation with **Herzberg** (fix hygiene to stop dissatisfaction, add motivators to raise engagement) and **expectancy** ($E \times I \times V$ --- any factor near zero kills it).
- **Emotional intelligence** --- self-awareness, self-regulation, motivation, empathy, social skill --- separates effective project leaders from merely competent ones.

## Practice Problems

<details>
<summary>Problem 1: Pick the style</summary>

Which leadership style fits each moment on the RC-vehicle project?

1. Kicking off planning for the WBS with the whole team. → **Democratic** (buy-in, reality-based plan).
2. A moulding press fails at 2 a.m. and the launch date is at risk. → **Command & control** (crisis; take charge).
3. A capable senior designer is bored and thinking of leaving. → **Coaching** (long-term development; varied, owned work).
4. A brand-new co-op student has never used the company's PLM system. → **Command & control** initially, moving to **coaching**.
5. The team has gelled and is hitting milestones on its own. → **Delegating / supporting** --- stay out of the way.

</details>

<details>
<summary>Problem 2: Expectancy diagnosis</summary>

A skilled technician has stopped putting in effort on validation testing. In a
one-on-one you learn they are certain the extra effort won't change the delivery
date because "purchasing will be late anyway." Which expectancy factor is
broken, and what do you do?

**Answer.** **Instrumentality** --- they no longer believe performance leads to
the outcome. Restoring it means making the link visible again: show how the test
results actually gate the next milestone, remove the purchasing blocker or
re-sequence around it, and close the loop when their work does move the date.

</details>

<details>
<summary>Problem 3 (discussion): the hands-on trap</summary>

You are the strongest CAD modeller on the team, and every time someone struggles
you take the model over and finish it yourself. The work gets done and is
high-quality. Why is this a leadership failure, and what should you do instead?

**Discussion points.** Taking over is the **hands-on** style overused: it is
disempowering, tanks morale, stops people learning, and makes you the
bottleneck --- the project's throughput is now capped by your personal hours. The
**coaching** move is to ask questions that lead the person to the fix, sit with
them while they do it, and follow up --- slower once, faster forever.

</details>

## Further Resources

```{r video-resources-04, echo=FALSE, results='asis'}
embed_youtube("ReRcHdeUG9Y", "Why Leaders Eat Last (Simon Sinek, 99U)")
```
