---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Work Breakdown Structures (WBS)

```{r setup-05, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Explain** what a WBS is and how it links scope, schedule, and budget.
2. **Decompose** a project into deliverables, work packages, and activities.
3. **Apply** the 100 % rule, the mutually-exclusive rule, and "plan outcomes, not actions."
4. **Choose** an appropriate level of decomposition using the 8/80 and single-reporting-period rules of thumb.
5. **Number** a WBS and write a WBS dictionary entry for a work package.
6. **Roll up** cost and effort from work packages to the total project.
7. **State** the six requirements of a valid project contract and the purpose of work authorization.

<details>
<summary>Before we begin: how do you eat an elephant?</summary>

One bite at a time. *"Project managers cannot manage projects. Project managers
manage pieces of projects."* The WBS is how you cut the elephant into bites.

</details>

## Introduction: from a scope statement to something you can manage

You have a signed scope statement for the RC-vehicle replica. It is still far
too big to schedule, budget, or assign. The **work breakdown structure** is the
bridge: it takes the total scope and breaks it into **deliverables**, then
**work packages**, then **activities** small enough that one person can own each
one and you can put a real number on it.

> **Definition.** A WBS is a **deliverable-oriented** grouping of project
> components that organizes and defines the **total scope** of the project.

A well-built WBS lets the team **plan every step**, **collect and organize** the
information each task needs, and **assign** manageable chunks to team members. It
also enables the **WBS dictionary**, which pins down the requirements for each
task.

::: {.callout-note}
**Industry snapshot.** An automotive program WBS separates *body*, *powertrain
integration*, *electrical*, *validation*, and *launch* --- each a deliverable, not
an action. A food-plant install WBS breaks out *utilities*, *process equipment*,
*controls*, *sanitation qualification*, and *commissioning*. A defense contract
WBS follows a mandated standard (MIL-STD-881) so the customer can track cost
against the exact same structure the contractor uses.
:::

```{r video-wbs-05, echo=FALSE, results='asis'}
embed_youtube("B8q4uz5XiZc", "Work Breakdown Structure (WBS) Explained")
```

## The hierarchy

```{r fig-wbs-tree-05, fig.cap="A partial WBS for the RC-vehicle project. Level 1 is the project; Level 2 the major deliverables; Level 3 the work packages. Numbering encodes the hierarchy.", fig.align='center', fig.width=13, fig.height=6}
w11 <- 0.5; w12 <- 3.2; w21 <- 6.0; w22 <- 8.7; w31 <- 11.4; w41 <- 14.1; w42 <- 16.8
d1 <- mean(c(w11, w12)); d2 <- mean(c(w21, w22)); d3 <- w31; d4 <- mean(c(w41, w42))
root <- mean(c(d1, d2, d3, d4))

nodes <- data.frame(
  id = c("root", "d1", "d2", "d3", "d4",
         "w11", "w12", "w21", "w22", "w31", "w41", "w42"),
  label = c("1.0 RC-Vehicle Replica Project",
            "1.1 Body & Frame", "1.2 Drive & Controls", "1.3 Assembly & Test", "1.4 Launch",
            "1.1.1 Outer body\ndesign", "1.1.2 Inner frame\ndesign",
            "1.2.1 Motor/controller\nprocurement", "1.2.2 Wiring\nharness",
            "1.3.1 Prototype\nassembly", "1.4.1 Packaging\ndesign", "1.4.2 Compliance\n& approvals"),
  x = c(root,  d1, d2, d3, d4,   w11, w12,  w21, w22,  w31,  w41, w42),
  y = c(3,  2,   2, 2,   2,   1,   1,    1,   1,    1,    1,   1)
)
edges <- data.frame(
  x = c(root,root,root,root, d1,d1, d2,d2, d3, d4,d4),
  y = c(3,3,3,3, 2,2, 2,2, 2, 2,2),
  xend = c(d1,d2,d3,d4, w11,w12, w21,w22, w31, w41,w42),
  yend = c(2,2,2,2, 1,1, 1,1, 1, 1,1)
)
ggplot() +
  geom_segment(data = edges, aes(x, y, xend = xend, yend = yend),
               color = "grey55", linewidth = 0.6) +
  geom_label(data = nodes, aes(x, y, label = label),
             size = 2.5 * diagram_text_scale, lineheight = 0.9,
             fill = c("#1B4F72", rep("#5DADE2", 4), rep("#D6EAF8", 7)),
             color = c("white", rep("white", 4), rep("black", 7)),
             fontface = "bold", label.size = 0) +
  annotate("text", x = w11 - 1.4, y = c(3, 2, 1),
           label = c("Level 1", "Level 2", "Level 3"),
           fontface = "italic", size = 3 * diagram_text_scale, hjust = 1) +
  coord_cartesian(xlim = c(w11 - 2.6, w42 + 1.4), ylim = c(0.5, 3.5)) +
  labs(title = "WBS: project -> deliverables -> work packages") +
  theme_void() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale,
                                  face = "bold", hjust = 0.5))
```

- **Level 1** --- the project. *What is the project?*
- **Level 2** --- the summary level: the project's major deliverables.
- **Level 3** --- the start of the **work package** level (more levels if the package is complex).

The same content can be shown as a **tree** (good for communicating structure)
or a **tabular / indented list** (good for working). Both carry the numbering.

## Design principles

### The 100 % rule

The WBS includes **100 % of the work** in the scope --- all deliverables, internal,
external, and interim --- and **no more**. At every level, the **sum of the
children equals the parent**, and nothing outside the scope appears.

### Worked Example 1: does this WBS roll up to 100 %?

The "1.1 Body & Frame" deliverable is estimated at **340 hours**. Its four work
packages are estimated at 120, 90, 70, and 60 hours.

Live cell --- change the parent estimate or the work-package hours and watch the 100 % check.

```{r calc-wbs-rollup-example-05, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "wbs-rollup-example-05",
  title = "WBS 100% Rule Check",
  icon = "sitemap",
  inputs = list(
    qwebr_calc_input("parent_hours", "Parent estimate (1.1 Body & Frame)", value = 340, min = 50, max = 800, step = 5, suffix = "h"),
    qwebr_calc_input("h_outer", "1.1.1 Outer body design", value = 120, min = 0, max = 400, step = 5, suffix = "h"),
    qwebr_calc_input("h_inner", "1.1.2 Inner frame design", value = 90, min = 0, max = 400, step = 5, suffix = "h"),
    qwebr_calc_input("h_select", "1.1.3 Material selection", value = 70, min = 0, max = 400, step = 5, suffix = "h"),
    qwebr_calc_input("h_review", "1.1.4 Design review & release", value = 60, min = 0, max = 400, step = 5, suffix = "h")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: wbs-rollup-example-05
parent_hours <- 340
h_outer  <- 120   # 1.1.1 Outer body design
h_inner  <- 90    # 1.1.2 Inner frame design
h_select <- 70    # 1.1.3 Material selection
h_review <- 60    # 1.1.4 Design review & release

children_hours <- c("1.1.1 Outer body design" = h_outer,
                    "1.1.2 Inner frame design" = h_inner,
                    "1.1.3 Material selection" = h_select,
                    "1.1.4 Design review & release" = h_review)

child_sum <- sum(children_hours)

# Chart: the four children stacked beside the parent estimate
stack <- rbind(cbind(children_hours, 0), c(0, parent_hours))
colnames(stack) <- c("Children", "Parent")
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
barplot(stack, col = c("#2980b9", "#16a085", "#8e44ad", "#e67e22", "#34495e"), ylim = c(0, max(child_sum, parent_hours) * 1.1), ylab = "Hours",
        main = "Do the children add up to the parent?")
abline(h = parent_hours, lty = 2, lwd = 3, col = "#e74c3c")

if (child_sum == parent_hours) {
  cat("OK: the children account for exactly 100% of the parent\n")
} else if (child_sum < parent_hours) {
  cat(sprintf("!!GAP: %g h of the parent's scope is in no work package!!\n", parent_hours - child_sum))
} else {
  cat(sprintf("!!OVER: the children exceed the parent by %g h (scope crept in)!!\n", child_sum - parent_hours))
}
cat(sprintf("Sum of children  %g h\n", child_sum))
cat(sprintf("Parent estimate  %g h\n", parent_hours))
cat(sprintf("Children as share of parent  %.1f%%\n", 100 * child_sum / parent_hours))
```

</details>

Here the children sum to 340 --- the deliverable is fully decomposed with nothing
missing and nothing extra.

### Mutually exclusive

**No overlap** between WBS elements. Overlap causes duplicated work, confusion
over who is responsible, and messy cost accounting. If element names are
ambiguous, the **WBS dictionary** disambiguates them.

### Plan outcomes, not actions

Define elements as **results** ("Training module completed"), not **actions**
("write script", "record video", "upload"). Action lists always capture too many
or too few steps --- breaking the 100 % rule --- and they over-prescribe *how* the
work is done, killing the team's ingenuity.

### Try it yourself: outcome or action?

```{r tryit-outcome-05, echo=FALSE, results='asis'}
oa_bins <- c("Outcome (a good WBS element)", "Action (rewrite it)")
tryit_sort(
  "outcome-or-action", "Is each WBS entry an outcome or an action?",
  bins = oa_bins,
  cards = data.frame(
    text = c("Compressor skid commissioned", "Write the test report", "Supplier selected and contracted",
             "Call three vendors", "Drawings released to manufacturing", "Run the line for two shifts",
             "Training module completed", "Record the demo video"),
    bin = c("Outcome (a good WBS element)", "Action (rewrite it)", "Outcome (a good WBS element)",
            "Action (rewrite it)", "Outcome (a good WBS element)", "Action (rewrite it)",
            "Outcome (a good WBS element)", "Action (rewrite it)"),
    why = c("A result you can mark done or not done.", "A verb phrase that prescribes how: rewrite as 'Test report issued'.",
            "A result.", "An action: rewrite as 'Vendor quotes obtained'.", "A result.",
            "An action: rewrite as 'Process capability demonstrated'.", "A result.",
            "An action: rewrite as 'Demo video published'."),
    stringsAsFactors = FALSE),
  intro = "Outcomes name a result; actions name a task.", question = NULL)

tryit_sort(
  "outcome-twin", "Outcome or action? (graded)", bins = oa_bins, show = FALSE,
  cards = data.frame(
    text = c("Welding jig qualified", "Email the drawings to the customer", "Packaging design approved",
             "Measure twenty samples", "Operator training delivered"),
    bin = c("Outcome (a good WBS element)", "Action (rewrite it)", "Outcome (a good WBS element)",
            "Action (rewrite it)", "Outcome (a good WBS element)"),
    stringsAsFactors = FALSE),
  question = "<p>A WBS should plan outcomes, not actions. For each entry, decide whether it is an outcome or an action.</p>")
```

### Try it yourself: does it roll up to 100 %?

```{r tryit-hundred-05, echo=FALSE, results='asis'}
hundred_bins <- c("OK", "Gap: scope missing", "Over: children exceed parent", "Overlap: not mutually exclusive")
tryit_sort(
  "hundred-rule", "What is wrong with each decomposition, if anything?",
  bins = hundred_bins,
  cards = data.frame(
    text = c("Parent 200 h; children 80 + 70 + 50 h",
             "Parent 300 h; children 120 + 100 + 60 h",
             "Parent 150 h; children 70 + 60 + 40 h",
             "'Prototype assembly' and 'Assemble prototype and rework' are listed under the same parent",
             "Parent 420 h; children 150 + 140 + 130 h",
             "Parent 90 h; children 40 + 30 h",
             "Two packages both include 'prepare test fixtures' in their descriptions",
             "Parent 250 h; children 100 + 90 + 80 h"),
    bin = c("OK", "Gap: scope missing", "Over: children exceed parent", "Overlap: not mutually exclusive",
            "OK", "Gap: scope missing", "Overlap: not mutually exclusive", "Over: children exceed parent"),
    why = c("200 = 200: children sum to the parent.", "280 is 20 h short of 300: some scope is in no package.",
            "170 exceeds 150 by 20 h: scope has crept in or the parent is understated.",
            "The same work appears twice: duplicated effort and unclear ownership.",
            "420 = 420.", "70 is 20 h short of 90.", "Shared work is counted twice and nobody owns it.",
            "270 exceeds 250 by 20 h."),
    stringsAsFactors = FALSE),
  intro = "Add the children and compare with the parent. Pick one verdict per card.", question = NULL)

tryit_sort(
  "hundred-twin", "The 100 % rule (graded)", bins = hundred_bins, show = FALSE,
  cards = data.frame(
    text = c("Parent 180 h; children 90 + 60 + 30 h", "Parent 360 h; children 150 + 120 + 70 h",
             "Parent 120 h; children 60 + 40 + 40 h",
             "'Install guard' and 'Install guard and test interlock' sit under one parent",
             "Parent 75 h; children 45 + 30 h"),
    bin = c("OK", "Gap: scope missing", "Over: children exceed parent", "Overlap: not mutually exclusive", "OK"),
    stringsAsFactors = FALSE),
  question = "<p>For each decomposition, decide whether it satisfies the 100 % rule, and if not, what is wrong with it.</p>")
```

## Level of detail

A work package at the **activity level** is a task that:

- can be **realistically and confidently estimated**;
- makes no practical sense to break down further;
- produces a **measurable deliverable**; and
- forms a unique package that **could be outsourced or contracted out**.

**Rules of thumb for how far to decompose:**

- **80-hour rule** --- no work package (single deliverable at the lowest level) exceeds ~80 hours of effort.
- **One reporting period** --- no package is longer than a single progress-reporting period (if you report monthly, nothing runs longer than a month).
- **8-hour floor** --- generally do not define a task below ~8 hours; below that you are micromanaging.
- **"If it makes sense"** --- apply common sense.

### Try it yourself: decomposition check

```{webr-r}
#| label: wbs-decomp-05
packages <- data.frame(
  id    = c("1.1.1", "1.2.1", "1.3.1", "1.3.4", "1.4.2"),
  name  = c("Outer body design", "Motor/controller procurement",
            "Prototype assembly", "Tighten one fastener",
            "Full regulatory approval campaign"),
  hours = c(64, 40, 72, 1.5, 260)
)

flag <- ifelse(packages$hours > 80, "TOO BIG  (> 80 h -- split it)",
        ifelse(packages$hours < 8,  "TOO SMALL (< 8 h -- roll it up)",
               "ok"))
out <- cbind(packages, verdict = flag)
print(out, row.names = FALSE)

cat(sprintf("\n%d of %d packages need re-sizing.\n",
            sum(flag != "ok"), nrow(packages)))
```

### Try it yourself: does it fit inside one reporting period?

The 8/80 rule checks **effort hours**; the one-reporting-period rule checks
**calendar span** against how often you report --- a different failure mode. A
package can be well under 80 hours and still run longer than your reporting
cycle if it is thin, spread-out work. Live cell --- change the reporting
cadence or a package's span and see which packages need to be split or
tracked by percent-complete instead.

```{r calc-wbs-reporting-period-05, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "wbs-reporting-period-05",
  title = "Reporting-Cycle Check",
  icon = "calendar-check",
  inputs = list(
    qwebr_calc_input("reporting_cycle_days", "Reporting cycle", value = 14, min = 3, max = 30, step = 1, suffix = "days", type = "slider"),
    qwebr_calc_input("span_concept", "1.1.1 Concept CAD", value = 10, min = 1, max = 60, step = 1, suffix = "days"),
    qwebr_calc_input("span_motor", "1.2.1 Motor/controller procurement", value = 25, min = 1, max = 60, step = 1, suffix = "days"),
    qwebr_calc_input("span_drop", "1.5.1 Drop/durability test", value = 5, min = 1, max = 60, step = 1, suffix = "days"),
    qwebr_calc_input("span_cert", "1.5.2 Toy-safety certification", value = 12, min = 1, max = 60, step = 1, suffix = "days")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: wbs-reporting-period-05
reporting_cycle_days <- 14   # biweekly status reports

span_concept <- 10    # 1.1.1 Concept CAD
span_motor   <- 25    # 1.2.1 Motor/controller procurement
span_drop    <- 5     # 1.5.1 Drop/durability test
span_cert    <- 12    # 1.5.2 Toy-safety certification

packages <- data.frame(
  id        = c("1.1.1", "1.2.1", "1.5.1", "1.5.2"),
  name      = c("Concept CAD", "Motor/controller procurement",
               "Drop/durability test", "Toy-safety certification"),
  span_days = c(span_concept, span_motor, span_drop, span_cert)
)
packages$too_long <- packages$span_days > reporting_cycle_days
n_bad <- sum(packages$too_long)

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
barplot(packages$span_days, names.arg = packages$id, col = ifelse(packages$too_long, "#e74c3c", "#16a085"),
        ylim = c(0, max(packages$span_days, reporting_cycle_days) * 1.15),
        ylab = "Span (days)", main = "Work-package span vs. reporting cycle")
abline(h = reporting_cycle_days, lty = 2, lwd = 3, col = "#2c3e50")

if (n_bad > 0) {
  cat(sprintf("!!%d of %d packages need attention!!\n", n_bad, nrow(packages)))
} else {
  cat(sprintf("All %d packages fit within one reporting cycle\n", nrow(packages)))
}
cat(sprintf("Reporting cadence  every %d days\n", reporting_cycle_days))
for (i in seq_len(nrow(packages))) {
  verdict <- if (packages$too_long[i]) {
    sprintf("!!%g days: split or report %% complete!!", packages$span_days[i])
  } else {
    sprintf("%g days: fits within one cycle", packages$span_days[i])
  }
  cat(sprintf("%s %s  %s\n", packages$id[i], packages$name[i], verdict))
}
```

</details>

### Try it yourself: is the package the right size?

```{r tryit-size-05, echo=FALSE, results='asis'}
size_bins <- c("About right", "Too big: split it", "Too small: roll it up", "Longer than a reporting period")
tryit_sort(
  "package-size", "Is each work package the right size? (status is reported weekly)",
  bins = size_bins,
  cards = data.frame(
    text = c("Fixture design: 40 h over 5 days",
             "Full environmental test campaign: 210 h over 3 weeks",
             "Tighten the guard screws: 1 h",
             "Supplier qualification audit: 30 h of effort spread over 3 weeks of waiting for replies",
             "Wiring harness layout: 72 h over 5 days",
             "Label the sample boxes: 2 h",
             "Complete tooling design: 150 h",
             "Prototype shake-down: 24 h over 4 weeks of intermittent runs"),
    bin = c("About right", "Too big: split it", "Too small: roll it up", "Longer than a reporting period",
            "About right", "Too small: roll it up", "Too big: split it", "Longer than a reporting period"),
    why = c("Between 8 h and 80 h and inside one week.", "Over 80 h.", "Under 8 h: micromanaging.",
            "The effort is fine but the span is longer than the weekly reporting period: split it or report % complete.",
            "Between 8 h and 80 h and inside one week.", "Under 8 h: roll it into a larger package.",
            "Over 80 h.", "Effort is fine; the span exceeds one reporting period."),
    stringsAsFactors = FALSE),
  intro = "Apply the 8 h floor, the 80 h ceiling and the one-reporting-period rule.", question = NULL)

tryit_sort(
  "package-size-twin", "Work package size (graded)", bins = size_bins, show = FALSE,
  cards = data.frame(
    text = c("Machine-vision commissioning: 120 h", "Deburr one bracket: 1.5 h",
             "Gauge calibration: 16 h over 2 days",
             "Customer approval loop: 12 h of effort over 4 weeks",
             "Wiring check: 60 h over 5 days"),
    bin = c("Too big: split it", "Too small: roll it up", "About right",
            "Longer than a reporting period", "About right"),
    stringsAsFactors = FALSE),
  question = "<p>Status is reported weekly. Using the 8-hour floor, the 80-hour ceiling and the one-reporting-period rule, decide whether each work package is the right size.</p>")
```

## Why the WBS matters

```{r fig-wbs-hub-05, fig.cap="The WBS feeds four downstream planning activities. Get it wrong and the estimate, the schedule, the staffing plan, and the risk register are all built on sand.", fig.align='center', fig.width=9, fig.height=5.5}
hub <- data.frame(x = 0, y = 0)
spokes <- data.frame(
  x = c(0, 1.3, 0, -1.3),
  y = c(1.3, 0, -1.3, 0),
  label = c("Estimating\n(cost & time)", "Network diagram\n(schedule)",
            "Risk\nidentification", "Staffing\n& responsibility")
)
ggplot() +
  geom_segment(data = spokes, aes(x = 0, y = 0, xend = x, yend = y),
               color = "grey55", linewidth = 0.7) +
  geom_point(data = hub, aes(x, y), size = 46, color = "#1B4F72") +
  annotate("text", x = 0, y = 0, label = "WBS", color = "white",
           fontface = "bold", size = 5 * diagram_text_scale) +
  geom_label(data = spokes, aes(x, y, label = label),
             size = 3 * diagram_text_scale, lineheight = 0.9,
             fill = "#D6EAF8", label.size = 0, fontface = "bold") +
  coord_equal(xlim = c(-2.2, 2.2), ylim = c(-2.1, 2.1)) +
  labs(title = "Everything downstream depends on the WBS") +
  theme_void() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale,
                                  face = "bold", hjust = 0.5))
```

## The WBS dictionary

The dictionary is used to **control what work is done and when**, **prevent scope
creep**, and **capture the effort** for each task. Physically, teams post the
work packages on a wall in logical work-flow order; **removing them in numerical
order generates the dictionary**. Each entry --- a **task description** --- records:

- Task description (what work is authorized)
- Quantified objectives and deliverables (cross-referenced to the charter)
- Product description; acceptance criteria (how the owner checks their own work)
- Assumptions; resources assigned; duration; cost; due date
- Interdependencies (before this task / after this task)
- Project manager approval and date

## Work authorization and contracts

Once the scope is decomposed into work packages, stakeholders formally agree on
project content --- the **work authorization**. If the client is external, this
includes the **contract sign-off**.

A valid project-management contract requires all six of:

1. The legal **right** to contract
2. Legal **age** (18 in Ontario and Quebec; 19 or 21 elsewhere in Canada)
3. **Sound mind** at signing
4. Valid, legal **consideration** for both parties
5. A **legal scope of work**
6. **Offer and acceptance**

Contracts also commonly contain: excusable delays; allowable costs; **liquidated
damages** (non-performance / penalty clauses); inspection criteria;
responsibility for correcting defects; and dispute-resolution mechanisms.

### Try it yourself: cost roll-up from the WBS

```{webr-r}
#| label: wbs-cost-rollup-05
# Work-package costs (direct), grouped by Level-2 deliverable.
wp <- data.frame(
  deliverable = c(rep("1.1 Body & Frame", 3),
                  rep("1.2 Drive & Controls", 2),
                  rep("1.3 Assembly & Test", 3),
                  rep("1.4 Launch", 2)),
  package = c("Outer body", "Inner frame", "Material selection",
              "Motor/controller", "Wiring harness",
              "Prototype assembly", "Drop test", "EMC test",
              "Packaging design", "Approvals"),
  cost = c(18000, 14000, 6000, 140000, 9000, 21000, 12000, 15000, 8000, 6000)
)

by_deliverable <- tapply(wp$cost, wp$deliverable, sum)
cat("Roll-up to Level 2:\n")
for (d in names(by_deliverable))
  cat(sprintf("  %-22s $%s\n", d, format(by_deliverable[d], big.mark = " ")))

cat(sprintf("\nLevel 1 (total project direct cost): $%s\n",
            format(sum(wp$cost), big.mark = " ")))
cat("\nThis roll-up only works because the WBS obeys the 100% and\n")
cat("mutually-exclusive rules -- every dollar is counted once, and all of them.\n")
```

### Try it yourself: roll the hours up the tree

The leaf work packages carry their hours. Work up the tree: each deliverable
is the sum of its packages, and the project is the sum of the deliverables.
Each team picks its own **data set**.

```{r tryit-rollup-05, echo=FALSE, results='asis'}
wbs_structure <- list(
  list(id = "1.1", label = "1.1 Enclosure",
       kids = c("1.1.1 Housing design", "1.1.2 Tooling quote", "1.1.3 Prototype housing")),
  list(id = "1.2", label = "1.2 Electronics",
       kids = c("1.2.1 PCB layout", "1.2.2 Firmware")),
  list(id = "1.3", label = "1.3 Test and release",
       kids = c("1.3.1 Compliance test", "1.3.2 Release review")))
wbs_hours <- list(c(60, 35, 70, 80, 95, 50, 30), c(45, 40, 60, 70, 110, 55, 25),
                  c(80, 20, 50, 60, 75, 40, 45), c(30, 50, 40, 90, 85, 65, 20))
wbs_build <- function(h) {
  nodes <- list(); edges <- list(); k <- 0; sums <- c()
  for (d in wbs_structure) {
    n <- length(d$kids)
    idx <- k + seq_len(n); k <- k + n
    sums[d$id] <- sum(h[idx])
    nodes[[length(nodes) + 1]] <- tryit_node(d$id, d$label, 2,
      blank = list(tryit_answer("Hours", sum(h[idx]), tol = 0, digits = 0)))
    edges[[length(edges) + 1]] <- c("1.0", d$id)
    for (i in seq_len(n)) {
      leaf_id <- strsplit(d$kids[i], " ")[[1]][1]
      nodes[[length(nodes) + 1]] <- tryit_node(leaf_id, d$kids[i], 3,
        given = c(Hours = sprintf("%g h", h[idx[i]])))
      edges[[length(edges) + 1]] <- c(d$id, leaf_id)
    }
  }
  root <- tryit_node("1.0", "1.0 Charging-station kit", 1,
                     blank = list(tryit_answer("Hours", sum(h), tol = 0, digits = 0)))
  list(nodes = c(list(root), nodes), edges = edges,
       answers = list(tryit_answer("Share of the project in 1.2 Electronics", 100 * sums[["1.2"]] / sum(h),
                                   tol = 0.5, suffix = "%")))
}
w1 <- wbs_build(wbs_hours[[1]])
tryit_diagram(
  "wbs-rollup-tree", "Charging-station kit: WBS hours", w1$nodes, w1$edges,
  direction = "down", arrows = FALSE, answers = w1$answers,
  teams = lapply(wbs_hours[-1], function(h) { w <- wbs_build(h); list(nodes = w$nodes, answers = w$answers) }),
  intro = "Level 1 is the project, Level 2 the deliverables, Level 3 the work packages.")

wt <- wbs_build(c(35, 45, 55, 100, 60, 30, 40))
tryit_diagram(
  "wbs-rollup-twin", "WBS hours roll-up (graded)", wt$nodes, wt$edges, direction = "down",
  arrows = FALSE, answers = wt$answers, show = FALSE,
  question = paste0("<p>The charging-station kit WBS below has work-package hours as given. Roll the ",
    "hours up to each deliverable and to the project total, then give the share of the project in ",
    "1.2 Electronics (nearest whole per cent). Packages: 1.1.1 = 35 h, 1.1.2 = 45 h, 1.1.3 = 55 h, ",
    "1.2.1 = 100 h, 1.2.2 = 60 h, 1.3.1 = 30 h, 1.3.2 = 40 h.</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

For data set 1: 1.1 = 60 + 35 + 70 = 165 h; 1.2 = 80 + 95 = 175 h; 1.3 =
50 + 30 = 80 h; the project is 165 + 175 + 80 = **420 h**. The electronics
branch is 175 / 420 = 42 % of the work. The roll-up only works because every
hour is in exactly one package: the 100 % rule and mutual exclusivity.

</details>

## Key Takeaways

- A **WBS** is a **deliverable-oriented** decomposition of the **total scope** into deliverables → work packages → activities. PMs manage the pieces.
- **Numbering** encodes the hierarchy: Level 1 project, Level 2 deliverables, Level 3+ work packages.
- Obey the three design rules: **100 % rule** (children sum to the parent, never more), **mutually exclusive** (no overlap), and **plan outcomes, not actions**.
- Size work packages with the **8/80 rule** and the **one-reporting-period** rule; below ~8 h you are micromanaging, above ~80 h you cannot estimate confidently.
- The WBS feeds **estimating, scheduling, staffing, and risk** --- errors propagate to all four.
- The **WBS dictionary** locks down each task's description, acceptance criteria, resources, and interdependencies, and helps prevent scope creep.
- **Work authorization** formalizes agreement on content; a valid contract needs all **six** legal elements and usually carries **liquidated-damages** and dispute clauses.

## Case Study: Meridian R/C — Week 5

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 5: Cutting the Trailblazer Into Bites"}

You're three weeks into the Meridian Trailblazer prototype, and Daniel Cho's
$800,000 charter ceiling is signed. It's time to stop describing the project
and start decomposing it. This week you sat down with Priya Nandakumar
(tooling), Marcus Webb (procurement), and Sofia Reyes (quality and test) to
turn "build a prototype that passes Kestrel Retail Group's durability testing
and toy-safety certification" into a WBS the team can actually schedule,
budget, and staff.

You started at Level 1 with **1.0 Meridian Trailblazer Prototype** and split
it into seven Level-2 deliverables, each owned by a named function so nobody
can later claim a piece of scope belongs to someone else:

- **1.1 Design & Engineering** — 1.1.1 Concept CAD, 1.1.2 Detailed design release
- **1.2 Tooling & Fabrication** — 1.2.1 Body tooling, 1.2.2 Frame tooling
- **1.3 Motor & Controls Integration**
- **1.4 Prototype Assembly**
- **1.5 Testing & Certification** — 1.5.1 Drop/durability test, 1.5.2 Toy-safety certification
- **1.6 Packaging & Client Sample Shipment**
- **1.7 Project Management** (cross-cutting)

Notice what this structure is *not*. It is not "call the tooling vendor" or
"email Sofia the test plan" — every element names a deliverable or result
(a released design, a qualified tool, an integrated drive system, a
certified sample) rather than an action, exactly as the outcomes-not-actions
rule demands. Priya pushed back once, wanting to fold "1.2 Tooling &
Fabrication" and "1.3 Motor & Controls Integration" into one bucket since
her team touches both — but that would blur ownership between tooling
(hers) and the MCS/Paragon procurement decision (Marcus's), so you kept them
mutually exclusive instead.

You also ran the 100% rule on 1.2: body tooling (1.2.1) and frame tooling
(1.2.2) are the *only* children of Tooling & Fabrication — nothing about
motor sourcing or assembly labour is allowed to hide inside that branch, and
nothing about tooling is allowed to leak into 1.3 or 1.4. And you sized
1.7 Project Management as its own Level-2 element specifically so PM effort
never gets smuggled into a technical deliverable's hours — the same mistake
you'll later coach a student through in the practice problems. Sofia flagged
that 1.5's two children stay far enough apart in scope — durability is
physical stress testing, certification is regulatory sign-off — that a
failure in one doesn't get miscoded as a failure in the other when Kestrel's
100,000-unit decision comes down to those two checkboxes.

:::

## Practice Problems

<details>
<summary>Problem 1: Fix the 100 % rule violation</summary>

The "1.3 Assembly & Test" deliverable is estimated at 300 h. Its packages are
listed as: Prototype assembly (110 h), Drop test (60 h), EMC test (70 h),
Prototype assembly rework (40 h), Write the test report (30 h), *and* "General
project management (50 h)."

What is wrong, and how do you fix it?

<details>
<summary>Solution</summary>

The children sum to 360 h --- 60 h **over** 100 % of the parent. Two
problems: "Prototype assembly rework" overlaps "Prototype assembly" (not
mutually exclusive --- fold rework into the assembly package or make it a
contingency), and "General project management" is **not part of this
deliverable's scope** --- PM effort belongs in its own Level-2 element (e.g.
"1.0.1 Project Management"), not smuggled into a technical deliverable.

</details>

</details>

<details>
<summary>Problem 2: Outcome or action?</summary>

Rewrite these action-style entries as outcome-style WBS elements:

1. "Call three carbon-fibre suppliers and get quotes"
2. "Run the mould 20 times and measure the parts"
3. "Email the drawings to the customer for approval"

<details>
<summary>Model answers</summary>

1. "Carbon-fibre supplier selected and quoted."
2. "Moulding process capability established."
3. "Customer design approval obtained."

Each is a result you can mark done or not done, and it does not
dictate the method.

</details>

</details>

<details>
<summary>Problem 3 (discussion): how deep should the RC-vehicle WBS go?</summary>

Your draft WBS has a work package "Design the outer body" estimated at 180 h and
spanning 6 weeks, and another, "Deburr the frame jig", estimated at 2 h. Using
the rules of thumb, what do you do with each?

<details>
<summary>Discussion points</summary>

"Design the outer body" breaks both the 80-hour rule and
(if you report every 2 weeks) the single-reporting-period rule --- split it into
concept, detailed CAD, and design-release packages. "Deburr the frame jig" is
below the 8-hour floor --- roll it into the jig-fabrication package rather than
tracking it as its own line.

</details>

</details>

## Further Resources

```{r video-resources-05, echo=FALSE, results='asis'}
embed_youtube("akO2Lf1fHmM", "What Is a Work Breakdown Structure? (Online PM Courses)")
embed_youtube("_7R_jjxVpK4", "Work Breakdown Structure Basics")
```
