---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Work Breakdown Structures (WBS)

```{r setup-05, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Explain** what a WBS is and how it links scope, schedule, and budget.
2. **Decompose** a project into deliverables, work packages, and activities.
3. **Apply** the 100 % rule, the mutually-exclusive rule, and "plan outcomes, not actions."
4. **Choose** an appropriate level of decomposition using the 8/80 and single-reporting-period rules of thumb.
5. **Number** a WBS and write a WBS dictionary entry for a work package.
6. **Roll up** cost and effort from work packages to the total project.
7. **State** the six requirements of a valid project contract and the purpose of work authorization.

<details>
<summary>Before we begin: how do you eat an elephant?</summary>

One bite at a time. *"Project managers cannot manage projects. Project managers
manage pieces of projects."* The WBS is how you cut the elephant into bites.

</details>

## Introduction: from a scope statement to something you can manage

You have a signed scope statement for the RC-vehicle replica. It is still far
too big to schedule, budget, or assign. The **work breakdown structure** is the
bridge: it takes the total scope and breaks it into **deliverables**, then
**work packages**, then **activities** small enough that one person can own each
one and you can put a real number on it.

> **Definition.** A WBS is a **deliverable-oriented** grouping of project
> components that organizes and defines the **total scope** of the project.

A well-built WBS lets the team **plan every step**, **collect and organize** the
information each task needs, and **assign** manageable chunks to team members. It
also enables the **WBS dictionary**, which pins down the requirements for each
task.

::: {.callout-note}
**Industry snapshot.** An automotive program WBS separates *body*, *powertrain
integration*, *electrical*, *validation*, and *launch* --- each a deliverable, not
an action. A food-plant install WBS breaks out *utilities*, *process equipment*,
*controls*, *sanitation qualification*, and *commissioning*. A defense contract
WBS follows a mandated standard (MIL-STD-881) so the customer can track cost
against the exact same structure the contractor uses.
:::

```{r video-wbs-05, echo=FALSE, results='asis'}
embed_youtube("B8q4uz5XiZc", "Work Breakdown Structure (WBS) Explained")
```

## The hierarchy

```{r fig-wbs-tree-05, fig.cap="A partial WBS for the RC-vehicle project. Level 1 is the project; Level 2 the major deliverables; Level 3 the work packages. Numbering encodes the hierarchy.", fig.align='center', fig.width=11, fig.height=6}
nodes <- data.frame(
  id = c("root", "d1", "d2", "d3", "d4",
         "w11", "w12", "w21", "w22", "w31", "w41", "w42"),
  label = c("1.0 RC-Vehicle Replica Project",
            "1.1 Body & Frame", "1.2 Drive & Controls", "1.3 Assembly & Test", "1.4 Launch",
            "1.1.1 Outer body\ndesign", "1.1.2 Inner frame\ndesign",
            "1.2.1 Motor/controller\nprocurement", "1.2.2 Wiring\nharness",
            "1.3.1 Prototype\nassembly", "1.4.1 Packaging\ndesign", "1.4.2 Compliance\n& approvals"),
  x = c(5,  1.5, 4, 6.5, 9,   0.6, 2.4,  3.3, 4.7,  6.5,  8.2, 9.8),
  y = c(3,  2,   2, 2,   2,   1,   1,    1,   1,    1,    1,   1)
)
edges <- data.frame(
  x = c(5,5,5,5, 1.5,1.5, 4,4, 6.5, 9,9),
  y = c(3,3,3,3, 2,2, 2,2, 2, 2,2),
  xend = c(1.5,4,6.5,9, 0.6,2.4, 3.3,4.7, 6.5, 8.2,9.8),
  yend = c(2,2,2,2, 1,1, 1,1, 1, 1,1)
)
ggplot() +
  geom_segment(data = edges, aes(x, y, xend = xend, yend = yend),
               color = "grey55", linewidth = 0.6) +
  geom_label(data = nodes, aes(x, y, label = label),
             size = 2.5 * diagram_text_scale, lineheight = 0.9,
             fill = c("#1B4F72", rep("#5DADE2", 4), rep("#D6EAF8", 7)),
             color = c("white", rep("white", 4), rep("black", 7)),
             fontface = "bold", label.size = 0) +
  annotate("text", x = -0.2, y = c(3, 2, 1),
           label = c("Level 1", "Level 2", "Level 3"),
           fontface = "italic", size = 3 * diagram_text_scale, hjust = 1) +
  coord_cartesian(xlim = c(-1.2, 10.6), ylim = c(0.5, 3.5)) +
  labs(title = "WBS: project -> deliverables -> work packages") +
  theme_void() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale,
                                  face = "bold", hjust = 0.5))
```

- **Level 1** --- the project. *What is the project?*
- **Level 2** --- the summary level: the project's major deliverables.
- **Level 3** --- the start of the **work package** level (more levels if the package is complex).

The same content can be shown as a **tree** (good for communicating structure)
or a **tabular / indented list** (good for working). Both carry the numbering.

## Design principles

### The 100 % rule

The WBS includes **100 % of the work** in the scope --- all deliverables, internal,
external, and interim --- and **no more**. At every level, the **sum of the
children equals the parent**, and nothing outside the scope appears.

### Worked Example 1: does this WBS roll up to 100 %?

The "1.1 Body & Frame" deliverable is estimated at **340 hours**. Its four work
packages are estimated at 120, 90, 70, and 60 hours.

Live cell --- change the parent estimate or the work-package hours and watch the 100 % check.

```{webr-r}
#| label: wbs-rollup-example-05
parent_hours   <- 340
children_hours <- c("1.1.1 Outer body design" = 120,
                    "1.1.2 Inner frame design" = 90,
                    "1.1.3 Material selection" = 70,
                    "1.1.4 Design review & release" = 60)

child_sum <- sum(children_hours)
cat("Children:\n"); print(children_hours)
cat(sprintf("\nSum of children : %d h\n", child_sum))
cat(sprintf("Parent estimate : %d h\n", parent_hours))

if (child_sum == parent_hours) {
  cat("OK - the children account for exactly 100% of the parent.\n")
} else if (child_sum < parent_hours) {
  cat(sprintf("GAP - %d h of the parent's scope is not captured by any work package.\n",
              parent_hours - child_sum))
} else {
  cat(sprintf("OVER - the children exceed the parent by %d h (>100%% -- scope crept in).\n",
              child_sum - parent_hours))
}
```

Here the children sum to 340 --- the deliverable is fully decomposed with nothing
missing and nothing extra.

### Mutually exclusive

**No overlap** between WBS elements. Overlap causes duplicated work, confusion
over who is responsible, and messy cost accounting. If element names are
ambiguous, the **WBS dictionary** disambiguates them.

### Plan outcomes, not actions

Define elements as **results** ("Training module completed"), not **actions**
("write script", "record video", "upload"). Action lists always capture too many
or too few steps --- breaking the 100 % rule --- and they over-prescribe *how* the
work is done, killing the team's ingenuity.

## Level of detail

A work package at the **activity level** is a task that:

- can be **realistically and confidently estimated**;
- makes no practical sense to break down further;
- produces a **measurable deliverable**; and
- forms a unique package that **could be outsourced or contracted out**.

**Rules of thumb for how far to decompose:**

- **80-hour rule** --- no work package (single deliverable at the lowest level) exceeds ~80 hours of effort.
- **One reporting period** --- no package is longer than a single progress-reporting period (if you report monthly, nothing runs longer than a month).
- **8-hour floor** --- generally do not define a task below ~8 hours; below that you are micromanaging.
- **"If it makes sense"** --- apply common sense.

### Try it yourself: decomposition check

```{webr-r}
#| label: wbs-decomp-05
packages <- data.frame(
  id    = c("1.1.1", "1.2.1", "1.3.1", "1.3.4", "1.4.2"),
  name  = c("Outer body design", "Motor/controller procurement",
            "Prototype assembly", "Tighten one fastener",
            "Full regulatory approval campaign"),
  hours = c(64, 40, 72, 1.5, 260)
)

flag <- ifelse(packages$hours > 80, "TOO BIG  (> 80 h -- split it)",
        ifelse(packages$hours < 8,  "TOO SMALL (< 8 h -- roll it up)",
               "ok"))
out <- cbind(packages, verdict = flag)
print(out, row.names = FALSE)

cat(sprintf("\n%d of %d packages need re-sizing.\n",
            sum(flag != "ok"), nrow(packages)))
```

## Why the WBS matters

```{r fig-wbs-hub-05, fig.cap="The WBS feeds four downstream planning activities. Get it wrong and the estimate, the schedule, the staffing plan, and the risk register are all built on sand.", fig.align='center', fig.width=9, fig.height=5.5}
hub <- data.frame(x = 0, y = 0)
spokes <- data.frame(
  x = c(0, 1.3, 0, -1.3),
  y = c(1.3, 0, -1.3, 0),
  label = c("Estimating\n(cost & time)", "Network diagram\n(schedule)",
            "Risk\nidentification", "Staffing\n& responsibility")
)
ggplot() +
  geom_segment(data = spokes, aes(x = 0, y = 0, xend = x, yend = y),
               color = "grey55", linewidth = 0.7) +
  geom_point(data = hub, aes(x, y), size = 46, color = "#1B4F72") +
  annotate("text", x = 0, y = 0, label = "WBS", color = "white",
           fontface = "bold", size = 5 * diagram_text_scale) +
  geom_label(data = spokes, aes(x, y, label = label),
             size = 3 * diagram_text_scale, lineheight = 0.9,
             fill = "#D6EAF8", label.size = 0, fontface = "bold") +
  coord_equal(xlim = c(-2.2, 2.2), ylim = c(-2.1, 2.1)) +
  labs(title = "Everything downstream depends on the WBS") +
  theme_void() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale,
                                  face = "bold", hjust = 0.5))
```

## The WBS dictionary

The dictionary is used to **control what work is done and when**, **prevent scope
creep**, and **capture the effort** for each task. Physically, teams post the
work packages on a wall in logical work-flow order; **removing them in numerical
order generates the dictionary**. Each entry --- a **task description** --- records:

- Task description (what work is authorized)
- Quantified objectives and deliverables (cross-referenced to the charter)
- Product description; acceptance criteria (how the owner checks their own work)
- Assumptions; resources assigned; duration; cost; due date
- Interdependencies (before this task / after this task)
- Project manager approval and date

## Work authorization and contracts

Once the scope is decomposed into work packages, stakeholders formally agree on
project content --- the **work authorization**. If the client is external, this
includes the **contract sign-off**.

A valid project-management contract requires all six of:

1. The legal **right** to contract
2. Legal **age** (18 in Ontario and Quebec; 19 or 21 elsewhere in Canada)
3. **Sound mind** at signing
4. Valid, legal **consideration** for both parties
5. A **legal scope of work**
6. **Offer and acceptance**

Contracts also commonly contain: excusable delays; allowable costs; **liquidated
damages** (non-performance / penalty clauses); inspection criteria;
responsibility for correcting defects; and dispute-resolution mechanisms.

### Try it yourself: cost roll-up from the WBS

```{webr-r}
#| label: wbs-cost-rollup-05
# Work-package costs (direct), grouped by Level-2 deliverable.
wp <- data.frame(
  deliverable = c(rep("1.1 Body & Frame", 3),
                  rep("1.2 Drive & Controls", 2),
                  rep("1.3 Assembly & Test", 3),
                  rep("1.4 Launch", 2)),
  package = c("Outer body", "Inner frame", "Material selection",
              "Motor/controller", "Wiring harness",
              "Prototype assembly", "Drop test", "EMC test",
              "Packaging design", "Approvals"),
  cost = c(18000, 14000, 6000, 140000, 9000, 21000, 12000, 15000, 8000, 6000)
)

by_deliverable <- tapply(wp$cost, wp$deliverable, sum)
cat("Roll-up to Level 2:\n")
for (d in names(by_deliverable))
  cat(sprintf("  %-22s $%s\n", d, format(by_deliverable[d], big.mark = " ")))

cat(sprintf("\nLevel 1 (total project direct cost): $%s\n",
            format(sum(wp$cost), big.mark = " ")))
cat("\nThis roll-up only works because the WBS obeys the 100% and\n")
cat("mutually-exclusive rules -- every dollar is counted once, and all of them.\n")
```

## Key Takeaways

- A **WBS** is a **deliverable-oriented** decomposition of the **total scope** into deliverables → work packages → activities. PMs manage the pieces.
- **Numbering** encodes the hierarchy: Level 1 project, Level 2 deliverables, Level 3+ work packages.
- Obey the three design rules: **100 % rule** (children sum to the parent, never more), **mutually exclusive** (no overlap), and **plan outcomes, not actions**.
- Size work packages with the **8/80 rule** and the **one-reporting-period** rule; below ~8 h you are micromanaging, above ~80 h you cannot estimate confidently.
- The WBS feeds **estimating, scheduling, staffing, and risk** --- errors propagate to all four.
- The **WBS dictionary** locks down each task's description, acceptance criteria, resources, and interdependencies, and helps prevent scope creep.
- **Work authorization** formalizes agreement on content; a valid contract needs all **six** legal elements and usually carries **liquidated-damages** and dispute clauses.

## Practice Problems

<details>
<summary>Problem 1: Fix the 100 % rule violation</summary>

The "1.3 Assembly & Test" deliverable is estimated at 300 h. Its packages are
listed as: Prototype assembly (110 h), Drop test (60 h), EMC test (70 h),
Prototype assembly rework (40 h), Write the test report (30 h), *and* "General
project management (50 h)."

What is wrong, and how do you fix it?

**Solution.** The children sum to 360 h --- 60 h **over** 100 % of the parent. Two
problems: "Prototype assembly rework" overlaps "Prototype assembly" (not
mutually exclusive --- fold rework into the assembly package or make it a
contingency), and "General project management" is **not part of this
deliverable's scope** --- PM effort belongs in its own Level-2 element (e.g.
"1.0.1 Project Management"), not smuggled into a technical deliverable.

</details>

<details>
<summary>Problem 2: Outcome or action?</summary>

Rewrite these action-style entries as outcome-style WBS elements:

1. "Call three carbon-fibre suppliers and get quotes"
2. "Run the mould 20 times and measure the parts"
3. "Email the drawings to the customer for approval"

**Model answers.** 1 → "Carbon-fibre supplier selected and quoted." 2 →
"Moulding process capability established." 3 → "Customer design approval
obtained." Each is a result you can mark done or not done, and it does not
dictate the method.

</details>

<details>
<summary>Problem 3 (discussion): how deep should the RC-vehicle WBS go?</summary>

Your draft WBS has a work package "Design the outer body" estimated at 180 h and
spanning 6 weeks, and another, "Deburr the frame jig", estimated at 2 h. Using
the rules of thumb, what do you do with each?

**Discussion points.** "Design the outer body" breaks both the 80-hour rule and
(if you report every 2 weeks) the single-reporting-period rule --- split it into
concept, detailed CAD, and design-release packages. "Deburr the frame jig" is
below the 8-hour floor --- roll it into the jig-fabrication package rather than
tracking it as its own line.

</details>

## Further Resources

```{r video-resources-05, echo=FALSE, results='asis'}
embed_youtube("akO2Lf1fHmM", "What Is a Work Breakdown Structure? (Online PM Courses)")
embed_youtube("_7R_jjxVpK4", "Work Breakdown Structure Basics")
```
