---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Network Diagrams and Activity-on-Node (AON)

```{r setup-06, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Build** an activity-on-node (AON) network from a precedence table.
2. **Apply** the rules for constructing a valid network diagram.
3. **Perform** the forward pass to find early start (ES) and early finish (EF).
4. **Perform** the backward pass to find late start (LS) and late finish (LF).
5. **Calculate** total float and identify the **critical path**.
6. **Compute** a three-point (PERT) activity duration.
7. **Choose** a strategy to shorten the critical path when the project is too long.

<details>
<summary>Before we begin: which task can slip without moving the deadline?</summary>

On any project some tasks have slack and some do not. Guessing wrong --- pouring
effort into a task with float while a zero-float task quietly slips --- is one of
the most common ways schedules die. The network diagram tells you which is which.

</details>

## Introduction: when can we launch?

The RC-vehicle WBS gives you a list of activities and their durations. It does
*not* tell you how long the whole project takes, because the activities are not
all independent --- you cannot assemble a prototype before the body is designed.
A **network diagram** captures those dependencies and answers the question the
sponsor actually cares about: **what is the earliest the project can finish, and
which activities decide that date?**

::: {.callout-note}
**Industry snapshot.** Every automotive program plan is a critical-path network:
the "job one" date is the length of the longest dependent chain. A food-plant
commissioning schedule uses the same math to decide whether sanitation
qualification can overlap controls checkout. A defense contractor reports
critical-path status to the customer monthly --- slipping a critical task is a
contractual event.
:::

```{r video-cpm-06, echo=FALSE, results='asis'}
embed_youtube("4oDLMs11Exs", "Forward and Backward Pass -> Project Duration and Critical Path (Engineer4Free)")
```

## From WBS to precedence table

For each activity you need its **duration** and its **immediate predecessors** ---
the activities that must finish before it can start. Rules for a valid AON
network:

1. Determine precedence ordering **before** drawing.
2. Networks flow **left to right**.
3. An activity cannot begin until **all** connected predecessors are complete.
4. Arrows show **precedence and logical flow** (they may cross, but minimize it).
5. Every activity has a **unique identifier**.
6. **No looping** back to an earlier activity.
7. It is common (not required) to start from a **single start node**...
8. ...and end at a **single end node**.

## AON node anatomy

Each activity is a box with six numbers:

```{r fig-node-anatomy-06, fig.cap="The six-number activity node. Forward pass fills the top row (ES, DUR, EF); backward pass fills the bottom (LS, FLT, LF).", fig.align='center', fig.width=8, fig.height=4.2}
ggplot() +
  annotate("rect", xmin = 0, xmax = 3, ymin = 0, ymax = 2, fill = "#D6EAF8",
           color = "#2E86C1", linewidth = 1) +
  annotate("segment", x = 0, xend = 3, y = 1, yend = 1, color = "#2E86C1") +
  annotate("segment", x = 1, xend = 1, y = 0, yend = 2, color = "#2E86C1") +
  annotate("segment", x = 2, xend = 2, y = 0, yend = 2, color = "#2E86C1") +
  annotate("text", x = c(0.5, 1.5, 2.5), y = 1.5,
           label = c("ES", "DUR", "EF"), fontface = "bold",
           size = 4 * diagram_text_scale) +
  annotate("text", x = c(0.5, 1.5, 2.5), y = 0.5,
           label = c("LS", "FLT", "LF"), fontface = "bold",
           size = 4 * diagram_text_scale) +
  annotate("text", x = 4.6, y = 1.5,
           label = "ES = earliest start\nDUR = duration\nEF = earliest finish",
           hjust = 0, size = 3 * diagram_text_scale, lineheight = 0.95) +
  annotate("text", x = 4.6, y = 0.5,
           label = "LS = latest start\nFLT = float (slack)\nLF = latest finish",
           hjust = 0, size = 3 * diagram_text_scale, lineheight = 0.95) +
  coord_equal(xlim = c(-0.3, 8.5), ylim = c(-0.4, 2.4)) +
  theme_void()
```

An activity with 2+ predecessors flowing **in** is a **merge**; one with 2+
successors flowing **out** is a **burst**.

## Worked Example: the RC-vehicle sub-project network

Ten activities (a simplified slice of the project):

```{r cpm-setup-06}
acts <- data.frame(
  id = LETTERS[1:10],
  dur = c(5, 8, 3, 4, 5, 7, 6, 8, 6, 5),
  pred = c("", "A", "B", "A", "C", "C", "D,E", "F", "G", "H,I"),
  stringsAsFactors = FALSE
)
knitr::kable(acts, col.names = c("Activity", "Duration (days)", "Predecessors"),
             align = c("c", "c", "l")) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                            full_width = FALSE)
```

### Forward pass (ES, EF)

Move left to right. $EF = ES + DUR$. Carry each EF forward as the successor's
ES --- **except at a merge, where the ES is the largest incoming EF.**

### Backward pass (LS, LF)

Move right to left from the project finish. $LS = LF - DUR$. Carry each LS back
as the predecessor's LF --- **except at a burst, where the LF is the smallest
outgoing LS.**

### Float and the critical path

$$FLT = LS - ES = LF - EF$$

Activities with **zero float** cannot slip without delaying the whole project ---
together they form the **critical path**.

Live cell --- a base-R critical-path solver. Run it, then change a duration or a
predecessor in `acts` and watch the project duration and the critical path move.

```{webr-r}
#| label: cpm-solve-06
# The ten activities: id, duration, and comma-separated predecessors.
acts <- data.frame(
  id = LETTERS[1:10],
  dur = c(5, 8, 3, 4, 5, 7, 6, 8, 6, 5),
  pred = c("", "A", "B", "A", "C", "C", "D,E", "F", "G", "H,I"),
  stringsAsFactors = FALSE
)

# A small critical-path solver (base R only).
cpm <- function(df) {
  df$pred <- strsplit(df$pred, ",")
  n <- nrow(df); id <- df$id
  ES <- EF <- LS <- LF <- setNames(rep(NA_real_, n), id)

  # Forward pass (ids are already in a valid topological order here)
  for (i in seq_len(n)) {
    p <- df$pred[[i]]; p <- p[p != ""]
    ES[i] <- if (length(p) == 0) 0 else max(EF[p])
    EF[i] <- ES[i] + df$dur[i]
  }
  proj <- max(EF)

  # Backward pass
  for (i in rev(seq_len(n))) {
    succ <- id[sapply(df$pred, function(pp) id[i] %in% pp)]
    LF[i] <- if (length(succ) == 0) proj else min(LS[succ])
    LS[i] <- LF[i] - df$dur[i]
  }

  data.frame(Activity = id, Dur = df$dur,
             ES = ES, EF = EF, LS = LS, LF = LF,
             Float = LS - ES,
             Critical = ifelse(LS - ES == 0, "  <- critical", ""))
}

res <- cpm(acts)
print(res, row.names = FALSE)
cat(sprintf("\nProject duration: %d days\n", max(res$EF)))
cat("Critical path:", paste(res$Activity[res$Float == 0], collapse = " - "), "\n")
```

```{r cpm-knitr-06, include=FALSE}
# Same solver, run in the page-build session so the float figure below and the
# Practice Problem 1 solution can reuse cpm() and res. (webR cells run only in
# the reader's browser and can't feed later figures.)
cpm <- function(df) {
  df$pred <- strsplit(df$pred, ",")
  n <- nrow(df); id <- df$id
  ES <- EF <- LS <- LF <- setNames(rep(NA_real_, n), id)
  for (i in seq_len(n)) {
    p <- df$pred[[i]]; p <- p[p != ""]
    ES[i] <- if (length(p) == 0) 0 else max(EF[p])
    EF[i] <- ES[i] + df$dur[i]
  }
  proj <- max(EF)
  for (i in rev(seq_len(n))) {
    succ <- id[sapply(df$pred, function(pp) id[i] %in% pp)]
    LF[i] <- if (length(succ) == 0) proj else min(LS[succ])
    LS[i] <- LF[i] - df$dur[i]
  }
  data.frame(Activity = id, Dur = df$dur, ES = ES, EF = EF, LS = LS, LF = LF,
             Float = LS - ES,
             Critical = ifelse(LS - ES == 0, "  <- critical", ""))
}
res <- cpm(acts)
```

```{r fig-cpm-float-06, echo=FALSE, fig.cap="Total float per activity. The zero-float bars (A, B, C, E, G, I, J) are the critical path; D has 12 days of slack, F and H have 2.", fig.align='center', fig.width=10, fig.height=4.5}
res$onCP <- res$Float == 0
ggplot(res, aes(x = Activity, y = Float, fill = onCP)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Float), vjust = -0.4, size = 3.6 * diagram_text_scale) +
  scale_fill_manual(values = c("TRUE" = "#C0392B", "FALSE" = "#5DADE2"),
                    labels = c("TRUE" = "Critical (float = 0)", "FALSE" = "Has float"),
                    name = NULL) +
  labs(title = "Where the slack is", x = "Activity", y = "Total float (days)") +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

The whole forward/backward/float sequence is one connected calculation --- change
any duration and every downstream number moves --- so work it as a block, not as
isolated steps.

## Three-point (PERT) duration estimates

When a duration is uncertain, estimate three values --- **optimistic** $o$,
**most likely** $m$, **pessimistic** $p$ --- and combine them:

$$t_e = \frac{o + 4m + p}{6}$$

### Try it yourself: PERT durations

```{webr-r}
#| label: pert-te-06
tasks <- data.frame(
  id = c("A", "B", "C", "D", "E"),
  o  = c(12, 7, 15, 9, 24),
  m  = c(15, 10, 19, 13, 26),
  p  = c(17, 13, 23, 16, 30)
)
tasks$te <- round((tasks$o + 4 * tasks$m + tasks$p) / 6, 2)
tasks$sd <- round((tasks$p - tasks$o) / 6, 2)   # activity standard deviation

print(tasks, row.names = FALSE)
cat(sprintf("\nSum of expected durations along this chain: %.2f days\n", sum(tasks$te)))
```

<details>
<summary>Why the weight of 4 on the most-likely value?</summary>

PERT assumes each activity duration follows a Beta distribution skewed toward
the most-likely value, with the optimistic and pessimistic values as the tails.
Weighting $m$ by 4 and $o$, $p$ by 1 each (dividing by 6) is the standard
approximation of that distribution's mean. The activity standard deviation
$(p - o)/6$ becomes important in [Chapter 8](08-Project-Scheduling.qmd), where
you combine them to find the probability of finishing on time.

</details>

## Shortening the critical path

Only work on the **critical path** shortens the project. Options, roughly in
order of preference:

1. **Eliminate** critical tasks (is it truly needed?).
2. **Re-plan series paths to run in parallel.**
3. **Overlap** sequential tasks (start the successor before the predecessor fully finishes --- *fast-tracking*).
4. **Shorten durations on critical activities** --- prefer earlier-starting tasks, then the longest, then the easiest, then the ones **cheapest to speed up**.

### Try it yourself: recompute the critical path after a change

```{webr-r}
#| label: path-duration-06
# All start-to-finish paths through the sub-project network, with activity durations.
dur <- c(A = 5, B = 8, C = 3, D = 4, E = 5, F = 7, G = 6, H = 8, I = 6, J = 5)

paths <- list(
  "A-B-C-E-G-I-J" = c("A","B","C","E","G","I","J"),
  "A-B-C-F-H-J"   = c("A","B","C","F","H","J"),
  "A-D-G-I-J"     = c("A","D","G","I","J")
)

path_len <- sapply(paths, function(p) sum(dur[p]))
cat("Baseline path lengths:\n"); print(path_len)
cat(sprintf("Critical path: %s (%d days)\n\n", names(which.max(path_len)), max(path_len)))

# Now crash B from 8 to 5 days and re-check:
dur["B"] <- 5
path_len2 <- sapply(paths, function(p) sum(dur[p]))
cat("After crashing B by 3 days:\n"); print(path_len2)
cat(sprintf("New critical path: %s (%d days)\n", names(which.max(path_len2)), max(path_len2)))
cat("Note how the critical path can *shift* to a different chain once you crash enough.\n")
```

## The downside of network diagrams

- Often **not drawn to scale** --- time is just a number, which can create a false sense of how much time is available.
- Can be **too large and complex** to be useful.
- **Faulty construction logic** can oversimplify or misrepresent the work.
- Sometimes used for work they do not suit.
- **Positive bias** creeps into PERT estimates (people are optimistic).

## Key Takeaways

- Build the **AON network** from a precedence table (durations + immediate predecessors); it flows left to right with no loops.
- **Forward pass**: $EF = ES + DUR$; at a **merge**, ES = the largest incoming EF. Gives the **project duration**.
- **Backward pass**: $LS = LF - DUR$; at a **burst**, LF = the smallest outgoing LS.
- **Float** $= LS - ES = LF - EF$. **Zero-float** activities form the **critical path** --- the longest dependent chain, and the only one worth shortening.
- **PERT**: $t_e = (o + 4m + p)/6$, activity SD $= (p - o)/6$.
- Shorten the critical path by **eliminating, parallelizing, overlapping, or crashing** critical tasks --- and re-check, because the critical path can shift.

## Practice Problems

<details>
<summary>Problem 1: Solve this network</summary>

| Activity | Predecessors | Duration |
|---|---|---|
| A | none | 3 |
| B | A | 4 |
| C | A | 2 |
| D | B | 5 |
| E | C | 1 |
| F | E | 2 |
| G | D, E | 4 |
| H | F, G | 3 |
| I | H | 6 |

Find the project duration and the critical path.

**Solution.**

```{r p1-cpm-06, echo=TRUE}
p1 <- data.frame(
  id = LETTERS[1:9],
  dur = c(3, 4, 2, 5, 1, 2, 4, 3, 6),
  pred = c("", "A", "A", "B", "C", "E", "D,E", "F,G", "H"),
  stringsAsFactors = FALSE
)
r1 <- cpm(p1)          # reuse the solver defined earlier in this chapter
print(r1, row.names = FALSE)
cat(sprintf("\nProject duration: %d days\n", max(r1$EF)))
cat("Critical path:", paste(r1$Activity[r1$Float == 0], collapse = " - "), "\n")
```

Duration 25 days; critical path A--B--D--G--H--I. (The chain A--B--D--G is what
pushes E, F, and C off the critical path --- D's 5 days arriving at the G merge
beat E's early finish by 6 days.)

</details>

<details>
<summary>Problem 2: PERT and the optimistic tail</summary>

An activity is estimated at $o = 6$, $m = 9$, $p = 24$ days.

(a) What is $t_e$? (b) Why is $t_e$ noticeably larger than $m$? (c) What is the activity standard deviation?

**Solution.** (a) $t_e = (6 + 36 + 24)/6 = 11$ days. (b) The pessimistic tail is
far from the most-likely value (24 vs 9) while the optimistic tail is close
(6 vs 9), so the distribution is **right-skewed** and its mean sits above $m$.
(c) SD $= (24 - 6)/6 = 3$ days --- a wide, risky activity.

</details>

<details>
<summary>Problem 3 (discussion): which task do you crash?</summary>

On the sub-project network, the critical path is A--B--C--E--G--I--J (38 days)
and the sponsor needs 34. Crash data: B costs \$500/day to shorten (max 3 days),
G costs \$1 200/day (max 2 days), E costs \$300/day (max 2 days). Which
combination gets you to 34 days at least cost, and what do you check afterward?

**Discussion points.** You need 4 days. Cheapest first: E for 2 days
(\$600), then B for 2 days (\$1 000) --- total \$1 600 for 4 days. Avoid G at
\$1 200/day. **Then re-run the network**: after removing 4 days from this chain,
another path (e.g. A--B--C--F--H--J) may now be critical or co-critical, and
crashing further would have to shorten *both*.

</details>

## Further Resources

```{r video-resources-06, echo=FALSE, results='asis'}
embed_youtube("AQ_Wr1Me__8", "PERT / Three-Point Estimating Explained (Cornelius Fichtner)")
embed_youtube("u-9o-qa3HDg", "Three-Point Estimates and PERT (PMC Lounge)")
```
