---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Risk Management

```{r setup-07, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Distinguish** risk from uncertainty, and threats from opportunities.
2. **Classify** a risk and write a well-formed risk statement.
3. **Run** a qualitative analysis with a probability--impact matrix.
4. **Compute** expected monetary value (EMV) and solve a simple decision tree.
5. **Select** a response strategy: accept, minimize, share, transfer, or reserve.
6. **Size** a contingency reserve and distinguish task contingency from managerial contingency.
7. **Explain** how FMEA supports project risk management, complete a simplified FMEA form, and compute a Risk Priority Number (RPN) from standardized Severity/Occurrence/Detection scales.

<details>
<summary>Before we begin: name one thing that could sink the RC-vehicle project</summary>

Now: how likely is it, and how bad would it be? You just did qualitative risk
analysis. The rest of this chapter makes it systematic.

</details>

## Introduction: the known unknowns

The RC-vehicle brief is a risk register waiting to be written. One candidate
motor supplier is **unproven** and its staff are **negotiating a first labour
contract**. The carbon-fibre supplier is **mid-restructuring** after a foreign
acquisition. Choosing a **new assembly line** brings a **learning curve**. None
of these is a certainty --- but each is *identifiable*, and that is exactly what
makes it a **risk** rather than an uncertainty.

> **Risk** = the probability or threat of loss, injury, or any negative
> occurrence caused by internal or external vulnerabilities, that **can be
> reduced by preemptive action**.
>
> An **uncertainty** is when *neither* the probability *nor* the mode of
> occurrence is known. If you do not know about it, you cannot act on it.

The core equation:

$$\text{Risk} = \text{Probability of event} \times \text{Impact of event}$$

Probability is usually a percentage; impact is dollars or schedule delay. **Risk
and opportunity are mirror images** --- an opportunity is an unforeseen event with
a *positive* outcome, handled with the same tools for the opposite end.

::: {.callout-note}
**Industry snapshot.** Aircraft makers such as Bombardier **minimize** part risk
by forcing every significant supplier into direct contact with their quality
department and auditing their processes. The European Space Agency and Airbus
**share** the risk of a new airframe across nations because no one country can
carry it. A fixed-price defense contract **transfers** budget-overrun risk to
the vendor.
:::

```{r video-emv-07, echo=FALSE, results='asis'}
embed_youtube("dD-Vygp_MiM", "Expected Monetary Value (EMV) Explained Simply (Online PM Courses)")
```

## The four-stage process

1. **Risk identification** --- determine the specific risk factors that can reasonably be expected to affect the project.
2. **Analysis of probability and consequences** --- how likely, and how bad.
3. **Risk mitigation strategies** --- steps to reduce the impact of the risks judged threatening enough to act on.
4. **Control and documentation** --- track the risks and build a **knowledge base** for future projects from the lessons learned.

## Risk identification and classification

Sort each identified risk into a class. A common (not exhaustive) list:

- **Financial** --- material price moves, currency, budget cuts
- **Technology** --- unproven design or process
- **Resources** --- key people, equipment availability
- **Schedule** --- dependency slips, long lead times
- **Weather**, **Political**, **Environmental**

Techniques: structured **brainstorming**, **interviewing** to get optimistic /
most-likely / pessimistic views of an event, **probability distributions** to
represent uncertain outcomes, and **expert judgment** to validate.

## Qualitative analysis: the probability--impact matrix

Plot each risk by **likelihood** (x-axis) and **impact** (y-axis). The
top-right is **red** (mitigation planned in advance), the middle band
**yellow** (team decides whether to plan), the bottom-left **green** (just stay
aware).

```{r fig-pi-matrix-07, fig.cap="Probability-impact matrix for the RC-vehicle project. Red-zone risks (unproven supplier failure, carbon-fibre disruption) get a written mitigation plan before anything goes wrong.", fig.align='center', fig.width=9.5, fig.height=6.5}
grid <- expand.grid(prob = 1:5, impact = 1:5)
grid$score <- grid$prob * grid$impact
grid$zone <- cut(grid$score, c(0, 6, 12, 25),
                 labels = c("Green", "Yellow", "Red"))

risks <- data.frame(
  risk = c("Unproven supplier can't deliver", "Carbon-fibre plant disruption",
           "New line learning curve overruns", "Labour action at supplier",
           "Steel price spike", "Minor drawing errors"),
  prob = c(3, 3, 4, 2, 3, 4),
  impact = c(5, 4, 3, 5, 2, 1),
  label_nudge = c(0.65, 0, 0, -0.65, 0, 0)
)

ggplot(grid, aes(prob, impact)) +
  geom_tile(aes(fill = zone), color = "white", linewidth = 1) +
  scale_fill_manual(values = c(Green = "#A9DFBF", Yellow = "#F9E79F", Red = "#F5B7B1"),
                    name = "Zone") +
  geom_point(data = risks, aes(prob, impact), inherit.aes = FALSE,
             size = 3, color = "#1B4F72") +
  geom_text(data = risks, aes(prob + label_nudge, impact, label = risk), inherit.aes = FALSE,
            size = 2.7 * diagram_text_scale, vjust = -1, hjust = 0.5) +
  scale_x_continuous(breaks = 1:5, labels = c("Very low", "Low", "Medium", "High", "Very high")) +
  scale_y_continuous(breaks = 1:5, labels = c("Negligible", "Minor", "Moderate", "Major", "Severe")) +
  labs(title = "Probability x Impact", x = "Probability", y = "Impact") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 9.5 * diagram_text_scale)
  )
```

## Quantitative analysis

Tools: **sensitivity analysis** (which risks move the outcome most),
**expected monetary value** (the probability-weighted average outcome),
**decision trees** (lay out choices and chance events), and **Monte Carlo**
simulation (run the network thousands of times with sampled durations/costs).

### Worked Example 1: EMV of a single risk

The "new assembly line learning curve" risk: a **40 %** chance it adds
**\$120 000** of rework and overtime; otherwise \$0.

Live cell --- change the probability or the impact of the risk.

```{r calc-emv-example-07, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "emv-example-07",
  title = "Expected Monetary Value Calculator",
  icon = "dice",
  inputs = list(
    qwebr_calc_input("p_occurs", "Probability the risk occurs", value = 0.4, min = 0, max = 1, step = 0.05, type = "slider"),
    qwebr_calc_input("impact", "Impact if it occurs", value = 120000, min = 0, max = 1000000, step = 5000, prefix = "$")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: emv-example-07
p_occurs <- 0.40
impact   <- 120000

emv <- p_occurs * impact + (1 - p_occurs) * 0

money <- function(x) paste0(ifelse(x < 0, "-", ""), "$", format(round(abs(x)), big.mark = " ", scientific = FALSE, trim = TRUE))

probs <- seq(0, 1, length.out = 100)
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
plot(probs * 100, probs * impact / 1000, type = "l", lwd = 4, col = "#2c3e50", xlab = "Probability of occurring (%)", ylab = "EMV ($ thousand)",
     main = "EMV grows with probability")
abline(v = p_occurs * 100, lty = 2, col = "#bdc3c7")
points(p_occurs * 100, emv / 1000, pch = 19, cex = 2.2, col = "#e74c3c")

cat(sprintf("EMV of this risk: %s\n", money(emv)))
cat(sprintf("Calculation  %.2f x %s\n", p_occurs, money(impact)))
```

</details>

### Try it yourself: decision tree --- buy the copper now or later?

You need **750 lb of copper** for the RC-vehicle wiring harness tooling. Buy it
**now** and pay **\$100** storage, or buy it **later** at the summer price ---
which is **\$5.58/lb with 90 % probability**, or **\$4.50/lb with 10 %
probability**.

```{r calc-decision-tree-07, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "decision-tree-07",
  title = "Decision-Tree EMV Calculator",
  icon = "code-branch",
  inputs = list(
    qwebr_calc_input("lb", "Quantity needed", value = 750, min = 100, max = 3000, step = 50, suffix = "lb"),
    qwebr_calc_input("storage", "Storage cost if bought now", value = 100, min = 0, max = 1000, step = 10, prefix = "$"),
    qwebr_calc_input("price_now", "Price if bought now", value = 5.58, min = 1, max = 10, step = 0.01, prefix = "$", suffix = "/ lb"),
    qwebr_calc_input("p_high", "Chance price stays high", value = 0.9, min = 0, max = 1, step = 0.05, type = "slider"),
    qwebr_calc_input("price_high", "Later price if it stays high", value = 5.58, min = 1, max = 10, step = 0.01, prefix = "$", suffix = "/ lb"),
    qwebr_calc_input("price_low", "Later price if it drops", value = 4.5, min = 1, max = 10, step = 0.01, prefix = "$", suffix = "/ lb")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: decision-tree-07
lb <- 750
storage <- 100
price_now <- 5.58

# Branch: buy now (price locked, pay storage)
buy_now <- price_now * lb + storage

# Branch: buy later (two chance outcomes)
p_high     <- 0.90
price_high <- 5.58
price_low  <- 4.50
p_low      <- 1 - p_high
fmt2 <- function(x) formatC(x, format = "f", digits = 2, big.mark = " ")
buy_later <- p_high * (price_high * lb) + p_low * (price_low * lb)

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
emvs <- c("Buy now" = buy_now, "Buy later" = buy_later)
bp <- barplot(emvs, col = c("#2980b9", "#e67e22"), ylim = c(0, max(emvs) * 1.15),
              ylab = "Expected cost ($)", main = "EMV of each branch")
text(bp, emvs, labels = paste0("$", format(round(emvs), big.mark = " ")), pos = 3, cex = 1.3)

cat(sprintf("Choose: %s (saves $%s on average)\n",
            ifelse(buy_later < buy_now, "BUY LATER", "BUY NOW"), fmt2(abs(buy_now - buy_later))))
cat(sprintf("EMV, buy now  $%s\n", fmt2(buy_now)))
cat(sprintf("EMV, buy later  $%s\n", fmt2(buy_later)))
cat(sprintf("Price if high (%.0f%%)  $%s\n", 100 * p_high, fmt2(price_high * lb)))
cat(sprintf("Price if low (%.0f%%)  $%s\n", 100 * p_low, fmt2(price_low * lb)))
```

</details>

<details>
<summary>What would flip the decision?</summary>

Raise the probability of the low summer price, widen the price gap, or cut the
storage cost toward zero and "buy later" wins by more. Push storage cost up
(a bonded warehouse, insurance, spoilage) or make the high price nearly
certain and "buy now" wins. A decision tree makes the tipping point explicit
instead of leaving it to gut feel.

</details>

## Response strategies

| Strategy | What it means | Example on our project |
|---|---|---|
| **Accept** | The risk is minor or unlikely enough to live with | Small drawing errors --- caught in review, low cost |
| **Minimize** | Reduce probability or impact | Audit the unproven supplier; put their staff in contact with your quality team |
| **Share** | Split the risk (and reward) with a partner | Co-develop the new control system with the supplier under a shared-savings deal |
| **Transfer** | Shift the risk to another party | Fixed-price contract (vendor eats overruns); **liquidated damages** for non-performance |
| **Contingency reserve** | Set aside funds/time for the risk if it hits | 10--15 % reserve against the carbon-fibre and learning-curve risks |
| **Other** | Anything else fit for the situation | Dual-source the motor; pre-qualify a backup |

For **opportunities**, the mirror strategies are **exploit, share, enhance,** and
**accept**.

## Contingency reserves

Two kinds:

- **Task (project) contingency** --- offsets overruns on individual tasks/work packages. **Managed by the project manager.** As work completes, reserve for finished tasks is released and cannot be reused.
- **Managerial contingency** --- a project-level buffer for scope changes forced by the customer, "acts of God," and other unforeseeable events. **Managed by senior management, not the PM.**

In construction it is common to set aside **10--15 %** of the price. Other
industries resist it --- the customer may read a visible reserve as a sign of
poor planning.

### Try it yourself: does the mitigation pay for itself?

```{r calc-mitigation-value-07, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "mitigation-value-07",
  title = "Does the Mitigation Pay for Itself?",
  icon = "shield-halved",
  inputs = list(
    qwebr_calc_input("p_fail_no_action", "Failure chance, do nothing", value = 0.35, min = 0, max = 1, step = 0.05),
    qwebr_calc_input("impact_if_fail", "Impact if it fails", value = 260000, min = 0, max = 1000000, step = 10000, prefix = "$"),
    qwebr_calc_input("mitigation_cost", "Mitigation cost", value = 22000, min = 0, max = 200000, step = 1000, prefix = "$"),
    qwebr_calc_input("p_fail_with_action", "Failure chance, with mitigation", value = 0.1, min = 0, max = 1, step = 0.05, type = "slider"),
    qwebr_calc_input("impact_with_action", "Impact if it still fails", value = 90000, min = 0, max = 1000000, step = 10000, prefix = "$")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: mitigation-value-07
# Risk: unproven supplier fails to deliver on time.
p_fail_no_action <- 0.35
impact_if_fail   <- 260000       # expedite alternative source + line downtime

# Mitigation: pre-qualify a backup supplier now.
mitigation_cost  <- 22000
p_fail_with_action <- 0.10       # backup ready -> failure much less damaging
impact_with_action <- 90000      # switch to backup, smaller hit

emv_before <- p_fail_no_action * impact_if_fail
emv_after  <- mitigation_cost + p_fail_with_action * impact_with_action
saving     <- emv_before - emv_after

money <- function(x) paste0(ifelse(x < 0, "-", ""), "$", format(round(abs(x)), big.mark = " ", scientific = FALSE, trim = TRUE))

stack <- cbind("Do nothing" = c(0, emv_before),
               "Pre-qualify backup" = c(mitigation_cost, emv_after - mitigation_cost))
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
barplot(stack / 1000, col = c("#7f8c8d", "#e67e22"), ylab = "Expected cost ($ thousand)",
        main = "Expected cost: spent now (grey) + loss (orange)")

if (saving >= 0) {
  cat(sprintf("Mitigation pays: expected saving %s\n", money(saving)))
} else {
  cat(sprintf("!!Mitigation does not pay: expected loss of %s!!\n", money(-saving)))
}
cat(sprintf("EMV, do nothing  %s\n", money(emv_before)))
cat(sprintf("EMV, pre-qualify backup  %s (incl. %s spent now)\n", money(emv_after), money(mitigation_cost)))
```

</details>

## FMEA as a risk tool

**Failure Mode and Effects Analysis** is often the first step of a reliability
study: review as many components, assemblies, and subsystems as possible to
identify **failure modes** (what can fail and how), their **causes**, and their
**effects** (the consequences). It is a structured, bottom-up way to surface
technical risks before they reach the customer --- and it feeds the same
probability--impact thinking used above. Instead of a single probability and a
single dollar impact, FMEA scores each failure mode on three separate 1--10
scales --- **Severity**, **Occurrence**, **Detection** --- and multiplies them
into a **Risk Priority Number (RPN)**.

### The FMEA form

Every failure mode gets one row on a standard form: what the step is supposed
to do, how it can fail, the effect of that failure, a severity rating, the
likely cause, an occurrence rating, what currently catches it, a detection
rating, the resulting RPN, and the action taken if the RPN is too high.

```{r fmea-form-07, echo=FALSE}
fmea_form <- data.frame(
  Field = c("Process Step", "Function", "Failure Mode", "Effect", "S",
            "Cause", "O", "Current Controls", "D", "RPN", "Action"),
  Example = c("Install motor controller",
              "Deliver throttle signal reliably at all speeds",
              "Controller intermittently drops signal",
              "Vehicle loses power mid-use; return; safety concern at speed",
              "8",
              "Unqualified solder joints from new (Paragon) supplier",
              "6",
              "Final functional test on assembly bench",
              "5",
              "8 x 6 x 5 = 240",
              "Require a qualification sample and burn-in test before contract (see Week 7 case study, R1)")
)
knitr::kable(fmea_form, align = c("l", "l"),
             caption = "FMEA form, one row, worked from the motor-controller risk.") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(1, bold = TRUE)
```

### Rating Severity, Occurrence, and Detection

Three people scoring the same failure mode should land on close to the same
numbers, so each factor is rated against a **predefined table**, not a gut
feel.

- **Severity (S)** --- how bad is the *effect* if the failure happens? Rated
  on the effect, not the cause --- so severity can really only be brought down
  by changing the design, not the process.
- **Occurrence (O)** --- how likely is the *cause* to happen?
- **Detection (D)** --- how likely are *current controls* to catch the failure
  before it reaches the customer? A 10 means no control exists at all; a 1
  means the process is error-proofed so the defect physically cannot get out.

```{r severity-scale-07, echo=FALSE}
severity <- data.frame(
  Rating = 10:1,
  Criteria = c(
    "Hazardous, no warning -- safety/regulatory failure with no warning",
    "Hazardous, with warning -- safety/regulatory failure, warning given",
    "Very high -- product/service inoperable, loss of primary function",
    "High -- operable, but at reduced performance",
    "Moderate -- operable, degradation noticeable to most customers",
    "Low -- operable, noticed by the average customer",
    "Very low -- operable, noticed only by a discriminating customer",
    "Minor -- nonconforming fit/finish, noticed by a discriminating customer",
    "Very minor -- noticed only by a very observant customer",
    "None -- no discernible effect"
  )
)
knitr::kable(severity, align = c("c", "l"),
             caption = "Severity rating scale (S).") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(1, bold = TRUE)
```

```{r occurrence-scale-07, echo=FALSE}
occurrence <- data.frame(
  Rating = 10:1,
  Criteria = c(
    "Very high -- failure almost inevitable",
    "Very high -- failures occur very frequently",
    "High -- failures occur frequently",
    "High -- failures occur often, on a regular basis",
    "Moderate -- occasional failures",
    "Moderate -- a moderate number of failures",
    "Low -- relatively few failures",
    "Low -- few failures",
    "Remote -- failure unlikely",
    "Remote -- failure nearly impossible"
  )
)
knitr::kable(occurrence, align = c("c", "l"),
             caption = "Occurrence rating scale (O).") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(1, bold = TRUE)
```

```{r detection-scale-07, echo=FALSE}
detection <- data.frame(
  Rating = 10:1,
  Criteria = c(
    "Absolute uncertainty -- no current control exists",
    "Very remote -- control will almost certainly not detect",
    "Remote -- control has a poor chance of detection",
    "Very low -- control has a low chance of detection",
    "Low -- control may detect",
    "Moderate -- control has a moderate chance of detection",
    "Moderately high -- control has a reasonably good chance of detection",
    "High -- control has a good chance of detection",
    "Very high -- control will almost certainly detect",
    "Almost certain -- error-proofed; the defect cannot reach the customer"
  )
)
knitr::kable(detection, align = c("c", "l"),
             caption = "Detection rating scale (D).") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(1, bold = TRUE)
```

### Calculating the Risk Priority Number

$$\text{RPN} = \text{Severity (S)} \times \text{Occurrence (O)} \times \text{Detection (D)}$$

RPN ranges from 1 (S=O=D=1) to 1000 (S=O=D=10). Rank failure modes by RPN and
work down the list --- but don't switch off judgment: a severity of 9 or 10
(a safety or regulatory hazard) usually earns action regardless of where the
RPN lands, because two very different S--O--D combinations can multiply out
to the same number.

Live cell --- change S, O, or D for any failure mode and watch the ranking move.

```{r calc-rpn-fmea-07, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "rpn-fmea-07",
  title = "FMEA Risk Priority Number Calculator",
  icon = "triangle-exclamation",
  inputs = list(
    qwebr_calc_input("S1", "Controller signal drop: Severity (S)", value = 8, min = 1, max = 10, step = 1),
    qwebr_calc_input("O1", "Controller signal drop: Occurrence (O)", value = 6, min = 1, max = 10, step = 1),
    qwebr_calc_input("D1", "Controller signal drop: Detection (D)", value = 5, min = 1, max = 10, step = 1),
    qwebr_calc_input("S2", "Panel delamination: Severity (S)", value = 9, min = 1, max = 10, step = 1),
    qwebr_calc_input("O2", "Panel delamination: Occurrence (O)", value = 3, min = 1, max = 10, step = 1),
    qwebr_calc_input("D2", "Panel delamination: Detection (D)", value = 6, min = 1, max = 10, step = 1),
    qwebr_calc_input("S3", "Paint overspray: Severity (S)", value = 2, min = 1, max = 10, step = 1),
    qwebr_calc_input("O3", "Paint overspray: Occurrence (O)", value = 5, min = 1, max = 10, step = 1),
    qwebr_calc_input("D3", "Paint overspray: Detection (D)", value = 3, min = 1, max = 10, step = 1)
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: rpn-fmea-07
# Three failure modes from the RC-vehicle build

# 1. Motor controller drops signal (unproven Paragon supplier)
S1 <- 8
O1 <- 6
D1 <- 5
RPN1 <- S1 * O1 * D1

# 2. Carbon-fibre panel delaminates
S2 <- 9
O2 <- 3
D2 <- 6
RPN2 <- S2 * O2 * D2

# 3. Minor paint overspray on trim
S3 <- 2
O3 <- 5
D3 <- 3
RPN3 <- S3 * O3 * D3

modes <- c("Controller signal drop", "Panel delamination", "Paint overspray")
short <- c("Controller", "Panel", "Overspray")
sev   <- c(S1, S2, S3)
rpns  <- c(RPN1, RPN2, RPN3)
ord   <- order(rpns, decreasing = TRUE)

# RPN is a sort, not a verdict: flag any Severity >= 8 for action regardless of rank.
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
barplot(rpns[ord], names.arg = short[ord], col = ifelse(sev[ord] >= 8, "#e74c3c", "#2980b9"),
        ylab = "RPN (S x O x D)", main = "Priority order by RPN")

cat(sprintf("Top priority: %s (RPN = %d)\n", modes[ord[1]], rpns[ord[1]]))
for (i in ord) {
  flag <- if (sev[i] >= 8) " - !!S >= 8: flag for action!!" else ""
  cat(sprintf("%s  RPN %d (S=%d, O=%d, D=%d)%s\n", modes[i], rpns[i], sev[i],
              c(O1, O2, O3)[i], c(D1, D2, D3)[i], flag))
}
```

</details>

<details>
<summary>Why doesn't the highest severity automatically win?</summary>

The panel delamination (S=9) has a *lower* RPN than the controller signal drop
(S=8) here, because the controller's occurrence and detection are both worse.
RPN is a useful sort, not a verdict: most teams still flag any S >= 8 for
action even if its RPN is not the top of the list, the same way the
probability--impact matrix earlier in this chapter still red-flags a
low-probability event if its impact is severe enough.

</details>

## Key Takeaways

- A **risk** is identifiable and can be acted on preemptively; an **uncertainty** cannot. **Risk = Probability × Impact.** Opportunities are the positive mirror.
- Four stages: **identify → analyze → mitigate → control & document** (build the knowledge base).
- **Qualitative**: plot risks on a **probability--impact matrix**; red-zone risks get a written plan *before* the event.
- **Quantitative**: **EMV** = probability × impact; **decision trees** compare choices with chance events; Monte Carlo for whole-network risk.
- Responses: **accept, minimize, share, transfer, reserve.** Pick per risk; ring-fence reserve against the biggest ones.
- **Task contingency** is the PM's; **managerial contingency** is management's, for the truly unforeseeable.
- **FMEA** systematically surfaces technical failure modes, causes, and effects, scoring each on **Severity, Occurrence, Detection** (predefined 1--10 tables) and multiplying them into an **RPN** to prioritize action.

## Case Study: Meridian R/C — Week 7

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 7: The Motor-Supplier Risk Register"}

You call a formal risk workshop this week, before a single dollar is committed to a motor-controller contract. Priya Nandakumar, Marcus Webb, and Sofia Reyes join you around the whiteboard; Daniel Cho, mindful of the \$800,000 charter ceiling he authorized back in Week 3, has asked for a dollar-costed register, not a list of worries. Kestrel Retail Group's 100,000-unit holiday order is still contingent on the Trailblazer prototype clearing durability testing and toy-safety certification, so anything that could derail either milestone earns a line on the register.

The motor-and-controller sourcing question is still wide open --- that decision does not land until Week 10 --- and it dominates the room. MCS, the trusted incumbent in Surrey, BC, is mid-move to a Mexican production line. Paragon, in Kingston, ON, is cheaper but unproven, and its staff are still negotiating their first labour contract. Marcus argues the team should keep an open mind toward Paragon's lower sticker price; Sofia counters that an unqualified controller is exactly the kind of risk that shows up in a durability-test failure report. Rather than settle the argument today, you convert it into a risk. Alongside it you register the MCS relocation, the carbon-fibre trim supplier's ongoing restructuring, and the ever-present chance that certification itself turns something up.

| ID | Risk | Probability | Impact | EMV |
|---|---|---|---|---|
| R1 | Paragon's unproven controller tech fails qualification | 30% | \$80,000 | \$24,000 |
| R2 | MCS's move to Mexico delays the first shipment | 25% | \$45,000 | \$11,250 |
| R3 | Fibro Source restructuring delays the carbon-fibre trim package | 20% | \$30,000 | \$6,000 |
| R4 | Toy-safety certification finds a defect requiring redesign | 15% | \$100,000 | \$15,000 |

R1 posts the highest EMV of the four --- not because it carries the biggest single impact (R4's \$100,000 redesign is worse on paper) but because its probability is twice R4's. That is the number Sofia holds up in the meeting: whichever way the Week 10 sourcing decision goes, Paragon cannot be treated as a drop-in substitute without a qualification sample first.

Response strategies, agreed before anyone leaves the room:

- **R1** --- mitigate (require a qualification sample before signing any contract) or avoid entirely (stay with MCS).
- **R2** --- accept, backed by a schedule contingency buffer.
- **R3** --- transfer (a penalty clause in the Fibro Source contract) or mitigate (line up a backup carbon-fibre source).
- **R4** --- mitigate, by running an internal pre-test ahead of the formal certification submission.

Nothing here forces the MCS/Paragon call yet --- that negotiation is still three weeks away --- but the register now puts a number on waiting too long to decide.

:::

## Practice Problems

<details>
<summary>Problem 1: The road-trip risk register (from the course exercise)</summary>

Six friends borrow an older car and drive to an unfamiliar beach. The weather
forecast is mixed with a chance of rain and high winds; one friend is a very
picky eater; the owner plans to replace the car soon due to problems; there is a
fireworks show and all-day music at a crowded venue. List the risks, classify
each, and give a mitigation strategy.

<details>
<summary>Model answer (partial)</summary>

| Risk | Class | Strategy |
|---|---|---|
| Get lost / arrive late | Schedule | *Minimize* --- offline map, leave early |
| Car breaks down | Technology / Resources | *Minimize* (check oil, tires) + *transfer* (roadside assistance) |
| Rain / high winds spoil the day | Weather | *Accept* + *minimize* (pack a tarp, check the radar) |
| Picky eater has nothing to eat | Resources | *Minimize* --- bring a known-safe option |
| No parking at a crowded venue | Resources | *Minimize* --- arrive before the crowd |
| Injury in the crowd / fireworks | Environmental / Safety | *Minimize* --- agree a meeting point, stay together |

</details>

</details>

<details>
<summary>Problem 2: EMV of competing suppliers</summary>

Supplier X: \$140 000, 15 % chance of a 3-week slip costing \$45 000.
Supplier Y: \$128 000, 35 % chance of a 3-week slip costing \$45 000.
On expected cost alone, which is cheaper?

<details>
<summary>Solution</summary>

```{r p2-emv-07, echo=TRUE}
X <- 140000 + 0.15 * 45000
Y <- 128000 + 0.35 * 45000
cat(sprintf("Supplier X expected cost: $%s\n", format(X, big.mark = " ")))
cat(sprintf("Supplier Y expected cost: $%s\n", format(Y, big.mark = " ")))
cat(sprintf("Cheaper on EMV: Supplier %s\n", ifelse(X < Y, "X", "Y")))
```

X: \$146 750. Y: \$143 750. Y is cheaper on expected cost --- but note X's risk is
*less likely*, which may matter more than the \$3 000 if a slip would blow the
holiday launch window. EMV informs the decision; it does not make it.

</details>

</details>

<details>
<summary>Problem 3 (discussion): red-zone risk with no cheap mitigation</summary>

The "carbon-fibre plant disruption" risk sits in the red zone, but the supplier
is the only qualified source and re-qualifying another would take four months you
do not have. What are your options?

<details>
<summary>Discussion points</summary>

You cannot *minimize* the probability much, so work the
**impact**: build a larger **contingency reserve** (time and money) against it;
**transfer** part of it via contract terms (penalty clauses, guaranteed
allocation); **share** it by co-investing in the supplier's modernization for
priority supply; carry **safety stock** of the critical carbon-fibre parts; and
design a **fallback** (an aluminium body variant) that could be substituted if
the supply fails, accepting a heavier product rather than a missed launch.

</details>

</details>

## Further Resources

```{r video-resources-07, echo=FALSE, results='asis'}
embed_youtube("vmkdHT9VmZg", "Decision Tree Analysis for Calculating EMV (PMC Lounge)")
embed_youtube("uC2XBGpIcAs", "Decision Tree and Expected Monetary Value (Cornelius Fichtner)")
```
