---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Project Scheduling

```{r setup-08, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Read and build** a Gantt chart and explain what it adds over a network diagram.
2. **Distinguish** the four precedence relationships (FS, FF, SS, SF) and use leads and lags.
3. **Combine** PERT activity estimates into a project mean and standard deviation.
4. **Calculate** the probability of finishing by a target date.
5. **Compute** an activity's crash cost slope and build a **least-cost crashing** sequence.
6. **Compare** crashing and fast-tracking, and state the risks of each.

<details>
<summary>Before we begin: a schedule is a promise --- how confident are you in it?</summary>

"The project finishes in 83 days" sounds precise. Is that a date you would bet
your bonus on, or a coin flip? This chapter turns a single-number schedule into
a probability you can actually stand behind.

</details>

## Introduction: from a network to a calendar

The AON network from [Chapter 6](06-Network-Diagrams-and-AON.qmd) tells you the
project takes *N* days. The **schedule** puts those days on a calendar, assigns
resources, and gives you something to track progress against. The classic tool
is the **Gantt chart**, developed by Henry Gantt between 1910 and 1915 and used
at scale for the first time on the **Hoover Dam** (1931--1936).

::: {.callout-note}
**Industry snapshot.** An automotive program's master schedule is a Gantt with
thousands of rows rolled up to a one-page milestone view for the steering
committee. A food-plant install schedule shows the trades stacked so the
project manager can see where two crews need the same space on the same day. A
defense schedule is baselined and every slip against it is reported.
:::

```{r video-3point-08, echo=FALSE, results='asis'}
embed_youtube("_YiAhMkBGbQ", "3-Point Estimation for Project Schedule and Cost")
```

## Gantt charts

A Gantt chart is a **time-phased network**: it links each activity to a
calendar. Linked to a schedule, it also **tracks progress against the plan**.
Benefits: easy to read; shows the network *and* the schedule together; supports
updating and control; useful for assigning resources; and easy to create.

```{r fig-gantt-08, fig.cap="A Gantt chart for the RC-vehicle sub-project. Bars are activities; the dark segments show the critical path. Arrows (implied by stacking) show that assembly cannot start until design finishes.", fig.align='center', fig.width=10, fig.height=5.5}
g <- data.frame(
  task = c("A Design outer body", "B Design inner frame", "C Material selection",
           "D Procure motor/controller", "E Wiring harness",
           "F Prototype assembly", "G Drop & EMC test", "H Packaging & approvals"),
  start = c(0, 5, 5, 8, 15, 20, 30, 30),
  dur   = c(5, 8, 3, 12, 5, 10, 8, 12),
  crit  = c(TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE)
)
g$end <- g$start + g$dur
g$task <- factor(g$task, levels = rev(g$task))

ggplot(g) +
  geom_segment(aes(x = start, xend = end, y = task, yend = task, color = crit),
               linewidth = 8, lineend = "butt") +
  geom_text(aes(x = start + dur / 2, y = task, label = paste0(dur, "d")),
            color = "white", fontface = "bold", size = 3.1 * diagram_text_scale) +
  scale_color_manual(values = c("TRUE" = "#C0392B", "FALSE" = "#5DADE2"),
                     labels = c("TRUE" = "Critical path", "FALSE" = "Has float"),
                     name = NULL) +
  labs(title = "RC-vehicle sub-project schedule", x = "Project day", y = NULL) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

## Precedence relationships

Most dependencies are **finish-to-start**, but there are four:

| Relationship | Meaning | Example |
|---|---|---|
| **Finish-to-Start (FS)** | B can't start until A finishes | Build the wall before you paint it |
| **Finish-to-Finish (FF)** | B can't finish until A finishes | Can't finish editing until writing is finished |
| **Start-to-Start (SS)** | B can't start until A starts (often simultaneous) | Start trimming edges when you start mowing |
| **Start-to-Finish (SF)** | A starting triggers B finishing (rare) | The new shift starting ends the old shift |

A **lead** lets a successor start early (overlap); a **lag** forces a wait
(e.g. "wait 2 days for the paint to cure" is FS + 2-day lag).

## Probabilistic scheduling with PERT

A single duration hides uncertainty. Using three-point estimates
(Chapter 6), for each activity:

$$t_e = \frac{o + 4m + p}{6}, \qquad \sigma = \frac{p - o}{6}, \qquad \sigma^2 = \left(\frac{p - o}{6}\right)^2$$

For the **project**, sum along the **critical path**:

$$\mu_{\text{project}} = \sum t_e, \qquad \sigma^2_{\text{project}} = \sum \sigma^2, \qquad \sigma_{\text{project}} = \sqrt{\sigma^2_{\text{project}}}$$

### Worked Example 1: project duration and its spread

The critical path is A--B--C--D--E with these three-point estimates:

Live cell --- change any activity's optimistic, most-likely, or pessimistic estimate.

```{webr-r}
#| label: pert-project-08
cp <- data.frame(
  id = c("A", "B", "C", "D", "E"),
  o  = c(12, 7, 15, 9, 24),
  m  = c(15, 10, 19, 13, 26),
  p  = c(17, 13, 23, 16, 30)
)
cp$te  <- (cp$o + 4 * cp$m + cp$p) / 6
cp$var <- ((cp$p - cp$o) / 6)^2

mu    <- sum(cp$te)
sigma <- sqrt(sum(cp$var))

cp_show <- cp
cp_show[c("te", "var")] <- lapply(cp_show[c("te", "var")], round, 2)
print(cp_show, row.names = FALSE)
cat(sprintf("\nProject mean duration : %.1f days\n", mu))
cat(sprintf("Project variance      : %.2f\n", sum(cp$var)))
cat(sprintf("Project std deviation : %.2f days\n", sigma))
cat(sprintf("Rough 95%% range        : %.0f to %.0f days\n", mu - 2 * sigma, mu + 2 * sigma))
```

### Try it yourself: probability of hitting a target date

Standardize the target: $z = \dfrac{\text{target} - \mu}{\sigma}$, then read the
normal cumulative probability.

```{webr-r}
#| label: prob-on-time-08
mu    <- 83.0     # from the worked example
sigma <- 2.42

for (target in c(80, 83, 86, 88)) {
  z <- (target - mu) / sigma
  cat(sprintf("Finish by day %2d:  z = %+5.2f  ->  P = %4.1f%%\n",
              target, z, 100 * pnorm(z)))
}

cat("\nWhat target gives 90% confidence?\n")
cat(sprintf("  day %.1f  (mu + %.2f x sigma)\n", mu + qnorm(0.90) * sigma, qnorm(0.90)))
```

<details>
<summary>Why quote "day 86 at 90 % confidence" instead of "day 83"?</summary>

Day 83 is the *mean* --- by definition you have only about a 50 % chance of
meeting it. Committing to the mean is committing to a coin flip. Adding a buffer
to reach 85--90 % confidence is how you turn a schedule into a promise you can
keep, and it is the honest version of the "padding" problem from
[Chapter 9](09-Cost-Estimation-and-Budgeting.qmd) --- one visible buffer at the
project level, not hidden padding in every task.

</details>

## Crashing the schedule

**When** you crash: the initial schedule was too optimistic; the market wants
the product sooner; the project has slipped badly; or the contract rewards
finishing early / punishes finishing late.

**How** you accelerate an activity: improve the productivity of existing
resources (focus the team, review best practice); change the working method
(swap drawing review for CAD model review); or add resources (overtime, extra
shifts, expedited freight).

::: {.callout-warning}
**Caution on overtime.** Research on engineering work found optimal marginal
output at about **4 hours of overtime per week**; by **10 hours** of overtime the
*marginal* extra output had fallen to **zero**. More hours is not more work.
:::

Every activity has a **crash cost slope** --- the cost to buy back one time unit:

$$\text{slope} = \frac{\text{crash cost} - \text{normal cost}}{\text{normal time} - \text{crash time}}$$

```{r fig-time-cost-08, fig.cap="Time-cost trade-off for one activity. Between the normal and crash points, buying time gets steadily more expensive per day; past the crash point you cannot buy any more.", fig.align='center', fig.width=9, fig.height=5}
normal_t <- 8; normal_c <- 14000
crash_t  <- 5; crash_c  <- 23000
line <- data.frame(
  t = c(crash_t, normal_t),
  c = c(crash_c, normal_c)
)
ggplot(line, aes(t, c)) +
  geom_line(color = "#2E86C1", linewidth = 1.2) +
  geom_point(size = 4, color = "#1B4F72") +
  annotate("text", x = normal_t, y = normal_c, label = "Normal point\n(8 wk, $14 000)",
           hjust = 1.1, size = 3.2 * diagram_text_scale, lineheight = 0.9) +
  annotate("text", x = crash_t, y = crash_c, label = "Crash point\n(5 wk, $23 000)",
           hjust = -0.1, size = 3.2 * diagram_text_scale, lineheight = 0.9) +
  annotate("text", x = 6.5, y = 15500,
           label = "slope = $9 000 / 3 wk\n= $3 000 per week",
           size = 3.2 * diagram_text_scale, fontface = "italic", lineheight = 0.9) +
  scale_y_continuous(labels = scales::dollar_format(big.mark = " ")) +
  scale_x_reverse(breaks = 5:8) +
  labs(title = "Time-cost trade-off for crashing",
       x = "Activity duration (weeks) - shorter to the right",
       y = "Activity cost") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Try it yourself: least-cost crashing

To shorten the project, crash **critical-path** activities in **cheapest-slope-
first** order, one time unit at a time, until you hit the target or run out of
crashable critical tasks.

```{webr-r}
#| label: crash-sequence-08
# Critical-path activities: normal/crash time and cost.
act <- data.frame(
  id         = c("A", "F", "H"),
  normal_t   = c(9, 6, 8),
  crash_t    = c(7, 5, 4),
  normal_c   = c(5500, 4000, 3500),
  crash_c    = c(7550, 9000, 7550)
)
act$slope     <- (act$crash_c - act$normal_c) / (act$normal_t - act$crash_t)
act$max_crash <- act$normal_t - act$crash_t

cat("Crash cost per day:\n")
print(act[c("id", "slope", "max_crash")], row.names = FALSE)

target_reduction <- 4          # days to remove from the project
order_by_cost <- act[order(act$slope), ]

remaining <- target_reduction
total_cost <- 0
cat(sprintf("\nGoal: remove %d days, cheapest first\n", target_reduction))
for (i in seq_len(nrow(order_by_cost))) {
  take <- min(remaining, order_by_cost$max_crash[i])
  if (take <= 0) break
  cost <- take * order_by_cost$slope[i]
  total_cost <- total_cost + cost
  remaining <- remaining - take
  cat(sprintf("  crash %s by %d day(s) @ $%.0f/day = $%s\n",
              order_by_cost$id[i], take, order_by_cost$slope[i],
              format(cost, big.mark = " ")))
}
cat(sprintf("\nTotal crash cost for %d days: $%s\n",
            target_reduction - remaining, format(total_cost, big.mark = " ")))
if (remaining > 0) cat(sprintf("Still %d day(s) short -- no cheap critical task left to crash.\n", remaining))
```

## Fast-tracking versus crashing

| | **Crashing** | **Fast-tracking** |
|---|---|---|
| Method | Add resources (overtime, staff, expediting) | Overlap tasks that were sequential |
| Cost | Always adds cost | Little or no added cost |
| Risk | Diminishing returns; overtime burnout | Rework if the predecessor changes after the successor started |

Both only help on the **critical path**, and after either one you must
**re-check the network** --- the critical path can move.

## Key Takeaways

- A **Gantt chart** puts the network on a calendar and lets you track progress; it shows activities, dependencies, and resources at once.
- Four precedence types: **FS** (default), **FF**, **SS**, **SF** --- with **leads** (overlap) and **lags** (forced wait).
- PERT project: $\mu = \sum t_e$ and $\sigma = \sqrt{\sum \sigma^2}$ along the **critical path**. The mean is a ~50 % date.
- Probability of on-time: $z = (\text{target} - \mu)/\sigma$, then the normal CDF. Commit to an ~85--90 % date, not the mean.
- **Crash cost slope** $=$ (crash cost − normal cost) / (normal time − crash time). Crash **critical tasks, cheapest slope first**.
- **Crashing** buys time with money; **fast-tracking** buys it with rework risk. Re-check the critical path after either.

## Case Study: Meridian R/C — Week 8

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 8: Can the Schedule Be Pulled In?"}

This week you pull up the Gantt chart derived from the fixed CPM network for the
Trailblazer prototype --- 10 activities, 67 working days, critical path
A--B--E--F--H--I--J --- and the picture is uncomfortably tight. Every activity
on that spine has zero float: concept CAD, detailed design release, motor &
controls procurement (E, 25 days), prototype assembly (F), toy-safety
certification (H, 12 days), packaging & client sample ship, and your sign-off
report to Renata Ford. The only slack anywhere in the network sits in body
tooling (C, 5 days float), frame tooling (D, 10 days float), and the
drop/durability test (G, 7 days float) --- none of which help you finish
sooner.

Daniel Cho, sitting in on the Friday steering call, asks the question you knew
was coming: with Kestrel's 100,000-unit holiday-season order riding on this
prototype, can the 67-day timeline be pulled in? You walk the team through the
two realistic levers, both of which only work because E and H --- and nothing
else in the network --- are actually critical.

**Crashing E** (motor & controls procurement) buys up to 5 days, at **$1,200
per day** for air freight and expedited customs --- a maximum of $6,000 to
shave the full 5 days off the 25-day activity.

**Crashing H** (toy-safety certification) buys up to 4 days, at **$900 per
day** for priority lab review --- a maximum of $3,600 to shave the full 4 days
off the 12-day activity.

There is also a **fast-tracking** option on H: instead of waiting for F
(assembly) to finish before booking the certification lab, Sofia Reyes could
pre-book the lab slot while assembly is still underway. That saves roughly 3
days at little direct cost --- but if a late change during assembly affects
the sample, the pre-booked test may have to be repeated, eating into the time
it was meant to save.

Laying the numbers side by side for Daniel, one thing jumps out: dollar for
dollar, **H is actually the cheaper activity to crash** ($900/day versus E's
$1,200/day), even though E offers more total days of compression (5 versus
4). Applying the chapter's own cheapest-slope-first rule, H is the activity to
crash first if only a few days are needed; E is the fallback if more days are
required than H alone can supply. Whichever lever gets pulled, the team agrees
to re-check the critical path afterward --- crashing or fast-tracking either
activity could shift where the true bottleneck sits.

:::

The question pool below tests this Week 8 episode specifically; it does not
retest the general chapter content already covered elsewhere.

## Practice Problems

<details>
<summary>Problem 1: Least-cost crash to a deadline</summary>

Project critical path A--F--H, currently 23 weeks; the contract needs 20. Crash
data: A \$513/wk (max 4 wk), F \$1 500/wk (max 1 wk), H \$950/wk (max 2 wk).
Find the least-cost way to remove 3 weeks.

**Solution.**

```{r p1-crash-08, echo=TRUE}
crash <- data.frame(
  id = c("A", "F", "H"),
  slope = c(513, 1500, 950),
  max = c(4, 1, 2)
)
need <- 3
crash <- crash[order(crash$slope), ]
left <- need; cost <- 0
for (i in seq_len(nrow(crash))) {
  take <- min(left, crash$max[i]); if (take <= 0) break
  cost <- cost + take * crash$slope[i]; left <- left - take
  cat(sprintf("Crash %s by %d wk @ $%d = $%s\n", crash$id[i], take, crash$slope[i],
              format(take * crash$slope[i], big.mark = " ")))
}
cat(sprintf("Total: $%s for %d weeks\n", format(cost, big.mark = " "), need - left))
```

Crash A by 3 weeks for \$1 539 --- far cheaper than touching F or H.

</details>

<details>
<summary>Problem 2: Confidence in a date</summary>

A project's critical path has $\mu = 60$ days and $\sigma = 4$ days. The customer
wants delivery in 64 days.

(a) What is the probability of meeting day 64? (b) What date should you commit to
for 95 % confidence?

**Solution.**

```{r p2-prob-08, echo=TRUE}
mu <- 60; sigma <- 4
cat(sprintf("P(finish by day 64) = %.1f%%\n", 100 * pnorm((64 - mu) / sigma)))
cat(sprintf("95%% confident date  = day %.1f\n", mu + qnorm(0.95) * sigma))
```

(a) ~84 %. (b) Day 66.6 --- commit to day 67.

</details>

<details>
<summary>Problem 3 (discussion): crash or fast-track the RC-vehicle launch?</summary>

The retailer moves the launch three weeks earlier. Design is on the critical
path and is followed by prototype assembly (also critical). You could add a
second designer and pay overtime (crash), or start building the frame jig from
the released frame drawings while the body design is still being finalized
(fast-track). Which, and why?

**Discussion points.** Fast-tracking the jig against the *frame* drawings is
low-cost and reasonably safe *if* the frame design is stable --- but if the body
design forces a late frame change, the jig is rework. Crashing design with a
second modeller adds cost and has coordination overhead (two people on one
model), but no rework risk. A common answer: fast-track the clearly stable parts,
crash the rest, and hold a contingency for the rework risk you accepted.

</details>

## Further Resources

```{r video-resources-08, echo=FALSE, results='asis'}
embed_youtube("6B7Qtc3rrQ4", "PERT / Three-Point Estimation Technique")
embed_youtube("4oDLMs11Exs", "Forward and Backward Pass -> Project Duration and Critical Path")
```
