---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Project Scheduling

```{r setup-08, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Read and build** a Gantt chart and explain what it adds over a network diagram.
2. **Distinguish** the four precedence relationships (FS, FF, SS, SF) and use leads and lags.
3. **Combine** PERT activity estimates into a project mean and standard deviation.
4. **Calculate** the probability of finishing by a target date.
5. **Compute** an activity's crash cost slope and build a **least-cost crashing** sequence.
6. **Compare** crashing and fast-tracking, and state the risks of each.

<details>
<summary>Before we begin: a schedule is a promise --- how confident are you in it?</summary>

"The project finishes in 83 days" sounds precise. Is that a date you would bet
your bonus on, or a coin flip? This chapter turns a single-number schedule into
a probability you can actually stand behind.

</details>

## Introduction: from a network to a calendar

The AON network from [Chapter 6](06-Network-Diagrams-and-AON.qmd) tells you the
project takes *N* days. The **schedule** puts those days on a calendar, assigns
resources, and gives you something to track progress against. The classic tool
is the **Gantt chart**, developed by Henry Gantt between 1910 and 1915 and used
at scale for the first time on the **Hoover Dam** (1931--1936).

::: {.callout-note}
**Industry snapshot.** An automotive program's master schedule is a Gantt with
thousands of rows rolled up to a one-page milestone view for the steering
committee. A food-plant install schedule shows the trades stacked so the
project manager can see where two crews need the same space on the same day. A
defense schedule is baselined and every slip against it is reported.
:::

```{r video-3point-08, echo=FALSE, results='asis'}
embed_youtube("_YiAhMkBGbQ", "3-Point Estimation for Project Schedule and Cost")
```

## Gantt charts

A Gantt chart is a **time-phased network**: it links each activity to a
calendar. Linked to a schedule, it also **tracks progress against the plan**.
Benefits: easy to read; shows the network *and* the schedule together; supports
updating and control; useful for assigning resources; and easy to create.

```{r fig-gantt-08, fig.cap="A Gantt chart for the RC-vehicle sub-project. Bars are activities; the dark segments show the critical path. Arrows (implied by stacking) show that assembly cannot start until design finishes.", fig.align='center', fig.width=10, fig.height=5.5}
g <- data.frame(
  task = c("A Design outer body", "B Design inner frame", "C Material selection",
           "D Procure motor/controller", "E Wiring harness",
           "F Prototype assembly", "G Drop & EMC test", "H Packaging & approvals"),
  start = c(0, 5, 5, 8, 15, 20, 30, 30),
  dur   = c(5, 8, 3, 12, 5, 10, 8, 12),
  crit  = c(TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE)
)
g$end <- g$start + g$dur
g$task <- factor(g$task, levels = rev(g$task))

ggplot(g) +
  geom_segment(aes(x = start, xend = end, y = task, yend = task, color = crit),
               linewidth = 8, lineend = "butt") +
  geom_text(aes(x = start + dur / 2, y = task, label = paste0(dur, "d")),
            color = "white", fontface = "bold", size = 3.1 * diagram_text_scale) +
  scale_color_manual(values = c("TRUE" = "#C0392B", "FALSE" = "#5DADE2"),
                     labels = c("TRUE" = "Critical path", "FALSE" = "Has float"),
                     name = NULL) +
  labs(title = "RC-vehicle sub-project schedule", x = "Project day", y = NULL) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

## Precedence relationships

Most dependencies are **finish-to-start**, but there are four:

| Relationship | Meaning | Example |
|---|---|---|
| **Finish-to-Start (FS)** | B can't start until A finishes | Build the wall before you paint it |
| **Finish-to-Finish (FF)** | B can't finish until A finishes | Can't finish editing until writing is finished |
| **Start-to-Start (SS)** | B can't start until A starts (often simultaneous) | Start trimming edges when you start mowing |
| **Start-to-Finish (SF)** | A starting triggers B finishing (rare) | The new shift starting ends the old shift |

A **lead** lets a successor start early (overlap); a **lag** forces a wait
(e.g. "wait 2 days for the paint to cure" is FS + 2-day lag).

### Try it yourself: which relationship is it?

```{r tryit-relations-08, echo=FALSE, results='asis'}
rel_bins <- c("Finish-to-Start (FS)", "Finish-to-Finish (FF)", "Start-to-Start (SS)", "Start-to-Finish (SF)")
tryit_sort(
  "precedence-types", "Which precedence relationship does each statement describe?",
  bins = rel_bins,
  cards = data.frame(
    text = c("The concrete must cure before the wall is painted.",
             "Editing cannot finish until the writing has finished.",
             "Trimming the edges starts when the mowing starts.",
             "The old system is switched off only once the new system has started up.",
             "Final testing cannot be completed until the last build is completed.",
             "Cleaning of each room can begin once painting in that room has begun.",
             "Assembly begins only after the long-lead part has arrived.",
             "The outgoing shift stops only when the incoming shift has started."),
    bin = c("Finish-to-Start (FS)", "Finish-to-Finish (FF)", "Start-to-Start (SS)", "Start-to-Finish (SF)",
            "Finish-to-Finish (FF)", "Start-to-Start (SS)", "Finish-to-Start (FS)", "Start-to-Finish (SF)"),
    why = c("The predecessor finishes, then the successor starts: FS, the usual case.",
            "B cannot finish until A finishes: FF.", "B starts when A starts: SS.",
            "A starting triggers B finishing: the rare SF.", "The two finishes are tied together: FF.",
            "The two starts are tied together: SS.", "FS: wait for the part, then start.",
            "The successor's start releases the predecessor's finish: SF."),
    stringsAsFactors = FALSE),
  intro = "Pick the relationship each statement describes.", question = NULL)

tryit_sort(
  "precedence-twin", "Precedence relationships (graded)", bins = rel_bins, show = FALSE,
  cards = data.frame(
    text = c("Wiring cannot start until the conduit is installed.",
             "Documentation cannot be completed until the software is completed.",
             "Surveying starts as soon as clearing starts.",
             "The temporary supply can be shut off only after the permanent supply has started.",
             "The first article is inspected once machining of it has ended."),
    bin = c("Finish-to-Start (FS)", "Finish-to-Finish (FF)", "Start-to-Start (SS)",
            "Start-to-Finish (SF)", "Finish-to-Start (FS)"),
    stringsAsFactors = FALSE),
  question = "<p>For each statement, choose the precedence relationship between the two activities.</p>")
```

## Probabilistic scheduling with PERT

A single duration hides uncertainty. Using three-point estimates
(Chapter 6), for each activity:

$$t_e = \frac{o + 4m + p}{6}, \qquad \sigma = \frac{p - o}{6}, \qquad \sigma^2 = \left(\frac{p - o}{6}\right)^2$$

For the **project**, sum along the **critical path**:

$$\mu_{\text{project}} = \sum t_e, \qquad \sigma^2_{\text{project}} = \sum \sigma^2, \qquad \sigma_{\text{project}} = \sqrt{\sigma^2_{\text{project}}}$$

### Worked Example 1: project duration and its spread

The critical path is A--B--C--D--E with these three-point estimates:

Live cell --- change any activity's optimistic, most-likely, or pessimistic estimate.

```{webr-r}
#| label: pert-project-08
cp <- data.frame(
  id = c("A", "B", "C", "D", "E"),
  o  = c(12, 7, 15, 9, 24),
  m  = c(15, 10, 19, 13, 26),
  p  = c(17, 13, 23, 16, 30)
)
cp$te  <- (cp$o + 4 * cp$m + cp$p) / 6
cp$var <- ((cp$p - cp$o) / 6)^2

mu    <- sum(cp$te)
sigma <- sqrt(sum(cp$var))

cp_show <- cp
cp_show[c("te", "var")] <- lapply(cp_show[c("te", "var")], round, 2)
print(cp_show, row.names = FALSE)
cat(sprintf("\nProject mean duration : %.1f days\n", mu))
cat(sprintf("Project variance      : %.2f\n", sum(cp$var)))
cat(sprintf("Project std deviation : %.2f days\n", sigma))
cat(sprintf("Rough 95%% range        : %.0f to %.0f days\n", mu - 2 * sigma, mu + 2 * sigma))
```

### Try it yourself: probability of hitting a target date

Standardize the target: $z = \dfrac{\text{target} - \mu}{\sigma}$, then read the
normal cumulative probability.

```{r calc-prob-on-time-08, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "prob-on-time-08",
  title = "Probability-of-On-Time Calculator",
  icon = "bell-concierge",
  inputs = list(
    qwebr_calc_input("mu", "Project mean duration", value = 83.0, min = 40, max = 200, step = 0.5, suffix = "days"),
    qwebr_calc_input("sigma", "Project standard deviation", value = 2.42, min = 0.5, max = 15, step = 0.01, suffix = "days"),
    qwebr_calc_input("confidence_pct", "Confidence you want", value = 90, min = 50, max = 99, step = 1, suffix = "%"),
    qwebr_calc_input("target_day", "Target finish day", value = 86, min = 60, max = 120, step = 1, type = "slider")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: prob-on-time-08
mu    <- 83.0     # from the worked example
sigma <- 2.42
confidence_pct <- 90    # confidence you want in the target
target_day     <- 86    # the day you are promising

z    <- (target_day - mu) / sigma
prob <- pnorm(z)
conf_day <- mu + qnorm(confidence_pct / 100) * sigma

x <- seq(mu - 4 * sigma, mu + 4 * sigma, length.out = 200)
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
plot(x, dnorm(x, mu, sigma), type = "n", xlab = "Project finish (day)", ylab = "Density",
     main = "Probability of finishing on time")
shade <- x <= target_day
if (any(shade)) polygon(c(x[shade][1], x[shade], target_day), c(0, dnorm(x[shade], mu, sigma), 0),
                        col = "#aed6f1", border = NA)
lines(x, dnorm(x, mu, sigma), lwd = 4, col = "#2c3e50")
abline(v = target_day, lty = 2, lwd = 2, col = "#e74c3c")
abline(v = conf_day, lty = 3, lwd = 2, col = "#16a085")

if (100 * prob >= confidence_pct) {
  cat(sprintf("P(finish by day %g) = %.1f%%: meets the %g%% goal\n", target_day, 100 * prob, confidence_pct))
} else {
  cat(sprintf("!!P(finish by day %g) = %.1f%%: below the %g%% goal!!\n", target_day, 100 * prob, confidence_pct))
}
cat(sprintf("z-score  %+.2f\n", z))
cat(sprintf("Day for %g%% confidence  day %.1f (mu + %.2f x sigma)\n",
            confidence_pct, conf_day, qnorm(confidence_pct / 100)))
```

</details>

<details>
<summary>Why quote "day 86 at 90 % confidence" instead of "day 83"?</summary>

Day 83 is the *mean* --- by definition you have only about a 50 % chance of
meeting it. Committing to the mean is committing to a coin flip. Adding a buffer
to reach 85--90 % confidence is how you turn a schedule into a promise you can
keep, and it is the honest version of the "padding" problem from
[Chapter 9](09-Cost-Estimation-and-Budgeting.qmd) --- one visible buffer at the
project level, not hidden padding in every task.

</details>

### Try it yourself: how likely is the target date?

Take the three-point estimates for the activities on the critical path and
work through to the probability of finishing by the target. Each team picks
its own **data set** (and its own target day).

```{r tryit-pert-prob-08, echo=FALSE, results='asis'}
pp_sets <- list(
  list(o = c(4, 8, 5, 6, 3),  m = c(6, 10, 7, 9, 4),  p = c(10, 16, 9, 18, 5),  target = 41),
  list(o = c(5, 10, 4, 7, 2), m = c(8, 12, 6, 10, 3), p = c(17, 14, 14, 13, 10), target = 45),
  list(o = c(3, 6, 8, 4, 5),  m = c(5, 9, 10, 6, 8),  p = c(13, 12, 12, 8, 17), target = 43),
  list(o = c(10, 2, 6, 12, 4), m = c(14, 5, 8, 15, 5), p = c(18, 14, 10, 24, 6), target = 52))
pp_table <- function(s) {
  data.frame(Activity = LETTERS[1:5], o = s$o, m = s$m, p = s$p,
             `t_e` = (s$o + 4 * s$m + s$p) / 6, Variance = ((s$p - s$o) / 6)^2)
}
pp_answers <- function(s) {
  d <- pp_table(s); mu <- sum(d$`t_e`); sg <- sqrt(sum(d$Variance)); z <- (s$target - mu) / sg
  list(tryit_answer("Project mean duration", mu, tol = 0.05, suffix = "days"),
       tryit_answer("Project standard deviation", sg, tol = 0.02, suffix = "days"),
       tryit_answer(sprintf("z for a target of day %d", s$target), z, tol = 0.02),
       tryit_answer(sprintf("Probability of finishing by day %d", s$target), 100 * pnorm(z), tol = 0.6, suffix = "%"),
       tryit_answer("Day to promise for 90 % confidence", mu + qnorm(0.90) * sg, tol = 0.1, suffix = "day"))
}
tryit_grid(
  "pert-probability", "Probability of finishing on time",
  data = pp_table(pp_sets[[1]]), blank = c("t_e", "Variance"), tol = 0.01, digits = 0,
  answers = pp_answers(pp_sets[[1]]),
  teams = lapply(pp_sets[-1], function(s) list(data = pp_table(s), answers = pp_answers(s))),
  hints = c(`t_e` = "(o + 4m + p) / 6.", Variance = "((p - o) / 6) squared."),
  intro = "Variance is the square of the standard deviation. Two decimal places.")

pp_twin <- list(o = c(6, 4, 9, 3, 7), m = c(8, 6, 12, 5, 9), p = c(16, 8, 21, 9, 17), target = 47)
tryit_grid(
  "pert-probability-twin", "Probability of finishing on time (graded)", show = FALSE,
  data = pp_table(pp_twin), blank = c("t_e", "Variance"), tol = 0.01, digits = 0,
  answers = pp_answers(pp_twin),
  question = paste0("<p>The critical path A to E has the three-point estimates below (days). Calculate ",
    "each activity's expected duration t<sub>e</sub> and variance, then the project mean and standard ",
    "deviation, the z-score and probability of finishing by day 47, and the day to promise for 90 % ",
    "confidence (two decimal places).</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

For data set 1: t_e = 6.33, 10.67, 7.00, 10.00, 4.00, which sum to a mean of
**38.0 days**; the variances 1.00, 1.78, 0.44, 4.00 and 0.11 sum to 7.33, so
sigma = **2.71 days**. For a target of day 41, z = (41 - 38) / 2.71 = 1.11 and
the normal CDF gives about **87 %**. For 90 % confidence use
38.0 + 1.2816 x 2.71 = **day 41.5**. The mean (38) alone would be a 50 % date.

</details>

### Try it yourself: simulate the schedule

The formula above assumes the project total is normally distributed. Check it
by brute force: each run draws every activity's duration from a PERT
(beta-shaped) distribution and adds them up. Press **Run again** for a new set
of 2 000 runs; in class, each team enters its own run number.

```{r calc-schedule-sim-08, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "schedule-sim-08",
  title = "Schedule Simulation",
  icon = "dice",
  inputs = list(
    qwebr_calc_input("seed", "Run number (team number)", value = 1, min = 1, max = 999, step = 1),
    qwebr_calc_input("target_day", "Target finish day", value = 80, min = 60, max = 110, step = 1, type = "slider")
  ),
  rerun = "seed",
  goal = "Move the target day until at least 90 % of the simulated runs finish on time. Does the formula's 90 % day agree?",
  goal_test = "pct_on_time >= 90"
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this simulation</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: schedule-sim-08
seed       <- 1
target_day <- 80

# The chapter's critical path A to E (days): optimistic, most likely, pessimistic.
o <- c(12, 7, 15, 9, 24)
m <- c(15, 10, 19, 13, 26)
p <- c(17, 13, 23, 16, 30)
runs <- 2000

set.seed(seed)
# PERT-beta: mean (o + 4m + p) / 6, spread set by the same three numbers.
a <- 1 + 4 * (m - o) / (p - o)
b <- 1 + 4 * (p - m) / (p - o)
total <- numeric(runs)
for (i in seq_along(o)) total <- total + o[i] + (p[i] - o[i]) * rbeta(runs, a[i], b[i])

pct_on_time <- 100 * mean(total <= target_day)
mu    <- sum((o + 4 * m + p) / 6)
sigma <- sqrt(sum(((p - o) / 6)^2))
formula_pct <- 100 * pnorm((target_day - mu) / sigma)

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
hist(total, breaks = 30, col = "#aed6f1", border = "white",
     xlab = "Project finish (day)", ylab = "Runs", main = "Simulated project durations")
abline(v = target_day, lty = 2, lwd = 3, col = "#e74c3c")

if (pct_on_time >= 90) {
  cat(sprintf("%.1f%% of runs finish by day %g: meets a 90%% promise\n", pct_on_time, target_day))
} else {
  cat(sprintf("%.1f%% of runs finish by day %g: !!below a 90%% promise!!\n", pct_on_time, target_day))
}
cat(sprintf("Normal-curve formula  %.1f%%\n", formula_pct))
cat(sprintf("Mean simulated finish  day %.1f\n", mean(total)))
cat(sprintf("Middle 90%% of runs  day %.1f to %.1f\n", quantile(total, 0.05), quantile(total, 0.95)))
```

</details>

## Crashing the schedule

**When** you crash: the initial schedule was too optimistic; the market wants
the product sooner; the project has slipped badly; or the contract rewards
finishing early / punishes finishing late.

**How** you accelerate an activity: improve the productivity of existing
resources (focus the team, review best practice); change the working method
(swap drawing review for CAD model review); or add resources (overtime, extra
shifts, expedited freight).

::: {.callout-warning}
**Caution on overtime.** Research on engineering work found optimal marginal
output at about **4 hours of overtime per week**; by **10 hours** of overtime the
*marginal* extra output had fallen to **zero**. More hours is not more work.
:::

Every activity has a **crash cost slope** --- the cost to buy back one time unit:

$$\text{slope} = \frac{\text{crash cost} - \text{normal cost}}{\text{normal time} - \text{crash time}}$$

```{r fig-time-cost-08, fig.cap="Time-cost trade-off for one activity. Between the normal and crash points, buying time gets steadily more expensive per day; past the crash point you cannot buy any more.", fig.align='center', fig.width=9, fig.height=5}
normal_t <- 8; normal_c <- 14000
crash_t  <- 5; crash_c  <- 23000
line <- data.frame(
  t = c(crash_t, normal_t),
  c = c(crash_c, normal_c)
)
ggplot(line, aes(t, c)) +
  geom_line(color = "#2E86C1", linewidth = 1.2) +
  geom_point(size = 4, color = "#1B4F72") +
  annotate("text", x = normal_t, y = normal_c, label = "Normal point\n(8 wk, $14 000)",
           hjust = 0.1, vjust = -1.3, size = 3.2 * diagram_text_scale, lineheight = 0.9) +
  annotate("text", x = crash_t, y = crash_c, label = "Crash point\n(5 wk, $23 000)",
           hjust = -0.1, size = 3.2 * diagram_text_scale, lineheight = 0.9) +
  annotate("text", x = 6.5, y = 15500,
           label = "slope = $9 000 / 3 wk\n= $3 000 per week",
           size = 3.2 * diagram_text_scale, fontface = "italic", lineheight = 0.9) +
  scale_y_continuous(labels = scales::dollar_format(big.mark = " "),
                     expand = expansion(mult = c(0.08, 0.08))) +
  scale_x_reverse(breaks = 5:8, expand = expansion(mult = c(0.1, 0.1))) +
  coord_cartesian(clip = "off") +
  labs(title = "Time-cost trade-off for crashing",
       x = "Activity duration (weeks) - shorter to the right",
       y = "Activity cost") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale),
    plot.margin = margin(t = 5.5, r = 65, b = 5.5, l = 65)
  )
```

### Try it yourself: least-cost crashing

To shorten the project, crash **critical-path** activities in **cheapest-slope-
first** order, one time unit at a time, until you hit the target or run out of
crashable critical tasks.

```{r calc-crash-sequence-08, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "crash-sequence-08",
  title = "Crash-Cost Calculator",
  icon = "bolt",
  inputs = list(
    qwebr_calc_input("target_reduction", "Days to remove from the project", value = 4, min = 1, max = 8, step = 1, type = "slider")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: crash-sequence-08
target_reduction <- 4          # days to remove from the project

# Critical-path activities: normal/crash time and cost.
act <- data.frame(
  id         = c("A", "F", "H"),
  normal_t   = c(9, 6, 8),
  crash_t    = c(7, 5, 4),
  normal_c   = c(5500, 4000, 3500),
  crash_c    = c(7550, 9000, 7550)
)
act$slope     <- (act$crash_c - act$normal_c) / (act$normal_t - act$crash_t)
act$max_crash <- act$normal_t - act$crash_t

order_by_cost <- act[order(act$slope), ]
order_by_cost$days_used <- 0

remaining <- target_reduction
total_cost <- 0
steps <- character(0)
for (i in seq_len(nrow(order_by_cost))) {
  take <- min(remaining, order_by_cost$max_crash[i])
  if (take <= 0) break
  cost <- take * order_by_cost$slope[i]
  total_cost <- total_cost + cost
  remaining <- remaining - take
  order_by_cost$days_used[i] <- take
  steps <- c(steps, sprintf("Crash %s by %g day(s) at $%s/day  $%s",
                            order_by_cost$id[i], take, format(round(order_by_cost$slope[i]), big.mark = " "),
                            format(round(cost), big.mark = " ")))
}

par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
bp <- barplot(order_by_cost$slope, names.arg = order_by_cost$id, col = ifelse(order_by_cost$days_used > 0, "#2980b9", "#bdc3c7"),
              ylim = c(0, max(order_by_cost$slope) * 1.2),
              xlab = "Activity (cheapest first)", ylab = "Crash cost ($ per day)",
              main = "Crash cost per day (blue = crashed)")
text(bp, order_by_cost$slope, labels = ifelse(order_by_cost$days_used > 0, paste(order_by_cost$days_used, "d"), ""),
     pos = 3, cex = 1.3)

if (remaining > 0) {
  cat(sprintf("!!Only %g of %g days can be crashed: $%s!!\n", target_reduction - remaining,
              target_reduction, format(round(total_cost), big.mark = " ")))
} else {
  cat(sprintf("Cheapest way to cut %g days: $%s\n", target_reduction, format(round(total_cost), big.mark = " ")))
}
for (s in steps) cat(s, "\n", sep = "")
```

</details>

### Try it yourself: crash slopes and the cheapest cut

All four activities are on the critical path. Work out each slope and each
activity's maximum crash, then the least cost of the cut asked for. Each team
picks its own **data set**.

```{r tryit-crash-08, echo=FALSE, results='asis'}
cr_sets <- list(
  list(t0 = c(10, 8, 6, 9), t1 = c(7, 6, 5, 6), c0 = c(20000, 15000, 9000, 18000), c1 = c(29000, 19000, 12500, 23400), cut = 5),
  list(t0 = c(12, 7, 10, 5), t1 = c(9, 5, 8, 4),  c0 = c(22000, 14000, 16000, 8000),  c1 = c(28000, 19000, 22000, 9800),  cut = 4),
  list(t0 = c(9, 7, 5, 8),   t1 = c(6, 4, 3, 7),  c0 = c(12000, 10000, 7000, 11000),  c1 = c(17400, 17500, 10200, 13300), cut = 6),
  list(t0 = c(6, 11, 8, 10), t1 = c(4, 8, 7, 6),  c0 = c(9000, 20000, 12000, 15000),  c1 = c(13400, 26900, 14100, 25000), cut = 5))
cr_table <- function(s) {
  data.frame(Activity = LETTERS[1:4], `Normal (wk)` = s$t0, `Crash (wk)` = s$t1,
             `Normal cost ($)` = s$c0, `Crash cost ($)` = s$c1,
             `Slope ($/wk)` = (s$c1 - s$c0) / (s$t0 - s$t1), `Max crash (wk)` = s$t0 - s$t1,
             check.names = FALSE)
}
cr_answers <- function(s) {
  d <- cr_table(s); d <- d[order(d$`Slope ($/wk)`), ]
  left <- s$cut; cost <- 0
  for (i in seq_len(nrow(d))) { take <- min(left, d$`Max crash (wk)`[i]); cost <- cost + take * d$`Slope ($/wk)`[i]; left <- left - take }
  list(tryit_answer(sprintf("Least cost to cut %d weeks", s$cut), cost, tol = 1, prefix = "$"))
}
cr_cols <- c("Slope ($/wk)", "Max crash (wk)")
tryit_grid(
  "crash-slopes", "Crash cost slopes", data = cr_table(cr_sets[[1]]), blank = cr_cols, tol = 0.5, digits = 0,
  answers = cr_answers(cr_sets[[1]]),
  teams = lapply(cr_sets[-1], function(s) list(data = cr_table(s), answers = cr_answers(s))),
  hints = c(`Slope ($/wk)` = "(crash cost - normal cost) / (normal time - crash time).",
            `Max crash (wk)` = "Normal time minus crash time."),
  intro = "Crash the cheapest slope first, up to each activity's maximum.")

cr_twin <- list(t0 = c(8, 10, 6, 7), t1 = c(6, 7, 4, 6), c0 = c(10000, 21000, 8000, 12000),
                c1 = c(14400, 27600, 12000, 14100), cut = 5)
tryit_grid(
  "crash-slopes-twin", "Crash cost slopes (graded)", show = FALSE,
  data = cr_table(cr_twin), blank = cr_cols, tol = 0.5, digits = 0, answers = cr_answers(cr_twin),
  question = paste0("<p>Four activities in series make up the critical path. Normal and crash times ",
    "(weeks) and costs ($) are given. Calculate each crash cost slope and maximum crash, then the ",
    "least cost of shortening the project by 5 weeks.</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

Slope = (crash cost - normal cost) / (normal time - crash time). For data set
1 the slopes are A \$3 000, B \$2 000, C \$3 500 and D \$1 800 per week. Cheapest
first: D for 3 weeks (\$5 400), then B for 2 weeks (\$4 000), so the 5-week cut
costs **\$9 400**. A and C are never touched. Re-check the network afterwards:
the critical path can shift.

</details>

## Fast-tracking versus crashing

| | **Crashing** | **Fast-tracking** |
|---|---|---|
| Method | Add resources (overtime, staff, expediting) | Overlap tasks that were sequential |
| Cost | Always adds cost | Little or no added cost |
| Risk | Diminishing returns; overtime burnout | Rework if the predecessor changes after the successor started |

Both only help on the **critical path**, and after either one you must
**re-check the network** --- the critical path can move.

## Key Takeaways

- A **Gantt chart** puts the network on a calendar and lets you track progress; it shows activities, dependencies, and resources at once.
- Four precedence types: **FS** (default), **FF**, **SS**, **SF** --- with **leads** (overlap) and **lags** (forced wait).
- PERT project: $\mu = \sum t_e$ and $\sigma = \sqrt{\sum \sigma^2}$ along the **critical path**. The mean is a ~50 % date.
- Probability of on-time: $z = (\text{target} - \mu)/\sigma$, then the normal CDF. Commit to an ~85--90 % date, not the mean.
- **Crash cost slope** $=$ (crash cost − normal cost) / (normal time − crash time). Crash **critical tasks, cheapest slope first**.
- **Crashing** buys time with money; **fast-tracking** buys it with rework risk. Re-check the critical path after either.

## Case Study: Meridian R/C — Week 8

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 8: Can the Schedule Be Pulled In?"}

This week you pull up the Gantt chart derived from the fixed CPM network for the
Trailblazer prototype --- 10 activities, 67 working days, critical path
A--B--E--F--H--I--J --- and the picture is uncomfortably tight. Every activity
on that spine has zero float: concept CAD, detailed design release, motor &
controls procurement (E, 25 days), prototype assembly (F), toy-safety
certification (H, 12 days), packaging & client sample ship, and your sign-off
report to Renata Ford. The only slack anywhere in the network sits in body
tooling (C, 5 days float), frame tooling (D, 10 days float), and the
drop/durability test (G, 7 days float) --- none of which help you finish
sooner.

Daniel Cho, sitting in on the Friday steering call, asks the question you knew
was coming: with Kestrel's 100,000-unit holiday-season order riding on this
prototype, can the 67-day timeline be pulled in? You walk the team through the
two realistic levers, both of which only work because E and H --- and nothing
else in the network --- are actually critical.

**Crashing E** (motor & controls procurement) buys up to 5 days, at **$1,200
per day** for air freight and expedited customs --- a maximum of $6,000 to
shave the full 5 days off the 25-day activity.

**Crashing H** (toy-safety certification) buys up to 4 days, at **$900 per
day** for priority lab review --- a maximum of $3,600 to shave the full 4 days
off the 12-day activity.

There is also a **fast-tracking** option on H: instead of waiting for F
(assembly) to finish before booking the certification lab, Sofia Reyes could
pre-book the lab slot while assembly is still underway. That saves roughly 3
days at little direct cost --- but if a late change during assembly affects
the sample, the pre-booked test may have to be repeated, eating into the time
it was meant to save.

Laying the numbers side by side for Daniel, one thing jumps out: dollar for
dollar, **H is actually the cheaper activity to crash** ($900/day versus E's
$1,200/day), even though E offers more total days of compression (5 versus
4). Applying the chapter's own cheapest-slope-first rule, H is the activity to
crash first if only a few days are needed; E is the fallback if more days are
required than H alone can supply. Whichever lever gets pulled, the team agrees
to re-check the critical path afterward --- crashing or fast-tracking either
activity could shift where the true bottleneck sits.

:::

## Practice Problems

<details>
<summary>Problem 1: Least-cost crash to a deadline</summary>

Project critical path A--F--H, currently 23 weeks; the contract needs 20. Crash
data: A \$513/wk (max 4 wk), F \$1 500/wk (max 1 wk), H \$950/wk (max 2 wk).
Find the least-cost way to remove 3 weeks.

<details>
<summary>Solution</summary>

```{r p1-crash-08, echo=TRUE}
crash <- data.frame(
  id = c("A", "F", "H"),
  slope = c(513, 1500, 950),
  max = c(4, 1, 2)
)
need <- 3
crash <- crash[order(crash$slope), ]
left <- need; cost <- 0
for (i in seq_len(nrow(crash))) {
  take <- min(left, crash$max[i]); if (take <= 0) break
  cost <- cost + take * crash$slope[i]; left <- left - take
  cat(sprintf("Crash %s by %d wk @ $%d = $%s\n", crash$id[i], take, crash$slope[i],
              format(take * crash$slope[i], big.mark = " ")))
}
cat(sprintf("Total: $%s for %d weeks\n", format(cost, big.mark = " "), need - left))
```

Crash A by 3 weeks for \$1 539 --- far cheaper than touching F or H.

</details>

</details>

<details>
<summary>Problem 2: Confidence in a date</summary>

A project's critical path has $\mu = 60$ days and $\sigma = 4$ days. The customer
wants delivery in 64 days.

(a) What is the probability of meeting day 64? (b) What date should you commit to
for 95 % confidence?

<details>
<summary>Solution</summary>

```{r p2-prob-08, echo=TRUE}
mu <- 60; sigma <- 4
cat(sprintf("P(finish by day 64) = %.1f%%\n", 100 * pnorm((64 - mu) / sigma)))
cat(sprintf("95%% confident date  = day %.1f\n", mu + qnorm(0.95) * sigma))
```

(a) ~84 %. (b) Day 66.6 --- commit to day 67.

</details>

</details>

<details>
<summary>Problem 3 (discussion): crash or fast-track the RC-vehicle launch?</summary>

The retailer moves the launch three weeks earlier. Design is on the critical
path and is followed by prototype assembly (also critical). You could add a
second designer and pay overtime (crash), or start building the frame jig from
the released frame drawings while the body design is still being finalized
(fast-track). Which, and why?

<details>
<summary>Discussion points</summary>

Fast-tracking the jig against the *frame* drawings is
low-cost and reasonably safe *if* the frame design is stable --- but if the body
design forces a late frame change, the jig is rework. Crashing design with a
second modeller adds cost and has coordination overhead (two people on one
model), but no rework risk. A common answer: fast-track the clearly stable parts,
crash the rest, and hold a contingency for the rework risk you accepted.

</details>

</details>

## Further Resources

```{r video-resources-08, echo=FALSE, results='asis'}
embed_youtube("6B7Qtc3rrQ4", "PERT / Three-Point Estimation Technique")
embed_youtube("4oDLMs11Exs", "Forward and Backward Pass -> Project Duration and Critical Path")
```
