---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Cost Estimation and Budgeting

```{r setup-09, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **List** the common sources of project cost and **classify** any cost against the four cost dichotomies (direct/indirect, fixed/variable, normal/expedited, recurring/non-recurring).
2. **Calculate** direct labour cost from a rate and hours, and **apply** an indirect-cost percentage.
3. **Select** an estimating method (ballpark, parametric, feasibility, definitive) appropriate to the project stage, and **state** its expected accuracy.
4. **Explain** how planning horizon, duration, people, structure, padding, and organizational culture affect estimate quality.
5. **Compare** top-down and bottom-up budgeting and **build** a bottom-up budget from a work breakdown structure using **activity-based costing**.
6. **Construct** a time-phased budget and **plot** the cumulative cost baseline (the S-curve).
7. **Size** a contingency reserve and **distinguish** it from a management reserve.

<details>
<summary>Before we begin: you have been asked for "a number" before you have any details</summary>

A stakeholder corners you: *"Ballpark it --- what will this project cost?"* You
have a one-page concept and nothing else. What do you say, and how do you say it
so the number does not follow you around for the next six months as a
commitment? Hold that thought.

</details>

## Introduction: the number that decides whether the project happens

Back to the running project. Your company must present a budget to **Senior
Management before the Scope and Charter are approved**. The project brief hands
you rough ranges to work from:

| Item | Range |
|---|---|
| Project Manager (salary) | \$60 000 -- \$80 000 |
| Team members (salary, each) | \$30 000 -- \$45 000 |
| Sundry / overhead (per hour) | \$8.00 -- \$15.00 |
| Motor / controller | \$100 000 -- \$175 000 |
| Carbon fibre | \$120 000 -- \$200 000 |
| Aluminium | \$50 000 -- \$115 000 |
| Tooling | \$100 000 -- \$180 000 |
| Testing | \$25 000 -- \$100 000 |
| Government approval (CSA, CE, MTO) | \$2 000 -- \$10 000 |
| Packaging / shipping | \$5 000 -- \$25 000 |

: Selected budget ranges from the RC-vehicle project brief

**Cost estimation is the first step in building a budget, and the budget
determines whether the project is feasible at all.** Two principles run through
everything below:

1. The more clearly you define each cost element up front, the fewer budgeting errors you make.
2. The more accurate your initial estimate, the more your budget reflects reality --- and the less likely a painful overrun later.

::: {.callout-note}
**Industry snapshot.** A defense contractor bidding a **firm-fixed-price**
contract must submit a *definitive* estimate (±5 %): if the real cost comes in
higher, the contractor eats the difference. A food company scoping a new
retort line at the concept stage lives with a *ballpark* (±30 %) and revises as
quotes arrive. Same tools, very different tolerance for error --- driven by where
in the project you are standing.
:::

```{r video-cost-mgmt-basics, echo=FALSE, results='asis'}
embed_youtube("iEndSB5IgQM", "The ABCs of Project Cost Management (Adriana Girdler)")
```

## Where project cost comes from

Cost management starts in the **planning phase**, when you identify the project's
human and material resources and lay out a **time-phased budget** that reflects
when each will be used. The common sources of cost:

- **Labour** --- the project team and anyone charging time to the project.
- **Materials** --- everything consumed to build the deliverable.
- **Subcontractors** --- external vendors, *and* internal support departments (test lab, drafting) that cross-charge.
- **Equipment and facilities** --- purchased, leased, or allocated.
- **Travel.**
- **Shipping** --- inbound and outbound, normal and expedited.

## The eight cost types

Every cost gets sorted into **one or more** of eight categories, arranged as
four either/or pairs. A single cost usually carries **two or three** labels at
once.

```{r tbl-cost-types-09}
cost_types <- data.frame(
  Dichotomy = c("Direct vs. Indirect", "", "Fixed vs. Variable", "",
                "Normal vs. Expedited", "", "Recurring vs. Non-recurring", ""),
  Type = c("Direct", "Indirect", "Fixed", "Variable",
           "Normal", "Expedited", "Recurring", "Non-recurring"),
  Meaning = c(
    "Clearly generated by the project (project labour, project materials, POs).",
    "Overhead: rent, property tax, benefits, utilities, depreciation, admin. Often applied as 20--50 % of direct cost.",
    "Does not change with usage (a 6-month equipment lease at a flat fee).",
    "Proportional to usage (machine-hour charges, fuel, mileage).",
    "Incurred during normal operation and working hours; includes overtime that was already scheduled.",
    "Unplanned cost to go faster: rush freight, extra hires, premium to a supplier for priority.",
    "Happens again through the project life: utilities, logistics, ongoing labour, monthly maintenance.",
    "One-time, usually at the start or end: training, prototype tooling, initial market analysis."
  )
)
knitr::kable(cost_types, align = c("l", "l", "l"),
             caption = "The four cost dichotomies (eight types).") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(2, bold = TRUE)
```

### Direct labour cost

The core calculation is simple:

$$\text{Total Direct Labour Cost} = \text{Direct Labour Rate} \times \text{Hours Spent}$$

Indirect cost is harder to pin to a task, so many organizations apply it as a
**percentage of direct cost**:

$$\text{Indirect Cost} = \text{Indirect Rate} \times \text{Direct Cost}, \qquad \text{Indirect Rate} \approx 20\%\text{--}50\%$$

### Worked Example 1: fully-burdened cost of a work package

The "Design body & frame" work package needs **1 mechanical designer for
220 h** at \$46/h and **1 junior engineer for 120 h** at \$32/h. Direct
materials for prototypes are \$4 500. The company applies indirect cost at
**35 %** of direct cost.

Live cell --- change the hours, rates, materials, or the indirect rate.

```{webr-r}
#| label: cost-wp-example-09
designer_hours <- 220; designer_rate <- 46
junior_hours   <- 120; junior_rate   <- 32
direct_materials <- 4500
indirect_rate <- 0.35

direct_labour <- designer_hours * designer_rate + junior_hours * junior_rate
direct_cost   <- direct_labour + direct_materials
indirect_cost <- indirect_rate * direct_cost
total_cost    <- direct_cost + indirect_cost

cat("Direct labour  :", sprintf("$%s", format(direct_labour, big.mark = " ")), "\n")
cat("Direct materials:", sprintf("$%s", format(direct_materials, big.mark = " ")), "\n")
cat("Direct cost    :", sprintf("$%s", format(direct_cost, big.mark = " ")), "\n")
cat("Indirect (35%) :", sprintf("$%s", format(indirect_cost, big.mark = " ")), "\n")
cat("-----\n")
cat("Fully-burdened work-package cost:",
    sprintf("$%s", format(round(total_cost), big.mark = " ")), "\n")
```

<details>
<summary>Practice: classify the cost (from the MGMT-3105 budgeting case study)</summary>

For each item, give **up to three** labels from: A Direct, B Indirect, C Fixed,
D Variable, E Recurring, F Non-recurring, G Normal, H Expedited.

| # | Cost item | Answer |
|---|---|---|
| 1 | Technician's hourly wages for project assembly | A Direct, D Variable, G Normal |
| 2 | Office rent for company headquarters | B Indirect, C Fixed, E Recurring |
| 3 | One-time purchase of specialized testing software | A Direct, F Non-recurring |
| 4 | Electricity used by production equipment | D Variable, E Recurring, B Indirect |
| 6 | Hiring extra workers to meet a tight deadline | H Expedited, A Direct |
| 9 | Rush shipment of components to meet schedule | H Expedited, A Direct |
| 12 | Cost of the project manager's salary | C Fixed, E Recurring, A Direct |
| 17 | Routine monthly maintenance of equipment | F ... *(trick: this is E Recurring, B Indirect)* |
| 23 | Equipment lease for six months (flat fee) | B Indirect, C Fixed, E Recurring |
| 40 | One-time cost for prototype tooling | A Direct, F Non-recurring, C Fixed |

The pattern: **anything charged straight to the project is Direct; overhead is
Indirect; anything you pay to go faster is Expedited; anything paid once is
Non-recurring.** Most items are genuinely two or three of these at once.

</details>

### The dichotomies on the running project

- **Fixed vs. variable.** The 6-month lease on the CMM is **fixed** --- the fee is the same whether you measure 50 parts or 5 000. The machine-hour charge on the rented moulding press is **variable**. Leasing an expensive machine is a real PM judgement call: will the usage justify the flat fee?
- **Normal vs. expedited.** Air-freighting circuit boards from the supplier to recover a slipped date is **expedited** and **direct**. The overtime you *planned* into the schedule from day one is still **normal**.
- **Recurring vs. non-recurring.** Prototype tooling and the initial market analysis are **non-recurring**. Utilities, logistics, and the team's salaries are **recurring** --- and it matters which is which the moment you build a time-phased budget.

## Estimating methods and their accuracy

The method you can use depends on how much you know, which depends on how far
into the project you are.

```{r tbl-estimate-methods-09}
methods <- data.frame(
  Method = c("Ballpark / order-of-magnitude", "Comparative / parametric",
             "Feasibility", "Definitive"),
  `When` = c(
    "Very early; scarce time/resources; deciding whether to even bid.",
    "A genuinely similar past project exists, with reliable archived actuals.",
    "Scope is clear and first supplier quotes are back (common in construction).",
    "Most design is complete; major POs can be written; detail is known."
  ),
  Accuracy = c("+/- 30 %", "+/- 15 %", "+/- 10 %", "+/- 5 %"),
  check.names = FALSE
)
knitr::kable(methods, align = c("l", "l", "c"),
             caption = "Estimating methods by project stage and typical accuracy.") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(3, bold = TRUE)
```

This is the **cone of uncertainty**: the possible range around an estimate is
widest at the concept stage and narrows as the project is defined.

```{r fig-cone-of-uncertainty-09, fig.cap="The cone of uncertainty. Early estimates can be off by +/- 30 %; the range narrows to +/- 5 % only once most design decisions are locked. The same $1.0M point estimate means very different things at each stage.", fig.align='center', fig.width=10, fig.height=5.5}
cone <- data.frame(
  stage = factor(c("Ballpark", "Parametric", "Feasibility", "Definitive"),
                 levels = c("Ballpark", "Parametric", "Feasibility", "Definitive")),
  x = 1:4,
  err = c(0.30, 0.15, 0.10, 0.05)
)
point_estimate <- 1.0  # $M

ggplot(cone, aes(x = x)) +
  geom_ribbon(aes(ymin = point_estimate * (1 - err),
                  ymax = point_estimate * (1 + err)),
              fill = "#5DADE2", alpha = 0.35) +
  geom_line(aes(y = point_estimate * (1 + err)), linewidth = 1, color = "#2E86C1") +
  geom_line(aes(y = point_estimate * (1 - err)), linewidth = 1, color = "#2E86C1") +
  annotate("segment", x = 0.7, xend = 4.3, y = point_estimate, yend = point_estimate,
           linetype = "dashed", color = "grey30") +
  annotate("text", x = 4.3, y = point_estimate, label = "point estimate  $1.0M",
           hjust = 1, vjust = -0.7, size = 3.3 * diagram_text_scale) +
  geom_point(aes(y = point_estimate * (1 + err)), size = 2.5, color = "#1B4F72") +
  geom_point(aes(y = point_estimate * (1 - err)), size = 2.5, color = "#1B4F72") +
  geom_text(aes(y = point_estimate * (1 + err),
                label = paste0("+", scales::percent(err, accuracy = 1))),
            vjust = -0.8, size = 3.2 * diagram_text_scale) +
  geom_text(aes(y = point_estimate * (1 - err),
                label = paste0("-", scales::percent(err, accuracy = 1))),
            vjust = 1.6, size = 3.2 * diagram_text_scale) +
  scale_x_continuous(breaks = 1:4, labels = levels(cone$stage)) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M", big.mark = " ")) +
  labs(title = "The cone of uncertainty", x = NULL, y = "Estimated project cost") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Try it yourself: parametric estimate and the accuracy band

A **parametric** estimate uses a cost-estimating relationship (CER) from history
--- for example, "past RC-vehicle projects cost about \$14.50 per assembled part
plus \$85 000 of fixed project cost." The live cell builds the estimate and
shows the band each method would put around it.

```{webr-r}
#| label: parametric-estimate-09
units             <- 120000
cost_per_unit_CER <- 14.50      # $/unit, from archived projects
fixed_project_cost<- 85000      # $, setup / PM / approvals

point_estimate <- units * cost_per_unit_CER + fixed_project_cost

bands <- data.frame(
  method   = c("Ballpark", "Parametric", "Feasibility", "Definitive"),
  accuracy = c(0.30,        0.15,         0.10,          0.05)
)
bands$low  <- round(point_estimate * (1 - bands$accuracy))
bands$high <- round(point_estimate * (1 + bands$accuracy))

cat(sprintf("Parametric point estimate: $%s\n\n",
            format(round(point_estimate), big.mark = " ")))
for (i in seq_len(nrow(bands))) {
  cat(sprintf("%-11s (+/- %2.0f%%):  $%s  ...  $%s\n",
              bands$method[i], bands$accuracy[i] * 100,
              format(bands$low[i],  big.mark = " "),
              format(bands$high[i], big.mark = " ")))
}
```

<details>
<summary>Why quote a range at all --- why not just give "the number"?</summary>

Because a single number reads as a **commitment**. If you say "\$1.8M" at the
concept stage and the definitive estimate later lands at \$2.0M, you have
"overrun by \$200k" --- even though \$2.0M was always inside the ±30 % ballpark
band. Quoting "\$1.3M--\$2.3M, ballpark" sets the expectation honestly and buys
you room to refine. A clear distinction between an **estimate** and a
**commitment** is one of the most important things a PM can establish early.

</details>

## What makes estimates good or bad

| Factor | Effect on estimate quality |
|---|---|
| **Planning horizon** | Estimates of near-term work are close to 100 % accurate; accuracy falls the further out the task sits. |
| **Project duration** | Long projects and new technology add uncertainty, especially if the scope statement is weak. |
| **People** | The closer the estimator's experience matches the work, the better. Teams that have worked together before form faster. Staff turnover hurts. |
| **Structure** | Dedicated teams deliver faster but tie up people full-time. Matrix organizations are more cost-efficient but slower. |
| **Padding** | Estimators add a safety margin. If *everyone* pads a little, the project total is badly overstated and stops reflecting reality. |
| **Organizational culture** | Some cultures reward padding; others prize accuracy. The consequences an estimator faces for a miss shape the number they give. |
| **Other** | Equipment downtime, sickness, vacation, holidays, legal limits, and project priority all leak into the estimate. |

### The cost of padding

Padding feels prudent per task. Watch what it does across a whole project.

```{r fig-padding-compounding-09, fig.cap="A 15 % 'just to be safe' pad added to every one of 40 tasks does not add 15 % to the project --- it adds 15 % to the mean, but the padded total sits far above any realistic outcome because the pads do not all get used.", fig.align='center', fig.width=10, fig.height=5}
set.seed(2025)
n_tasks <- 40
true_mean <- 40           # hours, honest most-likely per task
# Honest estimates vary around the truth; padded estimates add 15% on top.
honest  <- rnorm(n_tasks, true_mean, 6)
padded  <- honest * 1.15

df <- data.frame(
  task = 1:n_tasks,
  honest_cum = cumsum(honest),
  padded_cum = cumsum(padded)
)

ggplot(df, aes(x = task)) +
  geom_line(aes(y = honest_cum, color = "Honest estimate"), linewidth = 1.1) +
  geom_line(aes(y = padded_cum, color = "Every task padded 15%"), linewidth = 1.1) +
  annotate("text", x = n_tasks, y = df$padded_cum[n_tasks],
           label = paste0("+", round(100 * (df$padded_cum[n_tasks] /
                          df$honest_cum[n_tasks] - 1)), "% on the total"),
           hjust = 1, vjust = -0.6, size = 3.4 * diagram_text_scale) +
  scale_color_manual(values = c("Honest estimate" = "#27AE60",
                                "Every task padded 15%" = "#C0392B")) +
  labs(title = "Per-task padding compounds into a meaningless total",
       x = "Cumulative task count", y = "Cumulative estimated hours",
       color = NULL) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

The fix is not "ban all safety margin" --- it is to **remove padding from the
work packages** and hold a single, visible **contingency reserve** at the
project level (covered below).

### Guidelines for estimating (cost and time)

- **Responsibility:** the person who will do the task makes the estimate (except for highly specialised work). It uses their experience and builds their buy-in.
- **Use several estimators:** independent estimates from knowledgeable people, reconciled, remove the extremes.
- **Assume normal conditions:** normal hours, normal resource levels --- and write the assumptions down.
- **Define the time units** and the estimating base.
- **Estimate each task independently;** be careful with chains of dependent tasks, and keep one estimator per task.
- **Keep contingencies out of the work packages.** Contingency is a separate, managed fund.

## Budgeting: top-down and bottom-up

The budget and the schedule are built **in tandem** --- the budget decides
whether the schedule's milestones are reachable.

```{r tbl-topdown-bottomup-09}
approaches <- data.frame(
  Aspect = c("How it starts", "Advantages", "Disadvantages"),
  `Top-down` = c(
    "Senior management sets the overall budget and major work-package figures from experience; the numbers are pushed down the hierarchy and refined.",
    "Management aggregate estimates are often quite accurate; imposes budget discipline and cost control.",
    "Creates friction: lower levels must fit management's fixed total. Turns into a zero-sum game between functions."
  ),
  `Bottom-up` = c(
    "Start from the WBS; apply direct and indirect cost to each activity; sum up to work package, deliverable, then project.",
    "Forces detailed planning; improves PM--functional-manager communication; gives management a clear basis for prioritizing scarce resources.",
    "Reduces management's control to oversight rather than initiation; can misalign project goals with company goals; fine-tuning is time-consuming."
  ),
  check.names = FALSE
)
knitr::kable(approaches, align = c("l", "l", "l"),
             caption = "Top-down vs. bottom-up budgeting.") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(1, bold = TRUE)
```

## Activity-based costing (ABC)

ABC treats a project as a series of **activities**, each of which **consumes
resources**. Cost the activities and you have costed the project. Four steps:

1. Identify the activities that consume resources and assign costs to them.
2. Identify the **cost drivers** for each activity (personnel, materials, machine time...).
3. Compute a **cost rate per driver unit** (e.g. labour \$/hour).
4. **Assign cost** = rate × units the project uses. *(Example: an engineer at \$60/h working 80 h = \$4 800.)*

### A preliminary ABC budget

Splitting each activity into its **direct** and **overhead** components:

```{r tbl-abc-budget-09}
abc <- data.frame(
  Activity   = c("Survey", "Design", "Clear site", "Foundation", "Framing", "Plumb & wire"),
  Direct     = c(3500, 7000, 3500, 6750, 8000, 3750),
  Overhead   = c(500, 1000, 500, 750, 2000, 1250)
)
abc$Total <- abc$Direct + abc$Overhead
abc <- rbind(abc, data.frame(Activity = "TOTAL",
                             Direct = sum(abc$Direct),
                             Overhead = sum(abc$Overhead),
                             Total = sum(abc$Total)))

knitr::kable(abc, align = c("l", "r", "r", "r"),
             format.args = list(big.mark = " "),
             caption = "Preliminary ABC budget: direct vs. overhead by activity.") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::row_spec(nrow(abc), bold = TRUE)
```

### Try it yourself: cost an activity, then a variance

The live cell costs a single activity from its drivers, then compares planned
vs. actual to get the **variance**. Change the rates, hours, or actuals.

```{webr-r}
#| label: abc-activity-variance-09
# --- Step 3 & 4: cost the "Assemble & test prototype" activity from drivers ---
labour_rate   <- 38;  labour_hours   <- 160
machine_rate  <- 22;  machine_hours  <- 45
materials     <- 2600
overhead_rate <- 0.30            # 30% of direct

direct   <- labour_rate * labour_hours + machine_rate * machine_hours + materials
overhead <- overhead_rate * direct
planned  <- direct + overhead

cat(sprintf("Direct   : $%s\n", format(round(direct),   big.mark = " ")))
cat(sprintf("Overhead : $%s\n", format(round(overhead), big.mark = " ")))
cat(sprintf("Planned activity cost: $%s\n\n", format(round(planned), big.mark = " ")))

# --- Variance: what actually happened ---
actual <- 11800
variance <- planned - actual          # positive = under budget (favourable)
pct <- 100 * variance / planned

verdict <- if (variance >= 0) "UNDER budget (favourable)" else "OVER budget (unfavourable)"
cat(sprintf("Actual   : $%s\n", format(actual, big.mark = " ")))
cat(sprintf("Variance : $%s  (%.1f%%)  -> %s\n",
            format(round(variance), big.mark = " "), pct, verdict))
```

::: {.callout-warning}
**Common mistake.** Contingency is **not** built into an ABC budget. ABC costs
the activities as they are planned to run; the contingency reserve is added
*afterward*, on top of the total.
:::

## The time-phased budget and the S-curve

A **time-phased budget** spreads each activity's cost across the calendar
periods when it will actually be spent. Summing down the columns gives
**monthly planned spend**; the running total is the **cumulative cost
baseline** --- the S-curve you will track the project against in
[Chapter 12](12-Project-Control-Systems.qmd).

```{r fig-s-curve-09, fig.cap="Time-phased budget for a six-activity project. Bars are planned spend per month; the line is the cumulative baseline (the S-curve). Spend is slow at first, steepest mid-project, and tapers at the end.", fig.align='center', fig.width=10, fig.height=6}
phased <- data.frame(
  month = factor(c("Jan", "Feb", "Mar", "Apr", "May"),
                 levels = c("Jan", "Feb", "Mar", "Apr", "May")),
  Survey       = c(4000, 3000,     0,     0,    0),
  Design       = c(   0, 5000,  8000,     0,    0),
  ClearSite    = c(   0, 1000,  4000,  2000,    0),
  Foundation   = c(   0,    0,  3000,  4000,    0),
  Framing      = c(   0,    0,     0,  3000, 4000),
  PlumbWire    = c(   0,    0,     0,     0, 2000)
)
phased$monthly <- rowSums(phased[, -1])
phased$cumulative <- cumsum(phased$monthly)

scale_factor <- max(phased$cumulative) / max(phased$monthly)

ggplot(phased, aes(x = month)) +
  geom_col(aes(y = monthly), fill = "#5DADE2", width = 0.6) +
  geom_line(aes(y = cumulative / scale_factor, group = 1),
            color = "#C0392B", linewidth = 1.2) +
  geom_point(aes(y = cumulative / scale_factor), color = "#C0392B", size = 3) +
  geom_text(aes(y = cumulative / scale_factor,
                label = scales::dollar(cumulative, big.mark = " ")),
            vjust = -1, size = 3.2 * diagram_text_scale, color = "#C0392B") +
  scale_y_continuous(
    name = "Planned spend this month",
    labels = scales::dollar_format(big.mark = " "),
    sec.axis = sec_axis(~ . * scale_factor, name = "Cumulative baseline (S-curve)",
                        labels = scales::dollar_format(big.mark = " "))
  ) +
  labs(title = "Time-phased budget and cumulative cost baseline", x = NULL) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

Matching the cost baseline to the schedule baseline is what lets you set
**milestones for both schedule and budget** and later tell whether a project
that has spent 60 % of its money is 60 %, 40 %, or 80 % done.

## Contingency reserves

A **budget contingency** is extra money set aside to cover the uncertainty that
every estimate carries --- because even the best budget is only an estimate and
something unforeseen always happens. Key points:

- Added **after all costs are identified**, as a cushion on top of the calculated total (commonly **10--15 %**; the MGMT-3105 Costs Chart uses **15 %**).
- **Not** included in an ABC budget --- it sits on top.
- The **PM controls a project contingency fund**; separately, **top management holds a management reserve** for genuinely unknown-unknown events.

Why stakeholders push back: the percentage can look arbitrary, or feel like
being asked to pay for the vendor's poor cost control. Decide early --- and write
into the scope statement --- whether contingency applies across **all** tasks or
is held for the tasks **critical** to project success.

```{r tbl-reserve-vs-reserve-09}
reserves <- data.frame(
  Feature = c("Covers", "Owned by", "Part of the cost baseline?", "Typical size"),
  `Contingency reserve` = c(
    "Identified risks / \"known-unknowns\" (small scope adjustments, iteration, Murphy's Law).",
    "Project manager.",
    "Yes --- inside the baseline.",
    "10--15 % of estimated cost (project-dependent)."
  ),
  `Management reserve` = c(
    "Unidentified risks / \"unknown-unknowns.\"",
    "Senior management / sponsor.",
    "No --- outside the baseline; released by management.",
    "Set by organizational policy."
  ),
  check.names = FALSE
)
knitr::kable(reserves, align = c("l", "l", "l"),
             caption = "Contingency reserve vs. management reserve.") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) |>
  kableExtra::column_spec(1, bold = TRUE)
```

### Try it yourself: build the RC-vehicle project budget

The live cell builds a rough **bottom-up** budget from the project brief's
ranges (using a pick inside each range), rolls up labour and materials, and adds
a contingency. Change the picks or the contingency rate and watch the total ---
then check it against what Senior Management is likely to approve.

```{webr-r}
#| label: rc-project-budget-09
# --- Labour (annualised salaries prorated to a 7-month project = 7/12) ---
months <- 7
pm_salary        <- 72000
n_team           <- 4
team_salary_each <- 38000
sundry_per_hour  <- 11
sundry_hours     <- 1800          # total support hours booked to the project

labour <- (pm_salary + n_team * team_salary_each) * months / 12 +
          sundry_per_hour * sundry_hours

# --- Materials (one pick inside each brief range) ---
materials <- c(
  chassis_casting   = 55000,
  fasteners         = 9000,
  motor_controller  = 140000,
  carbon_fibre      = 0,          # not selected this time
  aluminium         = 85000,
  welding           = 25000,
  circuit_boards    = 32000,
  gov_approval      = 6000,
  testing           = 60000,
  tooling           = 150000,
  safety_equipment  = 50000,
  packaging_shipping= 15000
)
material_total <- sum(materials)

subtotal      <- labour + material_total
contingency_r <- 0.15
contingency   <- contingency_r * subtotal
grand_total   <- subtotal + contingency

cat(sprintf("Labour subtotal      : $%s\n", format(round(labour),        big.mark = " ")))
cat(sprintf("Material subtotal     : $%s\n", format(round(material_total),big.mark = " ")))
cat(sprintf("Subtotal             : $%s\n", format(round(subtotal),      big.mark = " ")))
cat(sprintf("Contingency (%.0f%%)     : $%s\n", contingency_r * 100,
            format(round(contingency), big.mark = " ")))
cat(sprintf("--------\nGRAND TOTAL          : $%s\n",
            format(round(grand_total), big.mark = " ")))
```

<details>
<summary>How would top-down vs. bottom-up change this number?</summary>

The cell above is **bottom-up**: it sums the pieces and lets the total fall out.
A **top-down** pass would start with a figure Senior Management is willing to
fund --- say "\$650k, no more" --- and force each line to fit. In practice you do
both: build bottom-up for defensibility, compare to the top-down ceiling, and
negotiate the gap before the Charter is signed. If bottom-up comes in well over
the ceiling, that is a **feasibility** finding, not a rounding problem.

</details>

## Key Takeaways

- Cost management starts in **planning**. Sources: labour, materials, subcontractors, equipment/facilities, travel, shipping.
- Every cost carries **two or three** of the eight labels: direct/indirect, fixed/variable, normal/expedited, recurring/non-recurring. **Direct labour cost = rate × hours**; indirect is often **20--50 % of direct**.
- Estimating method follows project stage: **ballpark ±30 %, parametric ±15 %, feasibility ±10 %, definitive ±5 %** --- the **cone of uncertainty**. Always quote a **range**, and keep "estimate" separate from "commitment."
- Estimate quality is driven by planning horizon, duration, people, structure, culture, and **padding** --- which compounds into a meaningless total. Remove pads from work packages; hold one visible contingency.
- **Top-down** imposes discipline but breeds friction; **bottom-up** (via **activity-based costing** on the WBS) forces real planning. ABC = identify activities → cost drivers → rate per unit → rate × units.
- A **time-phased budget** produces the **cumulative cost baseline (S-curve)** you track against later.
- **Contingency reserve** (10--15 %, PM-owned, inside the baseline, covers known-unknowns) is distinct from the **management reserve** (sponsor-owned, outside the baseline, covers unknown-unknowns). Contingency is added **after** ABC, never inside it.

## Case Study: Meridian R/C — Week 9

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 9: The Bottom-Up Budget Goes to Daniel Cho"}

Six weeks after Daniel Cho signed off on the $800,000 charter ceiling, you owe
him something more defensible than a ballpark: a definitive, bottom-up budget
built activity by activity off the WBS you cut with Priya Nandakumar, Marcus
Webb, and Sofia Reyes back in Week 5. This week you assembled it, work
package by work package, and sat down with Daniel for sign-off.

You started with labour, costing each team member's time against the
prototype phase: yourself as PM at $72,000, Priya (tooling) at $42,000,
Marcus (procurement) at $38,000, and Sofia (quality and test) at $36,000 —
a labour subtotal of **$188,000**. Then you rolled up the direct-cost line
items Marcus, Priya, and Sofia had been pricing with suppliers all along: the
motor and controller package MCS had quoted at $145,000, the aluminium for
the body and frame (the Week 1 material decision) at $95,000, the body and
frame tooling dies at $150,000, testing and certification fees at $60,000,
government and toy-safety approval fees at $6,000, and packaging and shipping
for the client sample run at $18,000 — a direct-cost subtotal of
**$474,000**.

Labour plus direct costs gives total direct costs of **$662,000**. Following
the course's Costs Chart convention — and remembering that contingency is
never buried inside an activity-based budget, only added on top once every
line item is identified — you applied a **15% contingency on the full
$662,000 direct-cost total**, not just the material-and-vendor subtotal. That
adds **$99,300**, for a **Budget at Completion (BAC) of $761,300**.

Daniel did the arithmetic you knew he would: $761,300 against his $800,000
ceiling leaves **$38,700 of headroom** — about 4.8% of the ceiling. Comfortable,
but not so much that he waved it through without questions. He pressed
hardest on the $150,000 tooling line (largest single item after labour) and
on why testing and certification, at $60,000, was priced near the top of the
brief's original $25,000–$100,000 range rather than the middle. You walked him
through the WBS lineage of each number — 1.2 Tooling & Fabrication, 1.5
Testing & Certification — and he signed off. The bottom-up number now becomes
the baseline you will track against once the S-curve is time-phased and,
later, once Earned Value tracking begins in Week 12.

:::

## Practice Problems

<details>
<summary>Problem 1: Fully-burdened labour and an indirect-rate swing</summary>

A test campaign needs 2 technicians for 3 weeks (37.5 h/week each) at \$34/h,
plus \$6 200 of consumables. Compute the fully-burdened cost at an indirect rate
of 25 %, and again at 45 %. How much does the indirect-rate assumption alone
move the number?

**Solution.**

```{r p1-burdened-09, echo=TRUE}
techs <- 2; weeks <- 3; hrs_week <- 37.5; rate <- 34
consumables <- 6200

direct_labour <- techs * weeks * hrs_week * rate
direct_cost   <- direct_labour + consumables

for (ir in c(0.25, 0.45)) {
  total <- direct_cost * (1 + ir)
  cat(sprintf("Indirect %2.0f%%:  direct $%s  +  indirect $%s  =  $%s\n",
              ir * 100,
              format(direct_cost, big.mark = " "),
              format(round(direct_cost * ir), big.mark = " "),
              format(round(total), big.mark = " ")))
}
cat(sprintf("\nSwing from the indirect-rate assumption alone: $%s\n",
            format(round(direct_cost * (0.45 - 0.25)), big.mark = " ")))
```

The direct cost is fixed at \$21 500, but the burdened total moves by \$4 300 ---
20 % of direct --- purely on the overhead assumption. Always state the indirect
rate you used.

</details>

<details>
<summary>Problem 2: Which estimating method, and what range?</summary>

For each situation, name the method and give the ± band around a \$500 000 point
estimate.

(a) A retailer asks, in a hallway, roughly what a new product line will cost.
You have a concept sketch.
(b) You have run three near-identical RC-vehicle launches in the last four years
with clean cost records.
(c) All major POs are drafted and 90 % of the design is released for a
firm-fixed-price defense bid.

**Solution.**

- (a) **Ballpark**, ±30 % → \$350 000 -- \$650 000.
- (b) **Comparative/parametric**, ±15 % → \$425 000 -- \$575 000.
- (c) **Definitive**, ±5 % → \$475 000 -- \$525 000. (Anything looser is a bad idea on a fixed-price bid.)

</details>

<details>
<summary>Problem 3: Variance analysis on the ABC budget</summary>

Using the preliminary ABC budget above (total planned \$63 250), the project
finishes with these actuals: Survey \$4 250, Design \$8 000, Clear site \$3 500,
Foundation \$8 500, Framing \$11 250, Plumb & wire \$5 150. Find the variance per
activity and overall, and flag the activities that need a post-mortem.

**Solution.**

```{r p3-variance-09, echo=TRUE}
planned <- c(Survey = 4000, Design = 8000, `Clear site` = 4000,
             Foundation = 7500, Framing = 10000, `Plumb & wire` = 5000)
actual  <- c(4250, 8000, 3500, 8500, 11250, 5150)

variance <- planned - actual            # + = under budget
tab <- data.frame(Activity = names(planned),
                  Planned = as.integer(planned),
                  Actual = as.integer(actual),
                  Variance = as.integer(variance),
                  Pct = round(100 * variance / planned, 1))
print(tab, row.names = FALSE)

cat(sprintf("\nTOTAL planned $%s  actual $%s  variance $%s (%.1f%%)\n",
            format(sum(planned), big.mark = " "),
            format(sum(actual),  big.mark = " "),
            format(sum(variance), big.mark = " "),
            100 * sum(variance) / sum(planned)))
```

Design and (roughly) Survey held. **Foundation (-13 %)** and **Framing (-13 %)**
blew through and drove the whole project \$2 150 (3.4 %) over --- those two get
the post-mortem.

</details>

<details>
<summary>Problem 4 (discussion): sizing contingency for the RC-vehicle project</summary>

The project brief flags several live risks: one candidate motor supplier is
unproven, that supplier's staff are negotiating a first labour contract, the
carbon-fibre supplier is mid-restructuring after an acquisition, and a new
assembly line would carry a learning curve. The subtotal budget is \$620 000.

Would you size the contingency at 10 %, 15 %, or 20 %? Would you spread it
across all tasks or ring-fence it for specific ones? Defend your answer.

**Discussion points.** With that many concentrated, plausible risks, **10 % is
light**; 15 % is defensible and matches the course's Costs Chart; 20 % is
arguable if the unproven-supplier and new-line risks are both realised. A strong
answer **ring-fences** most of the reserve against the supplier and
assembly-line line items (the tasks critical to success) rather than smearing it
evenly, and states that split in the scope statement so stakeholders are not
surprised when it is drawn down.

</details>

## Further Resources

- **Cost estimating techniques (analogous, parametric, bottom-up):** curated explainer videos below.
- Preview of [Chapter 12 --- Project Control Systems](12-Project-Control-Systems.qmd): the S-curve you built here becomes the baseline for **earned value** analysis.

```{r video-resources-09, echo=FALSE, results='asis'}
embed_youtube("EyPFi0YO32M", "The Basics of Project Cost Management (ProjectManager)")
embed_youtube("hfn9VCAj4Dw", "Cost Estimates - Key Concepts in Project Management (David McLachlan)")
embed_youtube("2IxoWVjsphc", "How to Estimate Project Costs Accurately")
embed_youtube("x0EgZnXdunw", "What Is Earned Value? (Ricardo Vargas) - looking ahead to project control")
```
