---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Negotiation and Conflict Management

```{r setup-10, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Identify** what a project manager negotiates and with whom.
2. **Prepare** for a negotiation: interests vs. positions, BATNA, reservation point, ZOPA.
3. **Apply** the principled-negotiation mindset (separate people from problem; focus on interests; win-win or no win).
4. **Determine** whether a zone of possible agreement exists and how wide it is.
5. **Classify** project conflict (goal-oriented, administrative, interpersonal) and its organizational vs. interpersonal causes.
6. **Select** a conflict-resolution method (confronting, compromising, withdrawal, smoothing, forcing) for a situation.
7. **Score** a multi-criteria supplier choice.

<details>
<summary>Before we begin: recall a negotiation you "won" but regretted</summary>

Did you win the point and lose the relationship? Most project negotiations are
with people you will work with again next week. That constraint changes
everything.

</details>

## Introduction: the project manager negotiates every day

As a project manager your day is full of negotiation, inside and outside the
organization. You negotiate for or with:

- **Resources** --- material, time, equipment, money
- **Contractors** --- costs, delivery schedules, quality of material, hours of work
- **Department managers** --- manpower, schedules

On the RC-vehicle project the sharp negotiations are obvious: the **motor and
control-system supply contract** (trusted MCS, now moving production to Mexico,
vs. unproven Paragon four hours away with staff negotiating a first labour
contract), and the recurring fight with **functional managers** over which
engineers you get and for how long.

::: {.callout-note}
**Industry snapshot.** An automotive buyer negotiating a multi-year harness
contract is really negotiating the *whole package* --- price, tooling ownership,
capacity guarantees, engineering support, payment terms. A defense subcontract
negotiation is bounded by regulation (allowable cost, audit rights). In a food
company, the PM negotiating a sanitation window with operations is negotiating
*downtime*, which is money.
:::

```{r video-negotiation-10, echo=FALSE, results='asis'}
embed_youtube("MTnQX11Xey0", "Project Management Negotiations: ZOPA, BATNA, and Win-Win (Parallel Project Training)")
```

## The negotiation mindset

- **Separate the people from the problem.** Everyone has ego, fears, insecurities --- do not tap into the other side's, or you lose control of both sides.
- **Stick to business.** Keep personalities out of it; a charm offensive fools no one.
- **Put yourself in their shoes.** Nobody likes a direct attack; it is counter-productive.
- **Win-win or no win.** Talk money **last** if possible; a negotiation is about the whole package, not just the number.

## Getting ready

There are three stages --- **before, during, after** --- and **before should be the
biggest.**

**Gather information about the other side:** their vital interests, goals,
costs, strategies, positions, needs; their product strengths; concession
patterns; past negotiating style; their **bottom line**; corporate deadlines and
time constraints; their **budget**; geographic limits. Also know your own
position, the players, and **who holds the power** --- including people behind the
scenes.

**Develop strategy:** aim high (with more than one win-win scenario in mind); be
realistic; rank your objectives by priority; know your **bottom line**; decide
whether deadlock or termination are tolerable outcomes; separate the
**necessary** items from the **expendable**; have a few **red herrings** ready;
plan how to help the other side sell the deal internally; find where your
interests and theirs intersect.

**Write the plan and an agenda:** your interests and theirs; your desired
outcomes and your estimate of theirs; your authority limits and theirs; any
fact sheet either side needs. The agenda names the meeting's goal, is published
in advance, and lists invitees and location.

### The three questions --- and who controls the room

Every negotiator should ask: **How much power do I have? What time pressure is
there? Do I trust the other side?**

And: *who controls a conversation?* **Generally the person asking the questions**,
because they set the direction. Aim to ask most of the questions --- politely ---
and recognise when it is being done to you.

## BATNA, reservation point, and ZOPA

- **BATNA** --- your Best Alternative To a Negotiated Agreement. What you will do if this deal falls through. A strong BATNA is your real source of power.
- **Reservation point** --- the worst deal you would still accept (a buyer's *maximum* price, a seller's *minimum* price), set by your BATNA.
- **ZOPA** --- the Zone Of Possible Agreement: the overlap between the two reservation points. If there is no overlap, **no deal is possible** at those terms.

### Worked Example 1: is there a ZOPA?

You are buying the motor/controller package. Your BATNA (Paragon's quote plus
the cost of the risk it carries) puts your **reservation price at \$168 000** ---
you would not pay MCS more than that. MCS's floor, given their costs and other
customers, is **\$150 000**.

Live cell --- move either side's reservation price and see whether a deal still exists.

```{r calc-zopa-example-10, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "zopa-example-10",
  title = "ZOPA Calculator",
  icon = "handshake",
  inputs = list(
    qwebr_calc_input("buyer_reservation", "Buyer's reservation price (most they'll pay)", value = 168000, min = 50000, max = 400000, step = 1000, prefix = "$"),
    qwebr_calc_input("seller_reservation", "Seller's reservation price (least they'll accept)", value = 150000, min = 50000, max = 400000, step = 1000, prefix = "$")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: zopa-example-10
buyer_reservation  <- 168000   # most you will pay
seller_reservation <- 150000   # least they will accept

zopa <- seller_reservation <= buyer_reservation

money <- function(x) paste0(ifelse(x < 0, "-", ""), "$", format(round(abs(x)), big.mark = " ", scientific = FALSE, trim = TRUE))

lo  <- min(buyer_reservation, seller_reservation)
hi  <- max(buyer_reservation, seller_reservation)
pad <- max(hi - lo, 20000) * 0.5
ticks <- pretty(c(lo - pad, hi + pad), n = 5)
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
plot(NA, xlim = c(lo - pad, hi + pad), ylim = c(0, 1), xaxt = "n", yaxt = "n",
     xlab = "Price ($)", ylab = "", main = "Zone of possible agreement")
rect(lo, 0.3, hi, 0.7, col = if (zopa) "#a9dfbf" else "#f5b7b1", border = NA)
axis(1, at = ticks, labels = format(ticks, big.mark = " ", scientific = FALSE, trim = TRUE))
points(seller_reservation, 0.5, pch = 19, cex = 3, col = "#2980b9")
points(buyer_reservation, 0.5, pch = 19, cex = 3, col = "#e67e22")
text(seller_reservation, 0.85, "Seller floor", col = "#2980b9", cex = 1.2)
text(buyer_reservation, 0.15, "Buyer ceiling", col = "#e67e22", cex = 1.2)

if (zopa) {
  cat(sprintf("ZOPA exists: %s to %s\n", money(seller_reservation), money(buyer_reservation)))
  cat(sprintf("ZOPA width  %s\n", money(buyer_reservation - seller_reservation)))
} else {
  cat("!!No ZOPA!!: the seller's floor is above the buyer's ceiling\n")
  cat(sprintf("Gap between the two sides  %s\n", money(seller_reservation - buyer_reservation)))
}
```

</details>

### Try it yourself: reservation points from a BATNA

```{r calc-zopa-calc-10, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "zopa-calc-10",
  title = "Reservation Points from a BATNA",
  icon = "scale-balanced",
  inputs = list(
    qwebr_calc_input("paragon_price", "BATNA: Paragon's price", value = 150000, min = 50000, max = 400000, step = 1000, prefix = "$"),
    qwebr_calc_input("paragon_risk", "Extra risk cost of using Paragon", value = 25000, min = 0, max = 150000, step = 1000, prefix = "$"),
    qwebr_calc_input("mcs_cost", "MCS cost to make", value = 132000, min = 50000, max = 400000, step = 1000, prefix = "$"),
    qwebr_calc_input("mcs_margin", "MCS minimum margin", value = 15000, min = 0, max = 100000, step = 1000, prefix = "$")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: zopa-calc-10
# Buyer's side: BATNA is Paragon at $150k, but carrying $25k of extra risk cost.
paragon_price   <- 150000
paragon_risk    <- 25000
buyer_reservation <- paragon_price + paragon_risk   # indifference point vs MCS

# Seller's side (MCS): cost to make + minimum acceptable margin.
mcs_cost   <- 132000
mcs_margin <- 15000
seller_reservation <- mcs_cost + mcs_margin

zopa <- seller_reservation <= buyer_reservation

money <- function(x) paste0(ifelse(x < 0, "-", ""), "$", format(round(abs(x)), big.mark = " ", scientific = FALSE, trim = TRUE))

lo  <- min(buyer_reservation, seller_reservation)
hi  <- max(buyer_reservation, seller_reservation)
pad <- max(hi - lo, 20000) * 0.5
ticks <- pretty(c(lo - pad, hi + pad), n = 5)
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
plot(NA, xlim = c(lo - pad, hi + pad), ylim = c(0, 1), xaxt = "n", yaxt = "n",
     xlab = "Price ($)", ylab = "", main = "Zone of possible agreement")
rect(lo, 0.3, hi, 0.7, col = if (zopa) "#a9dfbf" else "#f5b7b1", border = NA)
axis(1, at = ticks, labels = format(ticks, big.mark = " ", scientific = FALSE, trim = TRUE))
points(seller_reservation, 0.5, pch = 19, cex = 3, col = "#2980b9")
points(buyer_reservation, 0.5, pch = 19, cex = 3, col = "#e67e22")
text(seller_reservation, 0.85, "Seller floor", col = "#2980b9", cex = 1.2)
text(buyer_reservation, 0.15, "Buyer ceiling", col = "#e67e22", cex = 1.2)

if (zopa) {
  cat(sprintf("ZOPA width: %s: a deal is possible\n", money(buyer_reservation - seller_reservation)))
} else {
  cat("!!No ZOPA!!: improve your BATNA, expand the package, or walk away\n")
}
cat(sprintf("Buyer reservation (max)  %s\n", money(buyer_reservation)))
cat(sprintf("Seller reservation (min)  %s\n", money(seller_reservation)))
if (zopa) cat(sprintf("Midpoint settlement  about %s\n", money((buyer_reservation + seller_reservation) / 2)))
```

</details>

<details>
<summary>Why does improving your BATNA move the whole negotiation?</summary>

Your reservation point is set by your BATNA. If you pre-qualify a credible
second supplier (a better BATNA), your reservation price drops --- you can
credibly walk at a lower number --- which shifts the ZOPA and the likely
settlement in your favour, *without* saying a word at the table. The best
negotiating move often happens before the meeting, in the form of a real
alternative.

</details>

### Try it yourself: reservation points and the ZOPA by hand

The buyer's reservation price comes from their BATNA, the seller's from their
costs. Work out both and the width of the zone of possible agreement. Each
team picks its own **data set**.

```{r tryit-zopa-10, echo=FALSE, results='asis'}
zp_sets <- list(
  c(alt = 150000, risk = 25000, cost = 132000, margin = 15000),
  c(alt = 90000,  risk = 8000,  cost = 85000,  margin = 12000),
  c(alt = 210000, risk = 15000, cost = 205000, margin = 30000),
  c(alt = 60000,  risk = 12000, cost = 50000,  margin = 5000))
zp_given <- function(v) c(
  "Buyer's BATNA: alternative supplier's price" = sprintf("$%s", format(v[["alt"]], big.mark = " ")),
  "Extra risk cost of using that alternative" = sprintf("$%s", format(v[["risk"]], big.mark = " ")),
  "Seller's cost to make" = sprintf("$%s", format(v[["cost"]], big.mark = " ")),
  "Seller's minimum margin" = sprintf("$%s", format(v[["margin"]], big.mark = " ")))
zp_answers <- function(v) {
  b <- v[["alt"]] + v[["risk"]]; s <- v[["cost"]] + v[["margin"]]
  list(tryit_answer("Buyer's reservation price (most they will pay)", b, tol = 1, prefix = "$"),
       tryit_answer("Seller's reservation price (least they will accept)", s, tol = 1, prefix = "$"),
       tryit_answer("ZOPA width (negative means there is no ZOPA)", b - s, tol = 1, prefix = "$"))
}
tryit_number(
  "zopa-by-hand", "Is there a zone of possible agreement?",
  answers = zp_answers(zp_sets[[1]]), given = zp_given(zp_sets[[1]]),
  teams = lapply(zp_sets[-1], function(v) list(given = zp_given(v), answers = zp_answers(v))),
  intro = "Reservation price = BATNA price plus its risk cost (buyer); cost plus minimum margin (seller).")

tryit_number(
  "zopa-twin", "Zone of possible agreement (graded)", show = FALSE,
  answers = zp_answers(c(alt = 300000, risk = 20000, cost = 270000, margin = 40000)),
  question = paste0("<p>You are buying a tooling package. Your BATNA is an alternative supplier at ",
    "$300 000, which would carry an extra $20 000 of risk cost. The seller's cost to make it is $270 000 ",
    "and they need at least a $40 000 margin. Calculate your reservation price, the seller's reservation ",
    "price, and the width of the ZOPA (enter a negative number if there is no ZOPA).</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

Buyer's reservation price = BATNA price + risk cost; seller's = cost to make +
margin. Data set 1: buyer \$175 000, seller \$147 000, so the ZOPA is
\$147 000 to \$175 000, width **\$28 000**. In data set 3 the seller's floor
(\$235 000) is above the buyer's ceiling (\$225 000): width **-\$10 000**, no
ZOPA at these terms; the buyer must improve their BATNA or change the package.

</details>

## Language and tells

A few sample moves: *"My colleagues and I cannot settle for that"* (establish
power early); *"What are you looking for in these negotiations?"* (make them open
first); *"Do you have the authority to commit to that?"* (probe their authority);
*"Let's set that aside and come back to it"* (avoid a hard no).

And listen past the words. *"To be brutally honest..."* and *"To tell you the
truth..."* often precede a lie. *"Off the record..."* --- nothing is off the
record; the negotiation is always on. Watch **congruence**: when words and body
language disagree, believe the body language.

## Conflict management

Conflict *"begins when you perceive that someone has frustrated, or is about to
frustrate, a major concern of yours."* It is **perceptual** (it doesn't matter
whether a wrong was actually done, only that it is perceived) and **dynamic**
(the original cause may no longer matter, yet the conflict persists).

Three types on projects:

- **Goal-oriented** --- disagreement over end results, scope outcomes, specs, priorities. Fuelled by vague goals that let people fill in their own interpretation (and hidden agendas).
- **Administrative** --- who has authority or control over functions, tasks, and decisions. The **matrix** structure breeds this --- everyone has two bosses.
- **Interpersonal** --- personality differences: work ethic, behavioural style, ego, personalities that don't mesh.

### Try it yourself: what type of conflict is it?

```{r tryit-conflict-types-10, echo=FALSE, results='asis'}
ct_bins <- c("Goal-oriented", "Administrative", "Interpersonal")
tryit_sort(
  "conflict-types", "Goal-oriented, administrative or interpersonal?",
  bins = ct_bins,
  cards = data.frame(
    text = c("Marketing and engineering disagree on what the finished prototype must demonstrate.",
             "Who signs off a design change: the PM or the functional manager?",
             "Two engineers have clashed on every project since a personal falling-out.",
             "Vague specifications let each side read the requirement its own way.",
             "A matrix employee takes instructions from two bosses who contradict each other.",
             "One teammate's blunt style keeps rubbing another the wrong way.",
             "The customer wants it lighter and the sponsor wants it cheaper.",
             "A disagreement over who controls the test-lab schedule."),
    bin = c("Goal-oriented", "Administrative", "Interpersonal", "Goal-oriented", "Administrative",
            "Interpersonal", "Goal-oriented", "Administrative"),
    why = c("Disagreement over end results.", "Authority over decisions: administrative.",
            "Personality and history: interpersonal.", "Vague goals let people fill in their own meaning.",
            "The two-bosses problem of the matrix: administrative.", "Style clash: interpersonal.",
            "Priorities conflict over end results.", "Authority and control over a function."),
    stringsAsFactors = FALSE),
  intro = "Remember: personality is the first diagnosis and usually the wrong one.", question = NULL)

tryit_sort(
  "conflict-types-twin", "Type of conflict (graded)", bins = ct_bins, show = FALSE,
  cards = data.frame(
    text = c("Quality and production disagree about the acceptance criteria.",
             "Two managers both claim the right to assign the test technician.",
             "A designer and a buyer simply do not get along.",
             "Nobody agrees what 'done' means for the first article.",
             "It is unclear whether the PM or the department head approves overtime."),
    bin = c("Goal-oriented", "Administrative", "Interpersonal", "Goal-oriented", "Administrative"),
    stringsAsFactors = FALSE),
  question = "<p>For each conflict, choose whether it is goal-oriented, administrative or interpersonal.</p>")
```

### Three views of conflict

| School | View | Response |
|---|---|---|
| **Traditional** | All conflict is bad | Suppress and eliminate it, fast |
| **Behavioral** | Conflict is natural and inevitable | Accept it and manage it well |
| **Interactionist** | A *controlled* amount of conflict is *beneficial* | Encourage some --- it prevents stagnation and drives innovation; too little = inertia, too much = chaos |

### Sources, ranked by intensity

```{r fig-conflict-sources-10, fig.cap="Sources of project conflict ranked by typical intensity (from PMBOK data). Schedules, priorities, and resources dominate; personality is last -- it is usually the first thing blamed and rarely the real cause.", fig.align='center', fig.width=9.5, fig.height=5}
cs <- data.frame(
  source = c("Schedules", "Project priorities", "Resources", "Technical opinions",
             "Administrative procedures", "Cost", "Personality"),
  intensity = 7:1
)
cs$source <- factor(cs$source, levels = rev(cs$source))
ggplot(cs, aes(intensity, source)) +
  geom_col(fill = "#5DADE2", width = 0.68) +
  geom_text(aes(label = intensity), hjust = -0.4, size = 3.6 * diagram_text_scale) +
  scale_x_continuous(limits = c(0, 8), expand = expansion(mult = c(0, 0.02))) +
  labs(title = "What project teams actually fight about",
       x = "Relative intensity (7 = highest)", y = NULL) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

**Organizational causes:** reward systems (functional managers hoard their best
people), scarce resources, uncertainty (weak matrix --- unclear who is in
charge), different mindsets among departments. **Interpersonal causes:**
misconceptions about others' behaviour, faulty communication (often *style*, not
content), personal grudges.

> *"It only becomes personal when the root cause of the conflict is not
> resolved. Personality is very frequently the first diagnosis"* --- and usually
> the wrong one.

### Conflict-resolution methods (PMBOK)

| Method | What it is | Best when |
|---|---|---|
| **Confronting** (problem-solving) | Solve the real, underlying problem | There is time and goodwill; the durable fix |
| **Compromising** | Each side gives something; partial satisfaction for both | Positions are far apart and a deadline looms |
| **Withdrawal / avoidance** | Retreat or postpone | The issue is trivial, or emotions need to cool |
| **Smoothing** | Emphasize agreement, play down differences | Preserving the relationship matters more right now |
| **Forcing** | Push one view at the other's expense | A true emergency, or a matter of safety/ethics |

### Try it yourself: choose the resolution method

```{r tryit-resolution-10, echo=FALSE, results='asis'}
res_bins <- c("Confronting", "Compromising", "Withdrawal", "Smoothing", "Forcing")
tryit_sort(
  "conflict-methods", "Which resolution method fits each situation?",
  bins = res_bins,
  cards = data.frame(
    text = c("A disagreement over a spec detail, with plenty of time and goodwill: solve the real problem.",
             "The two sides are far apart and the gate review is tomorrow: each gives something.",
             "A trivial scheduling grumble, with tempers high: postpone it.",
             "Preserving a key relationship matters more than winning this point today.",
             "A safety interlock has been bypassed on the line: stop it now.",
             "Repeated hand-off errors need a joint working session on the root cause.",
             "Time is short, so split the difference between the two delivery dates.",
             "Emphasize what the two teams agree on and play down the disagreement for now."),
    bin = c("Confronting", "Compromising", "Withdrawal", "Smoothing", "Forcing", "Confronting",
            "Compromising", "Smoothing"),
    why = c("Confronting solves the underlying problem; it needs time and goodwill.",
            "Positions far apart and a deadline looming: compromise.",
            "Trivial issue or emotions need to cool: withdraw.",
            "Smoothing protects the relationship.", "Forcing is for a true emergency or a safety or ethics matter.",
            "The durable fix is to confront the real problem together.",
            "Each side gives something: compromise.", "Emphasizing agreement is smoothing."),
    stringsAsFactors = FALSE),
  intro = "Pick the method that fits best.", question = NULL)

tryit_sort(
  "conflict-methods-twin", "Conflict resolution method (graded)", bins = res_bins, show = FALSE,
  cards = data.frame(
    text = c("A minor wording dispute in a memo; let it lie for now.",
             "A genuine design disagreement and two weeks to resolve it properly.",
             "Both teams give up part of their request to hit tomorrow's deadline.",
             "An unsafe procedure must stop immediately, whatever the objections.",
             "The relationship with the customer is fragile this week, so stress common ground."),
    bin = c("Withdrawal", "Confronting", "Compromising", "Forcing", "Smoothing"),
    stringsAsFactors = FALSE),
  question = "<p>For each situation, choose the conflict-resolution method that fits best.</p>")
```

### Try it yourself: score the supplier choice

Multi-criteria scoring turns "MCS feels safer" into a defensible number.

```{webr-r}
#| label: supplier-scoring-10
criteria <- c("Quality record", "Delivery reliability", "Price",
              "Technical capability", "Supply-chain risk")
weight   <- c(0.30, 0.25, 0.15, 0.15, 0.15)   # must sum to 1

# Score each supplier 1 (poor) to 5 (excellent) on each criterion.
mcs     <- c(5, 4, 3, 3, 2)   # trusted, but production moving to Mexico
paragon <- c(3, 3, 4, 5, 3)   # unproven, new tech, plant 4 h away, labour talks

mcs_total     <- sum(weight * mcs)
paragon_total <- sum(weight * paragon)

tab <- data.frame(Criterion = criteria, Weight = weight,
                  MCS = mcs, Paragon = paragon)
print(tab, row.names = FALSE)
cat(sprintf("\nWeighted score  MCS: %.2f   Paragon: %.2f\n", mcs_total, paragon_total))
cat(sprintf("Recommended: %s\n", ifelse(mcs_total >= paragon_total, "MCS", "Paragon")))
cat("Change the weights to your project's priorities and watch the answer move.\n")
```

### Try it yourself: weighted supplier scoring by hand

Multiply each score by its weight, then total each supplier's column. Each
team picks its own **data set**.

```{r tryit-scoring-10, echo=FALSE, results='asis'}
sc_crit <- c("Quality record", "Delivery reliability", "Price", "Technical capability")
sc_sets <- list(
  list(w = c(.40, .30, .20, .10), Alpha = c(4, 3, 5, 2), Beta = c(5, 4, 2, 3), Gamma = c(3, 5, 4, 5)),
  list(w = c(.35, .25, .25, .15), Alpha = c(5, 2, 4, 3), Beta = c(3, 5, 3, 4), Gamma = c(4, 4, 5, 2)),
  list(w = c(.50, .20, .20, .10), Alpha = c(3, 5, 4, 4), Beta = c(4, 2, 5, 2), Gamma = c(5, 3, 2, 5)),
  list(w = c(.30, .30, .25, .15), Alpha = c(2, 4, 5, 5), Beta = c(5, 3, 3, 2), Gamma = c(4, 5, 2, 3)))
sc_table <- function(s) {
  data.frame(Criterion = sc_crit, Weight = s$w, `Alpha score` = s$Alpha, `Beta score` = s$Beta,
             `Gamma score` = s$Gamma, `Alpha x weight` = s$Alpha * s$w, `Beta x weight` = s$Beta * s$w,
             `Gamma x weight` = s$Gamma * s$w, check.names = FALSE)
}
sc_answers <- function(s) {
  list(tryit_answer("Alpha total", sum(s$Alpha * s$w), tol = 0.01),
       tryit_answer("Beta total", sum(s$Beta * s$w), tol = 0.01),
       tryit_answer("Gamma total", sum(s$Gamma * s$w), tol = 0.01))
}
sc_cols <- c("Alpha x weight", "Beta x weight", "Gamma x weight")
tryit_grid(
  "supplier-scoring", "Weighted supplier score",
  data = sc_table(sc_sets[[1]]), blank = sc_cols, tol = 0.01, digits = 2,
  answers = sc_answers(sc_sets[[1]]),
  teams = lapply(sc_sets[-1], function(s) list(data = sc_table(s), answers = sc_answers(s))),
  intro = "Scores run from 1 (poor) to 5 (excellent); the weights sum to 1.")

sc_twin <- list(w = c(.40, .20, .25, .15), Alpha = c(4, 5, 3, 2), Beta = c(3, 3, 5, 4), Gamma = c(5, 2, 2, 5))
tryit_grid(
  "supplier-scoring-twin", "Weighted supplier score (graded)", show = FALSE,
  data = sc_table(sc_twin), blank = sc_cols, tol = 0.01, digits = 2, answers = sc_answers(sc_twin),
  question = paste0("<p>Three suppliers were scored from 1 to 5 against four weighted criteria. ",
    "Calculate each score x weight, then each supplier's weighted total (two decimal places).</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

Weighted total = sum of (score x weight) down each supplier's column. In data
set 1: Alpha 3.70, Beta 3.90, Gamma 4.00, so Gamma is recommended. The ranking
depends on the weights: shift weight from Price to Quality and re-run the sums
to see how fragile a narrow win can be.

</details>

## Key Takeaways

- PMs negotiate constantly --- for resources, with contractors, with functional managers. **Before** is the biggest stage; prepare more than you think you need.
- **Separate people from the problem**, focus on **interests** not positions, aim for **win-win or no win**, and talk money last.
- **BATNA** sets your **reservation point**; the overlap of both reservation points is the **ZOPA**. No ZOPA, no deal --- improve your BATNA or change the package.
- The person **asking the questions** usually controls the conversation.
- Project conflict is **perceptual and dynamic**. Types: **goal-oriented, administrative, interpersonal**. The **interactionist** view holds that a controlled amount of conflict is healthy.
- Teams fight most about **schedules, priorities, and resources** --- **personality** is last and is usually a misdiagnosis of an unresolved root cause.
- Match the resolution method --- **confronting, compromising, withdrawal, smoothing, forcing** --- to the situation; confronting is the durable fix.

## Case Study: Meridian R/C — Week 10

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 10: The MCS/Paragon Showdown"}

By Week 10, the motor and controller package --- the single biggest line item
in Meridian's budget and the longest-lead, critical-path activity in the
schedule --- stops being a formality. The Week 9 budget assumed MCS's original
quote: **\$145,000** for the full package, with a **25-day lead time**. It is
the number already baked into every schedule and cost baseline you have built
so far.

Then Marcus Webb, your procurement lead, brings you a competing quote.
Paragon, the newer Kingston, Ontario supplier, will do the same package for
**\$118,000** --- in just **20 days**. Marcus wants to switch immediately.
Sofia Reyes, your quality and test engineer, pushes back hard: Paragon's
controller technology is unproven, and the Week 7 risk register already
flagged it as **R1** --- a 30% chance the technology fails qualification, at
an \$80,000 impact, for an EMV of **\$24,000**. Add that risk cost to
Paragon's sticker price and the real, risk-adjusted cost is
**\$118,000 + \$24,000 = \$142,000** --- narrowly cheaper than MCS on paper,
but not the bargain it first appears.

You will not let the team split the difference by shouting. Your BATNA going
into the negotiation is simple and safe: proceed with MCS at \$145,000, the
proven supplier, with no qualification risk. But a strong BATNA is leverage,
not a decision --- so instead of picking a side in the Marcus/Sofia standoff,
you take Paragon's quote back to MCS. MCS knows the stakes: losing this
contract could cost them the follow-on 100,000-unit production order Kestrel
Retail Group is set to place once Meridian's prototype passes testing. Facing
a credible alternative, MCS concedes --- dropping its price to **\$136,000**,
a **\$9,000 (6.2%)** reduction from its original quote, while holding its
25-day lead time.

The outcome resolves both negotiations at once. Meridian keeps its proven
supplier and captures real, verified savings, with no new qualification risk
introduced. And the Marcus/Sofia conflict is settled not by forcing one side
to lose, but through **compromising**: Marcus gets the lower price he wanted,
Sofia keeps the proven supplier she insisted on. Neither gets everything;
both get enough.

:::

## Practice Problems

<details>
<summary>Problem 1: Name the cause of conflict</summary>

Which cause does each statement point to?

1. "I can't talk to her after what her son said to my son."
2. "Sales management here are lazy --- they always leave by 4:00."
3. "She barely looked at my report; she's already decided to ignore it."
4. "Our computers are slow because their division always gets the best ones."
5. "The worse your project does at year-end, the better mine looks."
6. "Why should I do that for the PM if my functional manager didn't tell me?"

<details>
<summary>Answers</summary>

1. **Personal grudge** (interpersonal).
2. **Different mindsets among departments** (organizational).
3. **Misconceptions about others' behaviour** (interpersonal).
4. **Scarce resources** (organizational).
5. **Reward systems** (organizational).
6. **Uncertainty / two bosses** (organizational, matrix).

</details>

</details>

<details>
<summary>Problem 2: ZOPA and a walk-away</summary>

Your BATNA for a tooling package is to build it in-house for \$98 000, so
that is the most you will pay. The vendor's floor is \$102 000.

(a) Is there a ZOPA? (b) What are your options?

<details>
<summary>Solution</summary>

```{r p2-zopa-10, echo=TRUE}
buyer_max  <- 98000   # set by your BATNA: the in-house cost
seller_min <- 102000
cat(ifelse(seller_min <= buyer_max, "ZOPA exists", "No ZOPA"), "\n")
cat(sprintf("Gap to close: $%s\n", format(seller_min - buyer_max, big.mark = " ")))
```

(a) No --- the vendor's floor is \$4 000 above your ceiling. (b) Exercise your
BATNA (build in-house at \$98 000), expand the package so the vendor's per-unit
economics improve, split tooling ownership, or find a third vendor.

</details>

</details>

<details>
<summary>Problem 3 (discussion): the functional manager skips your stand-up</summary>

The maintenance supervisor on your project has missed the first two morning
meetings. How could this signal *each* of: reward systems, scarce resources,
uncertainty, different mindsets, misconceptions, faulty communication, personal
grudge?

<details>
<summary>Discussion points</summary>

*Reward systems*: their boss doesn't value project work,
so they deprioritize it. *Scarce resources*: they are triple-booked and the
meeting lost. *Uncertainty*: they are not sure the meeting applies to them.
*Different mindsets*: maintenance culture sees planning meetings as talk, not
work. *Misconceptions*: they think you have already sidelined their input.
*Faulty communication*: the invite never reached them, or the time was
ambiguous. *Grudge*: a past project soured them on you. The point: **do not
jump to the grudge** --- confront it by asking, and you will usually find one of
the structural causes.

</details>

</details>

## Further Resources

```{r video-resources-10, echo=FALSE, results='asis'}
embed_youtube("JEoFHePrp6k", "Negotiation Strategy: ZOPA (Mediator Academy)")
embed_youtube("kJWF2oYzTq0", "BATNA and ZOPA in Negotiations")
```
