---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Resource Management

```{r setup-11, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Name** the constraint types on a project and focus on the resource constraints.
2. **Build** a resource histogram and spot over-allocation.
3. **Apply** resource **leveling** and the heuristics that decide which task moves first.
4. **Distinguish** time-constrained from resource-constrained scheduling.
5. **Apply** resource **smoothing** within available float.
6. **Explain** the multi-project resource problem and the priority rules used to resolve it.
7. **Quantify** the cost of multitasking on effective capacity.

<details>
<summary>Before we begin: your plan assumed the person was free --- were they?</summary>

Almost every first-draft project schedule is built as if the organization has
unlimited trained people standing by. It never does. This chapter is what
happens when the plan meets the staffing reality.

</details>

## Introduction: the optimism of the first schedule

The RC-vehicle schedule from [Chapter 8](08-Project-Scheduling.qmd) quietly
assumed every resource was available exactly when needed. Reality: there is **one
CMM** for validation, the **millwrights** are shared with production, the design
engineers are **borrowed** from a functional department, and the new assembly
line has a **learning curve**. *Building plans under the assumption of unlimited
resources inevitably leads to schedule slippage and extra cost as reality
overtakes the optimism.*

Projects run inside three kinds of constraint --- **technical**, **physical**, and
**resource** --- and this chapter is about the last one: **people, materials,
space, money, equipment.**

::: {.callout-note}
**Industry snapshot.** An automotive company with one set of climate chambers
watches validation projects stack up behind that single resource --- the classic
bottleneck. A food plant schedules every project around the *one* weekend
sanitation window. A defense program's cleared personnel are a fixed pool, so
every new task is a reallocation, not an addition.
:::

```{r video-resource-leveling-11, echo=FALSE, results='asis'}
embed_youtube("JlUXah6fsLI", "Resource Allocation and Leveling: Tutorial and Example (AF Math & Engineering)")
```

## Resource loading and the histogram

A **resource histogram** plots how much of a resource is required over time.
Stack each activity's demand day by day and the over-allocations jump out ---
provided you remember the chart assumes each person is available **100 % of the
time**, which is rarely true.

```{r fig-histogram-11, fig.cap="Resource histogram before and after leveling, against a limit of 9 people/day. Leveling shifts a task with float so the peak on day 12 no longer breaches the limit -- without extending the project.", fig.align='center', fig.width=10, fig.height=5}
days <- 1:14
before <- c(6,6,6,6,6, 9,9,9,9,10, 8,10,9,7)
after  <- c(6,6,6,6,6, 9,9,9,9,9,  9,9,9,7)
df <- data.frame(
  day = rep(days, 2),
  people = c(before, after),
  state = rep(c("Before leveling", "After leveling"), each = length(days))
)
df$state <- factor(df$state, levels = c("Before leveling", "After leveling"))

ggplot(df, aes(day, people)) +
  geom_col(aes(fill = people > 9), width = 0.75) +
  geom_hline(yintercept = 9, linetype = "dashed", color = "#C0392B", linewidth = 0.9) +
  annotate("text", x = 2, y = 9.6, label = "limit = 9/day",
           color = "#C0392B", size = 3 * diagram_text_scale, hjust = 0) +
  scale_fill_manual(values = c("FALSE" = "#5DADE2", "TRUE" = "#C0392B"), guide = "none") +
  facet_wrap(~state, ncol = 1) +
  scale_x_continuous(breaks = days) +
  labs(title = "Leveling removes the over-allocation peak",
       x = "Project day", y = "People required") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    strip.text = element_text(size = 11 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 9.5 * diagram_text_scale)
  )
```

### Try it yourself: build the histogram

Four tasks run on a fixed schedule. For each day, add up the people needed
and decide whether the day is over the limit. Each team picks its own **data
set** (and its own limit).

```{r tryit-histogram-11, echo=FALSE, results='asis'}
hg_sets <- list(
  list(start = c(1, 3, 6, 4), dur = c(5, 6, 5, 4), ppl = c(3, 4, 2, 3), limit = 8),
  list(start = c(1, 2, 5, 7), dur = c(4, 5, 5, 4), ppl = c(4, 3, 4, 2), limit = 6),
  list(start = c(1, 2, 4, 8), dur = c(6, 4, 6, 3), ppl = c(2, 5, 3, 4), limit = 8),
  list(start = c(1, 2, 3, 6), dur = c(3, 6, 3, 5), ppl = c(5, 3, 4, 3), limit = 9))
hg_table <- function(s) {
  days <- 1:10
  d <- sapply(1:4, function(k) ifelse(days >= s$start[k] & days < s$start[k] + s$dur[k], s$ppl[k], 0))
  colnames(d) <- paste("Task", LETTERS[1:4])
  total <- rowSums(d)
  data.frame(Day = days, d, Limit = s$limit, Total = total,
             `Over the limit?` = ifelse(total > s$limit, "Yes", "No"), check.names = FALSE)
}
hg_answers <- function(s) {
  t <- hg_table(s)
  list(tryit_answer("Peak demand", max(t$Total), tol = 0, suffix = "people"),
       tryit_answer("Number of days over the limit", sum(t$Total > s$limit), tol = 0, suffix = "days"))
}
hg_choice <- list(`Over the limit?` = c("Yes", "No"))
tryit_grid(
  "histogram-by-hand", "Resource histogram: people needed per day",
  data = hg_table(hg_sets[[1]]), blank = c("Total", "Over the limit?"), choice = hg_choice,
  tol = 0, digits = 0, answers = hg_answers(hg_sets[[1]]),
  teams = lapply(hg_sets[-1], function(s) list(data = hg_table(s), answers = hg_answers(s))),
  intro = "The columns for Tasks A to D show the people each task needs on that day (0 if it is not running).")

hg_twin <- list(start = c(1, 3, 4, 7), dur = c(4, 5, 5, 4), ppl = c(3, 2, 4, 5), limit = 7)
tryit_grid(
  "histogram-twin", "Resource histogram (graded)", show = FALSE,
  data = hg_table(hg_twin), blank = c("Total", "Over the limit?"), choice = hg_choice,
  tol = 0, digits = 0, answers = hg_answers(hg_twin),
  question = paste0("<p>Four tasks need the numbers of people shown on each day. The resource limit is ",
    "7 people per day. Complete the Total for each day and say whether it is over the limit, then ",
    "give the peak demand and the number of days over the limit.</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

Add the four task columns for each day. In data set 1 the totals are 3, 3,
7, 10, 10, 9, 9, 6, 2, 2 against a limit of 8, so days 4 to 7 are over the
limit: four days, with a peak of 10. Those are the days leveling has to fix,
by moving tasks that have float.

</details>

## Resource leveling

> **PMBOK:** *a technique in which start and finish dates are adjusted based on
> resource constraints, with the goal of balancing demand for resources against
> the available supply.*

Two objectives: (1) make sure resources are available at the right time, and
(2) let each activity run with the **smoothest possible transitions** across
resource-usage levels.

When you must choose which activity to move, work through them in this order:

1. Activities with the **smallest float**
2. Activities with the **smallest duration**
3. Activities with the **lowest ID** (earliest in the WBS)
4. Activities with the **most successors**
5. Activities requiring the **most resources**

### Worked Example 1: which task do you move?

Day 12 needs 10 people against a limit of 9. Four activities are running that
day:

Live cell --- change the float, duration, successors, or people for each task.

```{webr-r}
#| label: leveling-example-11
running_day12 <- data.frame(
  task = c("D", "E", "H", "J"),
  float = c(12, 0, 0, 5),
  duration = c(6, 6, 7, 3),
  successors = c(1, 2, 1, 0),
  people = c(3, 3, 3, 1)
)
print(running_day12, row.names = FALSE)

# Only tasks with float can move without extending the project.
movable <- running_day12[running_day12$float > 0, ]
# Among movable tasks, prefer smallest float, then smallest duration.
pick <- movable[order(movable$float, movable$duration), ][1, ]
cat(sprintf("\nMovable tasks: %s\n", paste(movable$task, collapse = ", ")))
cat(sprintf("Move first: task %s (float %d, duration %d) -- shift it later within its slack.\n",
            pick$task, pick$float, pick$duration))
cat("E and H are on the critical path (float 0) -- moving them would extend the project.\n")
```

### Try it yourself: which task do you move first?

```{r tryit-leveling-order-11, echo=FALSE, results='asis'}
tryit_order(
  "leveling-order", "Order the tasks by the leveling rule",
  steps = c("S: float 2 days, duration 4 days, ID 5",
            "Q: float 2 days, duration 4 days, ID 8",
            "P: float 2 days, duration 6 days, ID 3",
            "R: float 5 days, duration 3 days, ID 2",
            "T: float 9 days, duration 7 days, ID 1"),
  intro = "All five can move. Put the one to try first at the top: smallest float, then smallest duration, then lowest ID.",
  question = NULL)

tryit_order(
  "leveling-order-twin", "Order the tasks by the leveling rule (graded)", show = FALSE,
  steps = c("H: float 1 day, duration 3 days, ID 4", "K: float 1 day, duration 3 days, ID 9",
            "M: float 1 day, duration 5 days, ID 2", "J: float 4 days, duration 2 days, ID 6"),
  question = paste0("<p>Four tasks running on an over-allocated day can all be moved within their float. ",
    "Using the chapter's rule (smallest float first, then smallest duration, then lowest ID), put them ",
    "in the order you would try to move them.</p>"))
```

## Time-constrained vs. resource-constrained

- **Time-constrained**: the end date is fixed; you add or reassign resources to protect it (this is closer to *smoothing*).
- **Resource-constrained**: the resource limit is fixed; if leveling within float is not enough, **the project duration extends** (undesirable, but sometimes unavoidable).

## Resource smoothing

Smoothing is done **after** leveling and **never extends the project duration**.
It uses available **float** to even out resource usage --- filling the valleys
from the peaks.

### Try it yourself: smooth a resource within its float

```{webr-r}
#| label: smoothing-11
# Weekly demand for one resource, and the resource's weekly limit.
weekly_demand <- c(52, 30, 61, 24, 45, 20, 58)
limit <- 45

peak <- max(weekly_demand)
total_work <- sum(weekly_demand)
weeks <- length(weekly_demand)
ideal_flat <- total_work / weeks

cat(sprintf("Peak week demand   : %d  (limit %d -> %d over)\n", peak, limit, max(0, peak - limit)))
cat(sprintf("Total work         : %d resource-weeks over %d weeks\n", total_work, weeks))
cat(sprintf("Perfectly smooth   : %.1f per week\n", ideal_flat))

if (ideal_flat <= limit) {
  cat("\nThe work FITS within the limit if spread evenly -> smoothing can solve this\n")
  cat("by shifting tasks into their float, no schedule change needed.\n")
} else {
  cat(sprintf("\nEven perfectly smoothed (%.1f/wk) still exceeds the limit -> this needs\n", ideal_flat))
  cat("leveling (extend duration) or more resource, not just smoothing.\n")
}
```

## Splitting a task

**Splitting** means interrupting continuous work on an activity, moving that
resource to another activity, then returning. It helps *only* when there are no
large **start-up or shutdown costs**, and **only some tasks can be split** at all
(you cannot pause a concrete pour).

## The multi-project resource problem

Organizations rarely buy new resources for each project --- they **share** them
across projects. The core challenge is **interdependency**: a resource decision
in one project ripples into others. Common problems:

- Inefficient use of resources
- Resource **bottlenecks**
- **Ripple effects**
- Heightened pressure on people to **multitask**

Any multi-project scheduling system tries to minimize three things:

1. **Schedule slippage** --- often carrying real financial penalty clauses per day late.
2. **Resource utilization** losses --- the unused-capacity gap; the goal is to fill it with other projects' work.
3. **In-process inventory (WIP)** --- work waiting on an unavailable resource, stacked behind a bottleneck (e.g. one shared climate chamber).

```{r fig-utilization-11, fig.cap="Resource utilization across a project. The shaded area is capacity the project does not use -- in a multi-project shop it should be filled with concurrent projects' work rather than left idle.", fig.align='center', fig.width=10, fig.height=4.8}
wk <- 1:14
used <- c(3,4,5,6,6,7,7,6,5,5,4,3,3,2)
cap  <- 8
df <- data.frame(wk, used, unused = cap - used)
df_long <- tidyr::pivot_longer(df, c(used, unused), names_to = "kind", values_to = "res")
df_long$kind <- factor(df_long$kind, levels = c("unused", "used"),
                        labels = c("Unused capacity", "Used by this project"))

ggplot(df_long, aes(wk, res, fill = kind)) +
  geom_col(width = 0.8) +
  geom_hline(yintercept = cap, linetype = "dashed", color = "grey30") +
  scale_fill_manual(values = c("Used by this project" = "#5DADE2",
                               "Unused capacity" = "#F5CBA7"), name = NULL) +
  scale_x_continuous(breaks = wk) +
  labs(title = "Where is the idle capacity?",
       x = "Project week", y = "Resource units") +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Priority rules for allocating shared resources

| Rule | Logic |
|---|---|
| **First in line** (first come, first served) | Simplest --- follow the master calendar; ignores priority and commercial value |
| **Greatest resource demand** | Schedule the most resource-hungry project first so bottlenecks surface early |
| **Greatest resource utilization** (most resource *types*) | Schedule the project that uses the widest mix of resources first, keeping all of them busy |
| **Minimum late finish** (earliest finish first) | Prioritize the project with the earliest late-finish date |
| **Mathematical programming** | Optimize (minimize total time / utilization loss / lateness); powerful but complex and slow for large problems |

### Try it yourself: which priority rule?

```{r tryit-priority-11, echo=FALSE, results='asis'}
pr_bins <- c("First in line", "Greatest resource demand", "Greatest resource utilization",
             "Minimum late finish", "Mathematical programming")
tryit_sort(
  "priority-rules", "Which priority rule is each decision using?",
  bins = pr_bins,
  cards = data.frame(
    text = c("Whichever project was approved first gets the CMM first.",
             "Put the project with the biggest peak demand in first so bottlenecks surface early.",
             "Favour the project using the widest mix of resource types, to keep them all busy.",
             "The project with the earliest late-finish date goes first.",
             "An optimization model minimizes total lateness across 40 projects.",
             "Follow the master calendar and ignore commercial value.",
             "One launch date is immovable and tighter than all the others, so that project goes first."),
    bin = c("First in line", "Greatest resource demand", "Greatest resource utilization",
            "Minimum late finish", "Mathematical programming", "First in line", "Minimum late finish"),
    why = c("First come, first served.", "Most resource-hungry first.", "Widest mix of resource types first.",
            "Earliest late-finish first.", "Optimization: powerful but slow for large problems.",
            "The simplest rule; ignores priority.", "Tightest deadline first."),
    stringsAsFactors = FALSE),
  intro = "Pick the rule each decision follows.", question = NULL)

tryit_sort(
  "priority-rules-twin", "Priority rules (graded)", bins = pr_bins, show = FALSE,
  cards = data.frame(
    text = c("Requests are served in the order they arrived.",
             "The project needing the most people goes first.",
             "The project whose deadline is nearest goes first.",
             "A solver finds the schedule with the least total lateness.",
             "The project that uses the most different kinds of resource goes first."),
    bin = c("First in line", "Greatest resource demand", "Minimum late finish",
            "Mathematical programming", "Greatest resource utilization"),
    stringsAsFactors = FALSE),
  question = "<p>For each description, choose the priority rule for allocating shared resources that it follows.</p>")
```

### Try it yourself: "40 hours is not one week's work"

```{r calc-multitask-11, echo=FALSE, results='asis'}
qwebr_calc_card(
  label = "multitask-11",
  title = "Multitasking Capacity Calculator",
  icon = "layer-group",
  inputs = list(
    qwebr_calc_input("nominal_hours", "Nominal weekly hours", value = 40, min = 10, max = 60, step = 1, suffix = "h"),
    qwebr_calc_input("switch_loss_per_extra_project", "Capacity lost per extra project", value = 0.15, min = 0, max = 0.3, step = 0.01),
    qwebr_calc_input("projects_carried", "Projects carried at once", value = 3, min = 1, max = 6, step = 1, type = "slider")
  )
)
```

<details class="qwebr-calc-source">
<summary>Show the live R code behind this calculator</summary>

The card above re-runs this exact webR code every time you change an input, substituting your values for the variables at the top. Edit the code directly and press **Run Code** to experiment further.

```{webr-r}
#| label: multitask-11
nominal_hours <- 40
# Each additional concurrent project imposes a context-switching loss.
switch_loss_per_extra_project <- 0.15   # 15% of capacity lost per extra project
projects_carried <- 3

effective_hours <- function(projects) {
  extras <- projects - 1
  max(nominal_hours * (1 - switch_loss_per_extra_project * extras), 0)
}

counts <- 1:6
eff <- sapply(counts, effective_hours)
par(mar = c(5, 5, 3, 1), cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)
bp <- barplot(eff, names.arg = counts, ylim = c(0, nominal_hours * 1.1),
              col = ifelse(counts == projects_carried, "#e74c3c", "#2980b9"),
              xlab = "Projects carried at once", ylab = "Effective hours per week",
              main = "Nominal hours are not effective hours")
abline(h = nominal_hours, lty = 2, lwd = 2, col = "#7f8c8d")

now <- effective_hours(projects_carried)
cat(sprintf("On %g project(s): %.1f effective hours of %g (%.0f%% productive)\n",
            projects_carried, now, nominal_hours, 100 * now / nominal_hours))
for (projects in counts) {
  cat(sprintf("On %d project(s)  %.1f effective hours (%.0f%% productive)\n",
              projects, eff[projects], 100 * eff[projects] / nominal_hours))
}
```

</details>

### Try it yourself: effective hours by hand

Each extra project you carry costs a share of your capacity. Work out what is
left. Each team picks its own **data set**.

```{r tryit-multitask-11, echo=FALSE, results='asis'}
mt_sets <- list(c(hrs = 40, loss = 15, n = 3), c(hrs = 38, loss = 12, n = 4),
                c(hrs = 45, loss = 10, n = 5), c(hrs = 35, loss = 20, n = 2))
mt_given <- function(v) c("Nominal weekly hours" = v[["hrs"]],
                          "Capacity lost per extra project" = sprintf("%g %%", v[["loss"]]),
                          "Projects carried" = v[["n"]])
mt_answers <- function(v) {
  eff <- function(n) v[["hrs"]] * (1 - v[["loss"]] / 100 * (n - 1))
  list(tryit_answer("Effective hours per week", eff(v[["n"]]), tol = 0.05, suffix = "h"),
       tryit_answer("Productive share of nominal hours", 100 * eff(v[["n"]]) / v[["hrs"]], tol = 0.5, suffix = "%"),
       tryit_answer("Hours lost per week", v[["hrs"]] - eff(v[["n"]]), tol = 0.05, suffix = "h"),
       tryit_answer("Effective hours with one project fewer", eff(v[["n"]] - 1), tol = 0.05, suffix = "h"))
}
tryit_number(
  "multitask-by-hand", "Effective capacity when multitasking",
  answers = mt_answers(mt_sets[[1]]), given = mt_given(mt_sets[[1]]),
  teams = lapply(mt_sets[-1], function(v) list(given = mt_given(v), answers = mt_answers(v))),
  intro = "Effective hours = nominal x (1 - loss x (projects - 1)).")

tryit_number(
  "multitask-twin", "Effective capacity (graded)", show = FALSE,
  answers = mt_answers(c(hrs = 42, loss = 14, n = 4)),
  question = paste0("<p>An engineer has 42 nominal hours per week. Each extra project carried at once ",
    "costs 14 % of nominal capacity. The engineer carries 4 projects. Calculate the effective hours per ",
    "week, the productive share of nominal hours (nearest whole per cent), the hours lost, and the ",
    "effective hours with one project fewer.</p>"))
```

<details class="tryit-solution">
<summary>Show the worked solution</summary>

With n projects there are n - 1 "extra" ones. Data set 1: 40 x (1 - 0.15 x 2) =
28 effective hours, 70 % productive, 12 h lost; dropping to 2 projects gives
40 x 0.85 = 34 h. Cutting one project recovers 6 h: the nominal 40 hours were
never real.

</details>

## Sustainability and resources

Resource management is not only about people-hours. The **circular-economy**
thinking from [Chapter 1](01-Sustainability.qmd) applies directly: minimize raw
material use, prefer recyclable and local inputs, design the project so
equipment and tooling can be reused on the next program, and track material and
energy as managed resources with their own KPIs --- not just labour and money.

## Key Takeaways

- Projects run inside **technical, physical, and resource** constraints; resources are people, materials, space, money, equipment.
- A **resource histogram** exposes **over-allocation** (against a 100 %-availability assumption).
- **Leveling** shifts tasks within their **float** to respect a resource limit; move tasks in order of **smallest float → smallest duration → lowest ID → most successors → most resources**. Critical-path tasks have no float, so moving them extends the project.
- **Time-constrained** protects the date with resources; **resource-constrained** may extend the date.
- **Smoothing** happens after leveling, uses float, and **never** extends duration.
- **Splitting** helps only when start-up/shutdown costs are low and the task can be split.
- Multi-project sharing creates **bottlenecks, ripple effects, and multitasking pressure**; minimize **slippage, utilization loss, and WIP**, and pick a **priority rule** deliberately.
- **Multitasking is not free** --- "40 hours" of nominal capacity is far less once someone carries several projects.

## Case Study: Meridian R/C — Week 11

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 11: Priya's 160% Week"}

By Week 11 the tooling phase is in full swing, and two activities are running side by side: C (body tooling build, 20 days) and D (frame tooling build, 15 days), both released from B (detailed design release) on day 18. Priya Nandakumar, Meridian's Manufacturing Engineer, is the only person qualified to supervise both dies at once — the matrix organization never gave her a dedicated tooling team, just herself and two die-shop technicians who report through her for the duration of the project.

You pull the resource histogram for the tooling window and the number jumps out immediately: Priya is booked at 100% on C and 60% on D at the same time — 160% of one person's capacity, a 60-percentage-point overallocation that no amount of overtime coffee is going to fix on its own.

Before moving anything, you go back to the CPM table. C has 5 days of float (LS day 23, LF day 43). D has 10 days of float (LS day 28, LF day 43). Neither is on the critical path A–B–E–F–H–I–J, so neither move threatens the 67-day finish — but D gives you twice the room to work with. You shift D's start later, inside its own float, so Priya's two commitments no longer collide the way they did on paper. C keeps running on its original clock; D slides toward its late-start date of day 28 and still lands exactly on its late-finish date of day 43, precisely when F (prototype assembly) needs both dies ready. The overallocation is resolved, and the 67-day project finish never moves.

While you're in the histogram, you also pull the underlying weekly team-hours demand for the whole tooling phase against the department's 45-hour/week capacity ceiling: 30, 48, 44, 52, and 38 hours across five consecutive weeks. Weeks 2 and 4 breach the limit, by 3 and 7 hours respectively — week 4 is the single worst overallocation — while weeks 1, 3, and 5 sit under capacity. Averaged out, the phase demands 42.4 hours per week, comfortably under the 45-hour limit. That average is the tell: there is enough slack in the valley weeks to absorb the peak weeks' extra hours using float alone, which makes this a smoothing problem rather than a leveling one — no added headcount, no extended schedule, just a better-shaped curve.

**Weekly team-hours demand, tooling phase (capacity limit: 45 hrs/week)**

| Week | Demand (hrs) | vs. 45-hr limit |
|---|---|---|
| 1 | 30 | 15 under |
| 2 | 48 | 3 over |
| 3 | 44 | 1 under |
| 4 | 52 | 7 over (peak) |
| 5 | 38 | 7 under |

**Activity float, for reference**

| Activity | Float | Critical? |
|---|---|---|
| B (detailed design release) | 0 days | Yes |
| C (body tooling build) | 5 days | No |
| D (frame tooling build) | 10 days | No |

:::

## Practice Problems

<details>
<summary>Problem 1: Level to the limit</summary>

Day 10 requires 10 people; the limit is 9. Running that day: task C (float 4,
duration 5, 4 people), task D (float 0, duration 6, 3 people), task F (float 0,
duration 6, 3 people). Which task moves, and what is the risk if none can?

<details>
<summary>Solution</summary>

```{r p1-level-11, echo=TRUE}
d10 <- data.frame(task = c("C", "D", "F"),
                  float = c(4, 0, 0), dur = c(5, 6, 6), ppl = c(4, 3, 3))
mov <- d10[d10$float > 0, ]
cat("Only task with float:", mov$task, "\n")
cat("Move C into its 4 days of slack -> day-10 demand drops to 6, under the limit.\n")
cat("If C could not move (no split, hard dependency), the project duration extends.\n")
```

</details>

</details>

<details>
<summary>Problem 2: Does this need leveling or smoothing?</summary>

A resource has a 40 h/week limit. Weekly demand over 6 weeks is
55, 20, 48, 15, 50, 22 h. Total 210 h.

<details>
<summary>Solution</summary>

```{r p2-smooth-11, echo=TRUE}
demand <- c(55, 20, 48, 15, 50, 22); limit <- 40
flat <- sum(demand) / length(demand)
cat(sprintf("Even spread: %.1f h/week vs limit %d\n", flat, limit))
cat(ifelse(flat <= limit,
           "Fits if spread evenly -> smoothing (shift within float) can solve it.\n",
           "Still over even when flat -> leveling (extend) or add resource.\n"))
```

35 h/week average is under the 40 h limit, so **smoothing** within float can
solve it without extending the project.

</details>

</details>

<details>
<summary>Problem 3 (discussion): the shared CMM bottleneck</summary>

Three projects, including yours, all need the single CMM for final validation in
the same two-week window. Using the priority rules, how would you decide the
order, and what could reduce the bottleneck?

<details>
<summary>Discussion points</summary>

Apply a rule and be explicit about it: **minimum late
finish** (whichever project has the tightest deadline goes first) is defensible
if one launch date is immovable; **greatest resource demand** surfaces the
bottleneck earliest. Reduce it by: staggering the validation tasks into each
project's float, renting a second CMM for the peak, outsourcing some measurement
to a metrology lab, or moving non-CMM checks off the critical window so the CMM
time is used only for what truly needs it.

</details>

</details>

## Further Resources

```{r video-resources-11, echo=FALSE, results='asis'}
embed_youtube("iSGWSGfG65U", "What Is Resource Leveling in Project Management?")
```
