---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Project Control Systems

```{r setup-12, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter you will be able to:

1. **Describe** the monitoring-and-control loop and the performance measurement baseline.
2. **Read** the three earned-value curves: planned value (PV), earned value (EV), and actual cost (AC).
3. **Compute** schedule and cost variance and the SPI / CPI performance indices.
4. **Forecast** with EAC, ETC, VAC, and TCPI, and interpret them.
5. **Run** integrated change control so the baseline stays meaningful.
6. **Report** status honestly and recognize the "watermelon" trap.

<details>
<summary>Before we begin: the project has spent 60 % of its money --- is it 60 % done?</summary>

Almost certainly not. "Percent of budget spent" and "percent of work finished"
are different numbers, and the gap between them is the whole subject of this
chapter.

</details>

## Introduction: from a plan to a control system

You have planned the RC-vehicle project --- scope, WBS, schedule, budget. It is
now **week 8**. The sponsor asks the only question that matters at this stage:
**are we on track, and if not, how far off, and what will it cost to finish?**

Answering that needs a **control system**: a closed loop that **measures**
progress, **compares** it to the baseline, **analyzes** the variance, and drives
a corrective **action** --- then measures again.

```{r fig-control-loop-12, fig.cap="The monitoring-and-control loop. Without the 'compare to baseline' step, status reporting is just activity; with it, you get variance you can act on.", fig.align='center', fig.width=9, fig.height=4.5}
loop <- data.frame(
  x = c(0, 2.2, 4.4, 6.6),
  y = 0,
  label = c("MEASURE\nprogress & cost", "COMPARE\nto baseline",
            "ANALYZE\nvariance", "ACT\ncorrective action")
)
ggplot(loop, aes(x, y)) +
  geom_label(aes(label = label), size = 3 * diagram_text_scale, lineheight = 0.9,
             fill = "#D6EAF8", label.size = 0, fontface = "bold") +
  annotate("segment", x = c(0.95, 3.15, 5.35), xend = c(1.55, 3.75, 5.95),
           y = 0, yend = 0, arrow = arrow(length = unit(0.18, "cm")), color = "grey40") +
  annotate("curve", x = 6.6, y = -0.35, xend = 0, yend = -0.35,
           curvature = 0.25, arrow = arrow(length = unit(0.18, "cm")), color = "grey40") +
  annotate("text", x = 3.3, y = -0.62, label = "repeat every reporting period",
           size = 2.8 * diagram_text_scale, fontface = "italic", color = "grey45") +
  coord_cartesian(xlim = c(-0.9, 7.5), ylim = c(-0.9, 0.6)) +
  theme_void()
```

The reference for "compare" is the **performance measurement baseline** --- the
approved **scope + schedule + cost** locked in at the end of planning
([Chapter 9](09-Cost-Estimation-and-Budgeting.qmd)'s S-curve is its cost
dimension).

::: {.callout-note}
**Industry snapshot.** Earned value management is **contractually mandated** on
large U.S. defense programs (EIA-748). Automotive programs run a lighter version
at the milestone level. A food-plant capital project tracks committed vs. spent
vs. installed. The math below is the same at every scale.
:::

```{r video-evm-12, echo=FALSE, results='asis'}
embed_youtube("x0EgZnXdunw", "What Is Earned Value? (Ricardo Vargas)")
```

## The three curves

Earned value management tracks three cumulative quantities against time:

| Term | Also called | Meaning |
|---|---|---|
| **PV** --- Planned Value | BCWS | The budgeted cost of the work **scheduled** to be done by now |
| **EV** --- Earned Value | BCWP | The budgeted cost of the work **actually done** by now |
| **AC** --- Actual Cost | ACWP | What has actually been **spent** to do that work |

**BAC** (Budget At Completion) is the total baseline budget --- where the PV curve
ends.

```{r fig-evm-curves-12, fig.cap="The three curves through week 8 of the RC-vehicle project (BAC = $620 000, off the top of this view). By the status date EV sits below PV -- behind schedule -- and AC sits above EV -- over cost. The vertical gaps at week 8 are the schedule and cost variances.", fig.align='center', fig.width=10, fig.height=6}
cutoff <- 8
wk <- seq(0, cutoff, by = 0.25)
# S-shaped ramps scaled to pass through the week-8 values in Worked Example 1.
shape <- function(t, k) (t / cutoff)^k
df <- data.frame(
  wk = wk,
  PV = 217000 * shape(wk, 1.8),
  EV = 186000 * shape(wk, 2.0),
  AC = 210000 * shape(wk, 1.7)
)

ggplot() +
  geom_line(data = df, aes(wk, PV, color = "PV (planned value)"), linewidth = 1.4) +
  geom_line(data = df, aes(wk, EV, color = "EV (earned value)"), linewidth = 1.4) +
  geom_line(data = df, aes(wk, AC, color = "AC (actual cost)"), linewidth = 1.4) +
  geom_vline(xintercept = cutoff, linetype = "dotted", color = "grey40") +
  geom_point(data = data.frame(m = c("PV", "EV", "AC"), y = c(217000, 186000, 210000)),
             aes(x = cutoff, y = y, color = m), size = 3, show.legend = FALSE) +
  annotate("text", x = cutoff + 0.12, y = 228000, hjust = 0,
           label = "PV  $217 000", size = 3 * diagram_text_scale, color = "#2E86C1") +
  annotate("text", x = cutoff + 0.12, y = 205000, hjust = 0,
           label = "AC  $210 000", size = 3 * diagram_text_scale, color = "#C0392B") +
  annotate("text", x = cutoff + 0.12, y = 178000, hjust = 0,
           label = "EV  $186 000", size = 3 * diagram_text_scale, color = "#27AE60") +
  annotate("text", x = cutoff - 0.25, y = 45000, hjust = 1,
           label = "status date\n(week 8)", size = 3 * diagram_text_scale, lineheight = 0.9) +
  scale_color_manual(values = c("PV (planned value)" = "#2E86C1",
                                "EV (earned value)" = "#27AE60",
                                "AC (actual cost)" = "#C0392B",
                                "PV" = "#2E86C1", "EV" = "#27AE60", "AC" = "#C0392B"),
                     breaks = c("PV (planned value)", "EV (earned value)", "AC (actual cost)"),
                     name = NULL) +
  scale_y_continuous(labels = scales::dollar_format(big.mark = " ")) +
  coord_cartesian(xlim = c(0, 9.7)) +
  labs(title = "Earned value: planned vs. earned vs. actual",
       x = "Project week", y = "Cumulative cost") +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

## The six core metrics

$$
\begin{aligned}
\text{Schedule Variance} \quad SV &= EV - PV \\
\text{Cost Variance} \quad CV &= EV - AC \\
\text{Schedule Performance Index} \quad SPI &= EV / PV \\
\text{Cost Performance Index} \quad CPI &= EV / AC
\end{aligned}
$$

| Reading | Schedule | Cost |
|---|---|---|
| **Ahead / under** | $SV > 0$, $SPI > 1$ | $CV > 0$, $CPI > 1$ |
| **On plan** | $SV = 0$, $SPI = 1$ | $CV = 0$, $CPI = 1$ |
| **Behind / over** | $SV < 0$, $SPI < 1$ | $CV < 0$, $CPI < 1$ |

### Worked Example 1: the RC-vehicle project at week 8

Baseline budget **BAC = \$620 000**. By week 8 the plan said **35 %** of the
work should be done; **30 %** is actually done; **\$210 000** has been spent.

Live cell --- change the planned %, the actual %, or the actual cost.

```{webr-r}
#| label: evm-example-12
BAC <- 620000
planned_pct <- 0.35
actual_pct  <- 0.30
AC <- 210000

PV <- planned_pct * BAC
EV <- actual_pct  * BAC

SV  <- EV - PV
CV  <- EV - AC
SPI <- EV / PV
CPI <- EV / AC

cat(sprintf("PV  = $%s      EV  = $%s      AC  = $%s\n",
            format(PV, big.mark = " "), format(EV, big.mark = " "),
            format(AC, big.mark = " ")))
cat(sprintf("SV  = $%s  (%s schedule)\n", format(SV, big.mark = " "),
            ifelse(SV < 0, "behind", "ahead")))
cat(sprintf("CV  = $%s  (%s budget)\n", format(CV, big.mark = " "),
            ifelse(CV < 0, "over", "under")))
cat(sprintf("SPI = %.3f   (getting %.0f cents of schedule per planned dollar)\n",
            SPI, SPI * 100))
cat(sprintf("CPI = %.3f   (getting %.0f cents of work per dollar spent)\n",
            CPI, CPI * 100))
```

The project is **behind schedule and over cost** --- and this is visible in week 8,
while there is still time to act, instead of as a surprise at the end.

## Forecasting

| Forecast | Formula | Meaning |
|---|---|---|
| **EAC** (Estimate At Completion) | $BAC / CPI$ | Total cost if current cost efficiency holds |
| EAC (schedule too) | $AC + \dfrac{BAC - EV}{CPI \times SPI}$ | If both cost *and* schedule performance persist |
| EAC (one-off overrun) | $AC + (BAC - EV)$ | If the overrun so far was a one-time event |
| **ETC** (Estimate To Complete) | $EAC - AC$ | Money still needed from here |
| **VAC** (Variance At Completion) | $BAC - EAC$ | Projected total over/under budget |
| **TCPI** (To-Complete Performance Index) | $\dfrac{BAC - EV}{BAC - AC}$ | The cost efficiency the *remaining* work must hit to still land on BAC |

### Try it yourself: EAC scenarios

```{webr-r}
#| label: eac-scenarios-12
BAC <- 620000
EV  <- 186000
AC  <- 210000
PV  <- 217000
CPI <- EV / AC
SPI <- EV / PV

eac_cpi       <- BAC / CPI
eac_oneoff    <- AC + (BAC - EV)
eac_cost_sched<- AC + (BAC - EV) / (CPI * SPI)

cat(sprintf("CPI = %.3f   SPI = %.3f\n\n", CPI, SPI))
cat(sprintf("EAC, cost efficiency continues : $%s\n", format(round(eac_cpi), big.mark = " ")))
cat(sprintf("EAC, overrun was a one-off     : $%s\n", format(round(eac_oneoff), big.mark = " ")))
cat(sprintf("EAC, cost AND schedule persist : $%s\n", format(round(eac_cost_sched), big.mark = " ")))

etc  <- eac_cpi - AC
vac  <- BAC - eac_cpi
tcpi <- (BAC - EV) / (BAC - AC)
cat(sprintf("\nETC (money still needed)  : $%s\n", format(round(etc), big.mark = " ")))
cat(sprintf("VAC (projected overrun)   : $%s\n", format(round(vac), big.mark = " ")))
cat(sprintf("TCPI (efficiency needed)  : %.3f  ", tcpi))
cat(ifelse(tcpi > 1.10, "-- that is a big ask; the BAC is likely unrealistic now.\n",
    ifelse(tcpi > 1.0, "-- recoverable, but the team must run leaner than planned.\n",
           "-- achievable.\n")))
```

<details>
<summary>Which EAC formula should you actually report?</summary>

It depends on *why* you are over. If a one-time event caused it (a failed test
you have since fixed), $AC + (BAC - EV)$ is fair. If the overrun reflects how the
team is fundamentally working (rates, productivity), $BAC/CPI$ is the honest
number. If you are *also* behind schedule and recovery will cost overtime, the
$CPI \times SPI$ version is the most conservative. Report the one your evidence
supports --- and say which assumption you made.

</details>

## Reading the S-curve honestly

- **Spend is not progress.** A project can be on budget (AC ≈ PV) and badly behind (EV ≪ PV) --- it just has not done the work yet.
- **The "watermelon" report** --- green on the outside, red on the inside. High-level status shows green because spend matches plan, while the earned value underneath is red.
- **Leading vs. lagging indicators.** SPI and CPI are somewhat lagging. Leading signals --- rising defect find-rate, slipping sub-milestones, key staff leaving --- warn you earlier. Watch both.

## Integrated change control

Control only works against a **meaningful baseline**. Every change request goes
through the same gate as in [Chapter 3](03-Scope-and-Organizational-Structures.qmd):
assess the impact on scope, schedule, and cost; approve or reject; and if
approved, **re-baseline** so future variance is measured against the new plan ---
not the old one you can no longer hit. **Configuration management** keeps the
product definition (drawings, specs, software versions) in step with the
approved baseline.

## Status reporting

The course's **Project Manager Report** and **Team Member Report** templates are
the reporting instruments: they roll up progress, cost, schedule, risks, and
issues on a fixed cadence. Good reporting is **honest** (report the red while it
is still fixable), **consistent** (same metrics every period so trends are
visible), and **actionable** (every variance has an owner and a next step). A
dashboard that only ever shows green is not a control system --- it is a
comfort blanket.

## Key Takeaways

- **Control** is a closed loop: **measure → compare to baseline → analyze variance → act**, every reporting period.
- Track **PV** (work scheduled), **EV** (work done), **AC** (money spent), all in budget dollars, against **BAC**.
- $SV = EV - PV$, $CV = EV - AC$, $SPI = EV/PV$, $CPI = EV/AC$. Below 1 / negative = behind / over.
- Forecast with **EAC** ($BAC/CPI$ is the common honest form), **ETC** $= EAC - AC$, **VAC** $= BAC - EAC$, **TCPI** $= (BAC-EV)/(BAC-AC)$.
- **Spend is not progress.** Beware the **watermelon** report; watch **leading** indicators, not just SPI/CPI.
- Keep the **baseline meaningful** through integrated change control and configuration management.
- Report **honestly, consistently, and actionably** --- every variance gets an owner.

## Case Study: Meridian R/C — Week 12

::: {.callout-tip .case-study title="📁 CASE STUDY FILE — Week 12: One Good Number Doesn't Make a Good Week"}

Day 50, and it's time to run the loop this chapter just walked you through
--- measure, compare, analyze, act --- against real Trailblazer numbers
instead of a textbook example. Sofia Reyes's weekly report lands with a
summary line that looks encouraging at a glance: spend is roughly tracking
the plan. You know better than to stop at a glance.

| Metric | Value |
|---|---|
| BAC (Budget At Completion) | \$761,300 |
| PV (Planned Value, day 50) | \$560,000 |
| EV (Earned Value, day 50) | \$520,000 |
| AC (Actual Cost, day 50) | \$545,000 |

You run the four core numbers yourself before the steering call. CV = EV −
AC = \$520,000 − \$545,000 = **−\$25,000** --- over budget. SV = EV − PV =
\$520,000 − \$560,000 = **−\$40,000** --- behind schedule. CPI = EV/AC =
520/545 ≈ **0.95**. SPI = EV/PV = 520/560 ≈ **0.93**. Running BAC/CPI forward
gives an EAC of roughly **\$798,000** --- a forecast, not a certainty, but not
a comfortable one either.

The schedule slip traces to one place: activity H, toy-safety certification,
which has sat on the critical path with zero float since the CPM network was
first built back in Week 6. The certification lab pushed back its review
date, and because H has no float, every day it slips is a day the whole
67-day project finish slips with it.

The cost story is the sharper lesson. Two weeks ago, Marcus Webb talked MCS
down from \$145,000 to **\$136,000** for the motor-and-controller package
--- a genuine \$9,000 win, hard-negotiated, real money, and the whole point
of using the competing Paragon quote as leverage. So why is AC still running
\$25,000 ahead of EV? Because elsewhere on the project, packaging samples had
to move by expedited freight to protect the schedule, and that bill quietly
ate the MCS saving and then some. A favourable variance on one activity,
read on its own, told the team nothing reliable about the project's overall
health --- exactly the trap this chapter's "watermelon" warning describes:
green on the outside, red once you look underneath.

There is a sliver of good news buried in the bad. At roughly \$798,000, the
EAC still sits just under Daniel Cho's \$800,000 charter ceiling from Week 3
--- a thin margin, but a margin, assuming nothing else slips before Week
13's closeout.

:::

## Practice Problems

<details>
<summary>Problem 1: Full EVM snapshot</summary>

BAC = \$400 000. At the status date: PV = \$180 000, EV = \$150 000,
AC = \$165 000. Compute SV, CV, SPI, CPI, EAC (BAC/CPI), ETC, VAC, TCPI, and
state the project's condition in one sentence.

**Solution.**

```{r p1-evm-12, echo=TRUE}
BAC <- 400000; PV <- 180000; EV <- 150000; AC <- 165000
SV <- EV - PV; CV <- EV - AC; SPI <- EV / PV; CPI <- EV / AC
EAC <- BAC / CPI; ETC <- EAC - AC; VAC <- BAC - EAC
TCPI <- (BAC - EV) / (BAC - AC)
cat(sprintf("SV=$%s  CV=$%s  SPI=%.3f  CPI=%.3f\n",
            format(SV, big.mark=" "), format(CV, big.mark=" "), SPI, CPI))
cat(sprintf("EAC=$%s  ETC=$%s  VAC=$%s  TCPI=%.3f\n",
            format(round(EAC), big.mark=" "), format(round(ETC), big.mark=" "),
            format(round(VAC), big.mark=" "), TCPI))
```

Behind schedule (SPI 0.83) and over cost (CPI 0.91); projected to finish about
\$40 000 over budget, and the remaining work must run at TCPI ≈ 1.06 to still
hit BAC.

</details>

<details>
<summary>Problem 2: On budget but in trouble</summary>

A project reports PV = \$500 000, EV = \$400 000, AC = \$400 000. The sponsor
sees "spend exactly matches earned value --- CPI = 1.0 --- we're fine." Are they?

**Answer.** Cost efficiency is fine (CPI = 1.0), but **SPI = EV/PV = 0.80** ---
the project has completed only 80 % of the work it should have by now. It is
**20 % behind schedule** and will either finish late or need to spend more
(overtime, added staff) to catch up, which will then push CPI below 1. This is
the classic watermelon: the cost view is green, the schedule view is red.

</details>

<details>
<summary>Problem 3 (discussion): re-baseline or not?</summary>

At week 8 the RC-vehicle project is forecast (EAC) at \$700 000 against a
\$620 000 BAC, mostly because the retailer added a second body colourway
(an approved scope change) and the new line's learning curve ran longer than
planned. Should you re-baseline?

**Discussion points.** Split the two causes. The **approved scope change** is a
legitimate reason to re-baseline --- the extra colourway was authorised, so the
budget and schedule should be formally updated and future variance measured
against the new plan. The **learning-curve overrun** is *performance* against
the plan, not a change to the plan --- that stays as variance you must recover or
explain. Re-baselining to bury performance problems ("we'll just move the
goalposts") destroys the value of the control system; re-baselining for
authorised changes keeps it honest.

</details>

## Further Resources

```{r video-resources-12, echo=FALSE, results='asis'}
embed_youtube("EEIu8-87nAY", "Earned Value Management Explained: PV, EV, AC, CPI, SPI")
embed_youtube("7DLyZjsrG14", "Earned Value Analysis Calculations Tutorial (CPI, SPI, CV, SV, EAC, ETC, TCPI)")
```
