```{r setup-index, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

# Welcome {-}

This book is the lecture-note companion for **MGMT-3105 -- Sustainability and
Business Practices**, a second-year course that introduces the philosophy of
*sustainability* and the discipline of *project management* to students in
electromechanical and mechanical engineering technology programs.

You are training for careers in the **automotive**, **food-processing**, and
**defense** industries. In every one of those sectors you will, sooner than you
expect, be handed a slice of a project: a line retrofit, a new product
introduction, a validation campaign, a plant move. This course is about the
structure that keeps such work on scope, on schedule, and on budget --- and
about doing it in a way that does not quietly borrow from the future to pay for
the present.

## Why a course that pairs sustainability with project management? {-}

At first glance they look like two subjects stapled together. They are not.
Every project consumes resources --- materials, energy, money, human effort ---
and every project leaves something behind: a product, a process, a waste stream,
a set of habits in the organization. **Project management is the set of tools
that decides how those resources are spent, and sustainability is the lens that
asks whether that spend was worth it beyond the next quarter.** A project
manager who can build a work breakdown structure but cannot reason about
lifecycle impact is only doing half the job.

```{r video-intro-pm, echo=FALSE, results='asis'}
embed_youtube("Jk-JwtScIlw", "What Is Project Management? (Association for Project Management)")
```

```{r video-intro-sustainability, echo=FALSE, results='asis'}
embed_youtube("7V8oFI4GYMY", "What Is Sustainable Development? (FN-sambandet)")
```

## Course map {-}

```{r fig-course-map, echo=FALSE, fig.cap="How the thirteen chapters of this book track the MGMT-3105 lecture sequence, grouped by project phase.", fig.align='center', fig.width=10, fig.height=6}
phases <- data.frame(
  chapter = 1:13,
  title = c(
    "Sustainability", "Intro & Communication", "Scope & Org. Structures",
    "Leadership & the PM", "Work Breakdown Structure", "Network Diagrams (AON)",
    "Risk Management", "Project Scheduling", "Cost Estimation & Budgeting",
    "Negotiation & Conflict", "Resource Management", "Project Control Systems",
    "Closeout & Termination"
  ),
  phase = c(
    "Foundations", "Foundations", "Initiation", "Initiation",
    "Planning", "Planning", "Planning", "Planning", "Planning",
    "Execution", "Execution", "Monitoring & Control", "Closing"
  )
)
phases$phase <- factor(phases$phase, levels = c(
  "Foundations", "Initiation", "Planning", "Execution",
  "Monitoring & Control", "Closing"
))
phases$row <- ave(phases$chapter, phases$phase, FUN = seq_along)

ggplot(phases, aes(x = phase, y = -row)) +
  geom_tile(aes(fill = phase), width = 0.9, height = 0.9,
            color = "white", linewidth = 1.2) +
  geom_text(aes(label = paste0(chapter, ". ", title)),
            color = "white", fontface = "bold", lineheight = 0.9,
            size = 2.9 * diagram_text_scale) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "MGMT-3105 at a glance", x = NULL, y = NULL) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 10 * diagram_text_scale, face = "bold",
                               margin = margin(t = 6))
  )
```

## Course learning outcomes {-}

On successful completion of MGMT-3105 you will be able to reliably:

1. Communicate the philosophy of project selection, including the Net Present Value method.
2. Communicate the theory of budgeting and scheduling.
3. Connect project-management concepts to real-world examples.
4. Identify the consequences to society of poor project planning.
5. Communicate the theory of progress management.
6. Discuss the problems of managing people and levelling allocated resources.
7. Communicate the organizational implications of closing a project.
8. Communicate negotiation theory.
9. Communicate effectively in a global environment.
10. Identify the use of sustainability in designs and in project management.

## The running project {-}

Throughout the book we return to one worked scenario, taken from the MGMT-3105
term project: your company designs and manages the launch of a **1/18-scale
radio-controlled vehicle replica** for the toy market, then wins a contract to
manufacture 100 000+ units for the following holiday season. You choose the
vehicle, the body and frame materials (steel, aluminium, or carbon fibre), and
one of two Canadian motor/control-system suppliers. Every chapter uses a piece
of that project --- its scope, its WBS, its schedule, its budget, its risks, its
closeout --- so the tools connect to a single concrete thing rather than a dozen
disconnected examples.

## How the book is organized {-}

Each chapter opens with measurable **learning objectives** and a real-world
**hook**, develops the theory with plain-language explanations and LaTeX
equations, and illustrates it with R-generated figures. Worked examples are
shown step by step; wherever an example is a self-contained calculation, it is
offered as a **live, editable R cell** you can run and modify in your browser.
Collapsible `Practice` and `Discussion` blocks let you test yourself before
revealing the answer, and each chapter closes with **curated YouTube videos**
for a second explanation in a different voice.

## Assessment (for the live course) {-}

| Component            | Timing   | Weight |
|----------------------|----------|-------:|
| Midterm test         | Week 7   |   25 % |
| Final test           | Week 14  |   25 % |
| Weekly quizzes       | Weekly   |   12 % |
| Weekly case studies  | Weekly   |   12 % |
| Assignment Phase A   | Week 7   |   10 % |
| Assignment Phase B   | Week 11  |   10 % |
| Assignment Phase C   | Week 14  |    6 % |

: MGMT-3105 grade breakdown {tbl-colwidths="[55,25,20]"}

Quizzes and case studies require attendance. Tests are cumulative, closed-book
with a handwritten aid sheet, and kit calculators are permitted. To pass, you
need an overall grade of 50 % **and** completion of the mandatory work.

## A note on academic integrity {-}

Write your reports and answers in your own words. Whenever you take an idea or a
passage from another author --- including a classmate, a previous year's work, or
an AI tool --- acknowledge it with quotation marks where appropriate and a proper
citation. If you are unsure how to credit shared work, ask **before** you
submit. By submitting, you put your name to the integrity of the work.

```{r write-pkg-bib, include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'dplyr', 'ggplot2', 'kableExtra'
), 'packages.bib')
```
